---
title: "Multinomial Logit"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
bibliography: references.bib
citations-hover: true
reference-location: document
warning: false
message: false
---

```{r}
#| include: false
library(tidyverse)
library(brms)  # attach here too: cached chunks below do not re-attach it
```

## The Inverse-Logit Function

Suppose $y_i \in \{0, 1\}$. Using logistic regression, we model $\Pr(y_i = 1)$ as a function of covariates, so that

$$
\Pr(Y_i = 1) = \frac{\exp(X_i \beta)}{1 + \exp(X_i \beta)}.
$$

The function $f(z) = \frac{e^z}{1 + e^z}$ is called the inverse-logit function and maps any real number $X_i \beta$ into a probability between 0 and 1.

Because the two probabilities must sum to one:

$$
\Pr(Y_i = 0) = 1 - \Pr(Y_i = 1)
             = \frac{1}{1 + \exp(X_i \beta)}.
$$

## The Softmax Function

The softmax generalizes the inverse-logit---it takes $J$ real inputs (i.e., from $\mathbb{R}^J$) and scales them to sum to one.

$$
\text{softmax}(z_j) = \frac{\exp(z_j)}{\sum_{k=1}^{J} \exp(z_k)}, \qquad j = 1,\ldots,J. 
$$
Because the $J$ outputs must sum to one, it is natural to interpret the outputs as *probabilities*.

---

The OJS widget below allows you to experiment with $J = 4$ inputs and see how the outputted probabilities change.

```{ojs}
//| echo: false

// Sliders (left column)
viewof z1 = Inputs.range([-5, 5], {step: 0.1, value: 1,   label: "z1"});
viewof z2 = Inputs.range([-5, 5], {step: 0.1, value: 0.5, label: "z2"});
viewof z3 = Inputs.range([-5, 5], {step: 0.1, value: -0.5,label: "z3"});
viewof z4 = Inputs.range([-5, 5], {step: 0.1, value: -1,  label: "z4"});

// Softmax pieces
labels = ["Pr(A)","Pr(B)","Pr(C)","Pr(D)"];
zs = [z1, z2, z3, z4];
exps = zs.map(Math.exp);
sumexp = exps.reduce((a, b) => a + b, 0);
ps = exps.map(e => e / sumexp);

// Data for stacked bar (one row, four segments)
data = ps.map((p, i) => ({ label: labels[i], value: p, row: "softmax" }));

// === Two-column layout: sliders (left) + live table (right) ===
{
const container = html`<div style="
  display:flex;
  gap:24px;
  align-items:flex-start;
  padding-top:8px;
"></div>`;

// Left column: sliders
const left = html`<div style="display:grid; gap:10px; min-width:220px;"></div>`;
left.append(viewof z1, viewof z2, viewof z3, viewof z4);

// Right column: table
const rows = labels.map((cls, i) => ({
Class: cls,
z: zs[i],
"exp(z)": exps[i],
Probability: ps[i]
}));

const right = html`<div style="
  min-width:240px;
  max-width:660px;
  padding:4px 0;
  font-size:2.00rem;
  line-height:1.1;
"></div>`;

right.append(
Inputs.table(rows, {
columns: ["Class", "z", "exp(z)", "Probability"],
format: {
z: d => d.toFixed(2),
"exp(z)": d => d.toFixed(3),
Probability: d => (d * 100).toFixed(1) + "%"
}
})
);

container.append(left, right);
return container;
}

// Stacked bar chart — bigger everything
Plot.plot({
width: 900,              // wider
height: 120,             // taller
marginLeft: 20,
marginRight: 20,
marginBottom: 50,
x: {
domain: [0, 1],
tickFormat: d => (d * 100).toFixed(0) + "%",
label: "Probability",
labelArrow: false,
labelAnchor: "center",
tickSize: 10,
tickPadding: 8
},
y: { axis: null },

// Bigger fonts everywhere
style: {
fontSize: "18px",        // main text size
fontFamily: "sans-serif"
},

color: {
legend: true,
label: "Class",
tickSize: 12
},

marks: [
Plot.barX(
data,
Plot.stackX({ x: "value", y: "row", fill: "label", sort: { fill: null } })
),
Plot.ruleX([0, 1])
]
})


```


## Adding covariates

Now suppose the outcome takes $J$ possible values:

$$
y_i \in \{1, 2, \ldots, J\}.
$$

**Example 1**: We might label *vote choice* in the US as (1) *abstain*, (2) *Republican*, (3) *Democrat*, or (4) *other*.

**Example 2**: We might label *coup attempts* in a given country-year as (0) *none attempted*, (1) *failed attempt*, or (2) *successful attempt*.

---

For each category $j$, define a linear predictor $\eta_{ij} = X_i \beta_j$. Here, each $\beta_j$ is a *vector* of coefficients.

We can use the *softmax* function to convert these linear predictors into probabilities that sum to one. 

$$
\Pr(Y_i = j) = \frac{\exp(\eta_{ij})}{\sum_{k=1}^{J} \exp(\eta_{ik})},
\qquad j = 1, \ldots, J.
$$

This is a generalization of the inverse-logit to $J$ categories. If $J = 2$, the softmax reduces to the inverse-logit.

---

However, this model is not identified because adding any constant to each of the $\eta_{ij}$ produces the same probabilities.

::: columns
::: column

**Example Inputs**

| Input $z_j$ | Softmax $p_j$ |
|------------:|--------------:|
| $-1$        | 0.0303 |
| $1$         | 0.2242 |
| $2$         | 0.6095 |
| $0.5$       | 0.1360 |

:::
::: column

**Adding a Constant ($+2$)**

| New Input $z_j + 2$        | Softmax $p_j$ |
|---------------------------:|--------------:|
| $-1 + 2 = 1$               | 0.0303 |
| $1 + 2 = 3$                | 0.2242 |
| $2 + 2 = 4$                | 0.6095 |
| $0.5 + 2 = 2.5$            | 0.1360 |

:::
:::

Adding a constant to all inputs leaves the softmax unchanged. This is why the multinomial logit model requires an identification constraint.

---

To identify the model, we set $\beta_J = 0$, so that $\eta_{iJ} = 0$. Then for $j = 1, \ldots, J-1$, we have

$$
\Pr(Y_i = j) =
\frac{\exp(X_i \beta_j)}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$

And for the "baseline" category $J$, we have

$$
\Pr(Y_i = J) =
\frac{1}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$

# Example: The 1992 presidential vote

## Data

- 1992 ANES [@anes1992]; same data file as the wk06 Bush approval example
- Outcome: `vote_1992`: Bush, Clinton, or Perot
- Predictors: `party_id` (−3 to 3), `economy` (1 = much better to 5 = much worse)
- Perot’s candidacy is what makes a nominal model necessary: Bush, Clinton, and Perot have no order.

```{r}
#| message: false
# load data; keep complete cases on the model's variables
vote <- read_csv("https://pos5747.github.io/data/bush-approval-1992.csv") |>
  select(vote_1992, party_id, economy) |>
  drop_na() |>
  mutate(vote_1992 = factor(vote_1992, levels = c("Bush", "Clinton", "Perot"))) |>
  glimpse()
```

## `multinom()` {.smaller}

```{r}
library(nnet)  # for multinom()

# fit multinomial logit; first level (Bush) is the baseline
fit_vote <- multinom(vote_1992 ~ party_id + economy, data = vote, trace = FALSE)
summary(fit_vote)
```

- One row of coefficients per non-baseline candidate each relative to Bush.

## Probabilities with {marginaleffects}

```{r}
library(marginaleffects)

# probability of each candidate as party ID varies
p <- predictions(fit_vote, newdata = datagrid(party_id = -3:3), type = "probs")
```

```{r}
#| echo: false
p |>
  as_tibble() |>
  select(group, party_id, estimate) |>
  pivot_wider(names_from = group, values_from = estimate) |>
  mutate(across(Bush:Perot, \(x) round(x, 2)))
```

## Probabilities with {marginaleffects}

```{r}
#| fig-asp: 0.45
ggplot(p, aes(x = party_id, y = estimate, ymin = conf.low, ymax = conf.high,
              color = group, fill = group)) +
  geom_ribbon(alpha = 0.2, color = NA) +
  geom_line() +
  labs(x = "Party ID", y = "Probability", color = "Candidate", fill = "Candidate")
```

- Pr(Perot) rises and then falls, even though both linear predictors are linear in party ID. The sign of a coefficient does not give the direction of that category’s probability.

## First differences

```{r}
# change in each candidate's probability as the economy goes from 2 to 4
comparisons(fit_vote,
            variables = list(economy = c(2, 4)),
            newdata = datagrid(grid_type = "mean_or_mode"),
            type = "probs")
```

- Perot’s economy coefficient is positive, but Clinton gains most of what Bush loses.

## `brm()` {.smaller}

```{r}
#| message: false
#| warning: false
#| cache: true
#| results: hide
library(brms)

# categorical() is the multinomial logit; first level (Bush) is the baseline
fit_vote_brm <- brm(vote_1992 ~ party_id + economy,
                    data = vote,
                    family = categorical(link = "logit"),
                    backend = "cmdstanr",
                    chains = 4,
                    cores = 4,
                    seed = 1234)
```

## `brm()` vs. `multinom()`

```{r}
# posterior means next to the ML estimates
fixef(fit_vote_brm)[, c("Estimate", "Q2.5", "Q97.5")]
coef(fit_vote)
```

## `brm()` + {marginaleffects}

```{r}
#| fig-asp: 0.4
# predictions() returns rowid but not the grid columns for this fit, so join them back
grid <- datagrid(model = fit_vote_brm, party_id = -3:3)
p_brm <- predictions(fit_vote_brm, newdata = grid) |>
  left_join(select(grid, rowid, party_id), by = "rowid")

ggplot(p_brm, aes(x = party_id, y = estimate, ymin = conf.low, ymax = conf.high,
                  color = group, fill = group)) +
  geom_ribbon(alpha = 0.2, color = NA) +
  geom_line() +
  labs(x = "Party ID", y = "Probability", color = "Candidate", fill = "Candidate")
```

# Callback: ordered logit with `brm()`

## Bush approval with `cumulative()` {.smaller}

- Last week’s `polr()` fit, refit with `brm(family = cumulative("logit"))`. Same parameterization: the cutpoints are `Intercept[j]`.

```{r}
#| message: false
#| warning: false
#| cache: true
#| results: hide
# the wk06 Bush approval data and model
approval <- read_csv("https://pos5747.github.io/data/bush-approval-1992.csv") |>
  mutate(bush_approval = factor(bush_approval, ordered = TRUE))

f_approval <- bush_approval ~ military_force + ideology_distance +
  economy + party_id + education

fit_approval <- MASS::polr(f_approval, data = approval, Hess = TRUE)
fit_approval_brm <- brm(f_approval,
                        data = approval,
                        family = cumulative(link = "logit"),
                        backend = "cmdstanr",
                        chains = 4,
                        cores = 4,
                        seed = 1234)
```

## Bush approval with `cumulative()`

```{r}
# posterior means next to the ML estimates (polr() calls the cutpoints zeta)
fixef(fit_approval_brm)[, c("Estimate", "Q2.5", "Q97.5")]
c(fit_approval$zeta, coef(fit_approval))
```

## Ordered vs. nominal

- Ordered (`cumulative()`) uses one set of coefficients; nominal (`categorical()`) uses J − 1 sets. Red State Part 3 in the exercises is a nominal case: did not vote, Harris, or Trump.

## References
