---
title: "Ideal Point Models"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
execute:
  echo: true
fig-align: center
fig-width: 8
fig-asp: 0.5
---


# Opening Comments

- Exam
- Workshop
- Paper

```{r echo=FALSE}
#| message: false
#| warning: false
options(width = 90)
options(digits = 3)

library(tidyverse)
theme_set(theme_minimal())
```

---

Here are the packages we're using:

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(marginaleffects)
library(modelsummary)
library(ggrepel)

# for fitting hierarchical models
library(lme4)
library(brms)  # see also rstanarm::stan_lmer()
library(rstan)

# for post-processing simulations
library(tidybayes)

# handle future conflicts
# - MASS v. dplyr for filter()
conflicted::conflict_prefer("filter", "dplyr")
# - MASS v. dplyr for select()
conflicted::conflict_prefer("select", "dplyr")

```

# Ideal Point Models

## `SupremeCourt` data

```{r}
# load MCMC package, which has SupremeCourt data
library(MCMCpack)

# load SupremeCourt data
# - for some reason, `sc <- MCMCpack::SupremeCourt` doesn't work 
data(SupremeCourt)

# quick look
glimpse(SupremeCourt)
```

## Tidy it up!

- Convert to long format
- Flip meaning of indicator

```{r}
sc <- SupremeCourt |>
  mutate(case_id = 1:n()) |>
  pivot_longer(Rehnquist:Breyer, 
               names_to = "justice", 
               values_to = "vote") |>
  mutate(appointed_by = case_when(justice %in% c("Breyer", "Ginsburg") ~ "Clinton",
                                  justice %in% c("Souter", "Thomas") ~ "Bush (41)",
                                  justice %in% c("Kennedy", "O'Connor", "Scalia") ~ "Reagan",
                                  justice == "Stevens" ~ "Ford",
                                  justice == "Rehnquist" ~ "Nixon",
                                  TRUE ~ NA_character_)) |>
  mutate(cons_vote = 1 - vote) |>
  mutate(case_id = paste0("Case ", case_id)) |>
  select(-vote) |>
  glimpse()
```

## Summarize

```{r}
sc |> 
  group_by(justice, appointed_by) |>
  summarize(prop_cons_votes = mean(cons_vote, na.rm = TRUE)) |>
  arrange(prop_cons_votes)
```

## Plot 1

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.5

sc_summary <- sc |> 
  group_by(justice, appointed_by) |>
  summarize(prop_cons_votes = mean(cons_vote, na.rm = TRUE)) |>
  ungroup() |>
  mutate(justice = reorder(justice, prop_cons_votes)) |>
  arrange(prop_cons_votes)

ggplot(sc_summary, aes(x = justice, y = prop_cons_votes)) + 
  geom_col(width = 0.5) + 
  geom_label(aes(label = scales::percent(prop_cons_votes, accuracy = 1)), hjust = 1.2) + 
  coord_flip()
```

## Plot 2

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4
library(ggrepel)

set.seed(1234)
ggplot(sc_summary, aes(x = prop_cons_votes, y = 0)) +
  geom_hline(yintercept = 0) +
  geom_point(size = 3) +
  geom_label_repel(
    aes(label = justice),
    nudge_y = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "y",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Proportion of Conserative Votes"  
    ) + 
  scale_y_continuous(limits = c(-.1, .1))
```

## Estimate "abilities" of justices and "difficulties" of cases

- ability is analagous to a justice's ideology.
- difficulty is analogous to a case's ideology.

```{r}
f <- cons_vote ~ (1 | justice) + (1 | case_id)
fit <- glmer(f, data = sc, family = binomial)
```

## Print the "abilities"

```{r}
ranef(fit)$justice |>
  as_tibble(rownames = "justice") |>
  rename(ideal_point = `(Intercept)`) |>
  arrange(ideal_point)
```

## Plot the abilities

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4

ip <- ranef(fit)$justice |>
  as_tibble(rownames = "justice") |>
  rename(ideal_point = `(Intercept)`) |>
  arrange(ideal_point)

set.seed(1234)
ggplot(ip, aes(x = ideal_point, y = 0)) +
  geom_hline(yintercept = 0) +
  geom_point(size = 3) +
  geom_label_repel(
    aes(label = justice),
    nudge_y = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "y",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Ideal Point"  
    ) + 
  scale_y_continuous(limits = c(-.1, .1))
```

## Add cases

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4

library(ggbeeswarm)

cases <- data.frame(case = as.numeric(ranef(fit)$case_id[, "(Intercept)"])) 

set.seed(1234)
ggplot(ip, aes(y = ideal_point, x = 0)) +
  geom_vline(xintercept = 0) +
  geom_beeswarm(data = cases, aes( y = case, x = 0), shape = 18, alpha = 0.3, color = "red") + 
  geom_beeswarm(size = 2, alpha = 0.5) +
  geom_label_repel(
    aes(label = justice),
    nudge_x = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "x",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Ideal Point"  
    ) + 
  scale_x_continuous(limits = c(-.1, .1)) + 
  coord_flip()
```

## Plot the logit curve

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4
mu <- as.numeric(fixef(fit))

beta <- ranef(fit)$case_id |>
  as_tibble(rownames = "case_id") |>
  rename(beta = `(Intercept)`)

cases_to_keep <- c("Case 20", "Case 5", "Case 2", "Case 42", "Case 22", "Case 33")

pred_lines <- crossing(justice_ideology = seq(-2, 2, length.out = 100), 
                      case_id = unique(sc$case_id)) |>
  left_join(beta) |>
  mutate(pr_cons_vote = plogis(mu + justice_ideology - beta)) |>
  mutate(case_id = reorder(case_id, pr_cons_vote)) |>
  filter(case_id %in% cases_to_keep)

pred_points <- sc |>
  left_join(beta) |> 
  left_join(ip) |>
  mutate(pr_cons_vote = plogis(mu + ideal_point - beta)) |>
  mutate(case_id = reorder(case_id, pr_cons_vote)) |>
    filter(case_id %in% cases_to_keep)

ggplot(pred_lines, aes(x = justice_ideology, y = pr_cons_vote)) + 
  geom_vline(xintercept = ip$ideal_point, color = "grey50", linewidth = 0.2) + 
  geom_line() + 
  geom_point(data = pred_points, aes(x = ideal_point, y = pr_cons_vote, shape = factor(cons_vote)), fill = "white") + 
  facet_wrap(vars(case_id)) + 
  scale_shape_manual(values = c(19, 21)) + 
  labs(subtitle = "Selected cases", 
       shape = "Conservative Vote", 
       y = "Probability of a Conservative Vote", 
       x = "Justice Ideal Point")
```



