---
title: "Week 3 Exercises"
distributed: 2026-09-08
show-hints: true
---

CTK things to add: references to notes as hints (i.e., links to specific sessions.), references to other questions if there's an important link. missing solution, check that notation in exercises matches notes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mean v. median {ref="mean-median"}

Suppose $y$ follows a normal model with mean $\mu$ and SD $\sigma$. Then $\operatorname{avg}(y)$ is the *best* unbiased estimator of $\mu$. But $\operatorname{median}(y)$ is also an unbiased estimator of $\mu$. Use a Monte Carlo simulation to show how $\operatorname{avg}(y)$ is a better estimator of $\mu$ than $\operatorname{median}(y)$.

::: hint

See [Section 4.2](https://pos5747.github.io/notes/wk03/01-sampling-distribution.html#example-normal-model) of the notes for the sampling distribution of $\operatorname{avg}(y)$ under a normal model, and [Section 4.1](https://pos5747.github.io/notes/wk03/01-sampling-distribution.html#example-the-toothpaste-cap-problem) for simulating a sampling distribution.

:::

::: sol

The average is better because the SE of the average estimator is smaller than the SE of the median estimator. 

```{r}
# parameters
mu <- 0
sigma <- 1
N <- 10 # sample size of y
n_mc_sims <- 10000   # monte carlo repetitions

# containers
mean_hat <- numeric(n_mc_sims)
median_hat <- numeric(n_mc_sims)

# simulate many data sets y; compute and store estimates for each
for (i in 1:n_mc_sims) {
  y <- rnorm(N, mean = mu, sd = sigma)
  mean_hat[i]   <- mean(y)
  median_hat[i] <- median(y)
}

sd(mean_hat)  # se of avg estimator
sd(median_hat)  # se of median estimator
```

:::


## Asymptotics as approximations {ref="asymptotics"}

For a large number of repetitions, do the following:

1. Simulate a data set $y$ with $N = 50$ observations from the exponential distribution with $\lambda = 2$.
1. Compute the point estimate $\hat{\lambda} = \frac{1}{\text{avg}(y)}$. Store this point estimate.

Compare the mean and SD of the simulated point estimates to the theoretical mean and SD of the *asymptotic* sampling distribution. We care, of course, about the *actual* sampling distribution, but we have to use an *asymptotic approximation* in practice. Discuss the differences and whether they seem important or unimportant.

Optional: Go beyond the mean and SD. Use a table of quantiles of the simulated point estimates and theoretical distribution and/or plots of the CDFs to compare the simulated (i.e., actual) and theoretical distributions.

::: hint

See [Section 6.2](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#theory) of the notes for the asymptotic normal approximation, including the callout there on treating asymptotic results as approximations. [Section 4.4.2](https://pos5747.github.io/notes/wk03/01-sampling-distribution.html#example-exponential-model) simulates the actual sampling distribution of $\hat{\lambda}$.

:::

::: hint

For the exponential model, the asymptotic sampling distribution of $\hat{\lambda}$ is $\hat{\lambda} \overset{a}{\sim} N\left(\lambda,\; \frac{\lambda^2}{N}\right)$. 

:::

::: sol

**Solution**

```{r}
# load packages
library(tinytable)
options(tinytable_tt_digits = 3)
options(tinytable_latex_placement = "H")

# set seed for reproducibility
set.seed(1234)

# parameters
lambda <- 2
N <- 50  # sample size
R <- 10000  # number of mc repetitions

# loop
lambda_hat <- numeric(R)  # container to store results
for (i in 1:R) {
 y <-  rexp(N, rate = lambda)  # simulate data set
 lambda_hat[i] <- 1/mean(y)  # compute and store estimate
}

# empirical summaries
emp_mean <- mean(lambda_hat)
emp_sd   <- sd(lambda_hat)

# asymptotic
asy_mean <- lambda
asy_sd   <- sqrt(lambda^2 / N)

# data frame of results
res <- data.frame(
  Statistic   = c("Mean", "SD"),
  Empirical   = c(emp_mean, emp_sd),
  Asymptotic  = c(asy_mean, asy_sd)
)

# print table
tt(
  res,
  digits  = 3,   
  align   = "lrr",
) 
```


```{r}
# quantile comparison table
probs <- c(0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99)
emp_q <- quantile(lambda_hat, probs = probs)
asy_q <- qnorm(probs, mean = asy_mean, sd = asy_sd)

# combine into data frame
quant <- data.frame(
  p = probs,
  empirical = emp_q,
  asymptotic = asy_q,
  difference = emp_q - asy_q
)

# make table
tt(
  quant,
  digits  = 3,   # 3 decimals for numbers
  align   = "lrr",
)
```

```{r}
#| message: false
#| warning: false

# load packages
library(tidyverse)

# plot ecdf and asymptotic distribution
gg_data <- data.frame(lambda_hat = lambda_hat)
ggplot(gg_data, aes(x = lambda_hat)) +
  stat_ecdf() +
  stat_function(fun = function(x) pnorm(x, 
                                        mean = asy_mean, 
                                        sd = asy_sd), 
                color = "red") +
  labs(
    title = "ECDF of Simulated ML Estimates vs Asymptotic Sampling Distribution",
    x = expression(hat(lambda)),
    y = "Cumulative Probability"
  )
```

*Discussion*

- **Center (bias).** The empirical mean of $\hat{\lambda}$ is about `r round(emp_mean, 3)`, compared to the true $\lambda=2.000$. This is a difference of roughly `r round(100*(emp_mean-asy_mean)/asy_mean, 1)`%. At $N=50$, this upward bias is present but quite small.
- **Spread.** The empirical SD is about `r round(emp_sd, 3)`, while the asymptotic SD is `r round(asy_sd, 3)`. This difference is `r round(100*abs(emp_sd-asy_sd)/asy_sd,1)`% of the asymptotic value, so the asymptotic variance formula provides an accurate description of variability.
- **Quantiles and tails.** For the median, the empirical value is `r round(emp_q["50%"],3)` compared to an asymptotic prediction of `r round(asy_q[5],3)`. At the 1% and 99% quantiles, the differences are about `r round(emp_q["1%"]-asy_q[1],3)` and `r round(emp_q["99%"]-asy_q[9],3)`, respectively, which correspond to differences of only a few percent relative to the true scale. The ECDF and normal CDF curves are nearly indistinguishable except in the tails.
- **Bottom line.** With $N=50$, the asymptotic normal distribution closely matches the finite-sample distribution of $\hat{\lambda}$. The bias is about `r round(100*(emp_mean-asy_mean)/asy_mean)`%, the SD differs by about `r round(100*abs(emp_sd-asy_sd)/asy_sd)`%, and quantile differences are small. The asymptotic normal approximation is accurate and reliable for practical use at this sample size.

:::

## German Tank Problem {ref="german-tank"}

Suppose a discrete uniform distribution from 0 to $K$. The pmf is $f(x; K) = \frac{1}{K + 1}, \quad x \in \{0, 1, \ldots, K\}$. The ML estimator is $\hat K = \max(y)$. The method of moments estimator is $\hat K = 2 \times \operatorname{avg}(y)$. Use a Monte Carlo simulation to evaluate bias and variance of each estimator for a small ($N = 3$), medium ($N = 25$), and large sample size ($N = 1,000$). Which estimator do you recommend?

*Note:* This model violates several of the regularity conditions, so the usual asymptotic guarantees do not hold.

::: hint

See [Section 4.3](https://pos5747.github.io/notes/wk03/01-sampling-distribution.html#bias) of the notes for bias and [Section 1.6](https://pos5747.github.io/notes/wk02/01-maximum-likelihood.html#principle-method-of-moments) for the method of moments.

:::

::: hint

[mean-median]{.ex-ref} similarly compares two estimators. However, the mean and median are both unbiased in that case, so only the variance matters.

:::

::: sol

**Solution**


```{r}
# load packages
library(tinytable)
library(dplyr)
options(tinytable_tt_digits = 3)
options(tinytable_latex_placement = "H")

# set seed for reproducibility
set.seed(1234)

# parameters
K  <- 100               # true K
Ns <- c(3, 25, 1000)    # sample sizes
R  <- 10000             # number of mc repetitions

# container: list of data frames
res_list <- list()

# loop over sample sizes
for (N in Ns) {
  K_hat_mle <- numeric(R)   # container for MLE
  K_hat_mom <- numeric(R)   # container for MOM
  
  for (i in 1:R) {
    y <- sample(0:K, N, replace = TRUE)  # simulate data set
    K_hat_mle[i] <- max(y)                # MLE
    K_hat_mom[i] <- 2 * mean(y)           # MOM
  }
  
  # empirical summaries
  emp_mean_mle <- mean(K_hat_mle)
  emp_sd_mle   <- sd(K_hat_mle)
  emp_mean_mom <- mean(K_hat_mom)
  emp_sd_mom   <- sd(K_hat_mom)
  
  # store results for this N
  res_list[[as.character(N)]] <- data.frame(
    N          = N,
    Estimator  = c("MLE", "MOM"),
    Mean       = c(emp_mean_mle, emp_mean_mom),
    SD         = c(emp_sd_mle,   emp_sd_mom),
    Bias       = c(emp_mean_mle - K, emp_mean_mom - K),
    MSE        = c(mean((K_hat_mle - K)^2), mean((K_hat_mom - K)^2))
  )
}

# combine all into one data frame
res <- bind_rows(res_list)

# print table
tt(
  res,
  digits = 5,
  align  = "lrrrrr",
)
```

Notice that the ML estimator is severely biased downward in small samples. But it also has a much smaller variance. It's not clear how to trade off these two. However, it's worth noting that the severely biased ML estimator has a *smaller* mean-squared error.

*Which estimator do you recommend?* There isn't an easy right answer here, just make an argument for one or the other from *principles* (e.g., bias, variance, MSE, or others).

:::

## Jensen's Inequality {ref="jensen"}

Suppose the usual exponential model. The ML estimate of the rate $\lambda$ is $\hat{\lambda} = \frac{1}{\text{avg}(y)}$. Using the invariance property, the ML estimate of the mean $\frac{1}{\lambda}$ is $\hat{\mu} = \frac{1}{\hat{\lambda}}$. Show analytically that the ML estimate of the mean is unbiased, while the ML estimate of the rate is biased. 

**Jensen's Inequality.**  If $g$ is a convex function and $X$ is a random variable with $\mathbb E[X]$ finite, then $\mathbb E\big[g(X)\big] \ge g \big(\mathbb E[X]\big)$,
with strict inequality unless $X$ is degenerate (i.e., constant almost surely).

::: hint

See [Section 4.3.2](https://pos5747.github.io/notes/wk03/01-sampling-distribution.html#example-exponential-distribution) of the notes.

:::

::: hint

First show that $\hat{\mu}$ is unbiased. Then use Jensen's inequality to show that $\hat{\lambda}$ is biased upward. You do not need to find the size of the bias, just use Jensen's inequality to show that the ML estimate of the rate is biased. To use Jensen's inequality, let $g(x)=1/x$, and notice that $g(x)$ is convex on $(0,\infty)$ (since $g''(x)=2/x^3>0$).

:::

::: sol

**Solution**

*1) Unbiasedness of the ML estimate of the mean*

$\hat{\mu} = \text{avg}(y) = \frac{1}{N} \sum_{i = 1}^N y_i$ is the sample mean of iid observations. $\mathbb E[y_i]=\frac{1}{\lambda}$. Then 

$$
\begin{aligned}
\mathbb E[\hat\mu] 
  &= \mathbb E \left[ \frac{1}{N} \sum_{i = 1}^N y_i \right] && \text{(definition of $\hat\mu$)}\\
  &= \frac{1}{N} \cdot \mathbb E \left[\sum_{i = 1}^N y_i \right] && \text{(linearity)}\\
  &= \frac{1}{N} \cdot \left[\sum_{i = 1}^N \mathbb E (y_i) \right] && \text{(sum of expectations)}\\
  &= \frac{1}{N} \cdot \left[\sum_{i = 1}^N \frac{1}{\lambda} \right] && \text{(each has mean $1/\lambda$)}\\
  &= \frac{1}{N} \cdot \left[N \cdot \frac{1}{\lambda} \right] && \text{(sum $N$ identical terms)}\\
  &= \frac{1}{\lambda} && \text{(simplify)}\\
  &= \mu && \text{(definition of $\mu$)}
\end{aligned}
$$

Because $\mathbb E(\hat\mu) = \mu$, $\hat{\mu}$ is an unbiased estimate of the mean.

*2) Bias of the ML estimate of the rate*

$\hat{\lambda} = \dfrac{1}{\text{avg}(y)} = \left(\dfrac{1}{N}\sum_{i=1}^N y_i\right)^{-1}$. Let $g(x)=1/x$, and notice that $g(x)$ is convex on $(0,\infty)$ (since $g''(x)=2/x^3>0$). Then

$$
\begin{aligned}
\mathbb E[\hat{\lambda}]
  &= \mathbb E\left[ \frac{1}{\text{avg}(y)} \right] && \text{(definition of $\hat{\lambda}$)}\\
  &= \mathbb E\left[g\big(\text{avg}(y)\big)\right] && \text{(definition of $g$ above)}\\
  &> g\big(\mathbb E[\text{avg}(y)]\big) && \text{(Jensen's inequality, $g$ convex)}\\
  &= \frac{1}{\mathbb E[\text{avg}(y)]} && \text{(apply $g$)}\\
  &= \frac{1}{\frac{1}{N}\sum_{i=1}^N \mathbb E(y_i)} && \text{(linearity of expectation)}\\
  &= \frac{1}{\frac{1}{N}\sum_{i=1}^N \frac{1}{\lambda}} && \text{(each $y_i$ has mean $1/\lambda$)}\\
  &= \frac{1}{\frac{1}{N}\cdot N \cdot \frac{1}{\lambda}} && \text{(sum of $N$ identical terms)}\\
  &= \lambda && \text{(simplify)}
\end{aligned}
$$

Since the inequality is strict for finite $N$ (the distribution of $\text{avg}(y)$ is non-degenerate), $\mathbb E[\hat{\lambda}] > \lambda$. Thus, the ML estimate of the **rate** is biased upward.

:::


## Poisson {ref="poisson"}

Suppose a Poisson model. The ML estimate of the mean parameter $\lambda$ is $\hat\lambda = \text{avg}(y)$. Find an estimate of the SE using the observed Fisher information.

::: hint

See [Section 6.2](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#theory) of the notes for the four steps that turn the log-likelihood into an SE estimate, and [Section 6.6](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#example-exponential-model) for those steps worked out for the exponential model.

:::

::: sol

**Solution**

The log-likelihood is $\ell(\lambda) = \sum_{i=1}^N \big[y_i \log \lambda - \lambda - \log(y_i!)\big]$.

First, find the second derivative with respect to $\lambda$.

$$
\begin{aligned}
\ell(\lambda)
&= \sum_{i=1}^N \big[y_i \log \lambda - \lambda - \log(y_i!)\big]
&& \text{Poisson log-likelihood} \\[6pt]
\frac{\partial \ell}{\partial \lambda}
&= \sum_{i=1}^N \left( y_i \cdot \frac{1}{\lambda} - 1 - 0 \right)
&& \text{use } \frac{d}{d\lambda}\log \lambda = \frac{1}{\lambda},\; \frac{d}{d\lambda}(-\lambda) = -1,\; \log(y_i!) \text{ is constant} \\[6pt]
&= \sum_{i=1}^N \left( \frac{y_i}{\lambda} - 1 \right)
= \frac{1}{\lambda} \sum_{i=1}^N y_i - N
&& \text{collect terms} \\[10pt]
\frac{\partial^2 \ell}{\partial \lambda^2}
&= \frac{\partial}{\partial \lambda} \left( \frac{1}{\lambda} \sum_{i=1}^N y_i - N \right)
&& \text{differentiate again} \\[6pt]
&= \sum_{i=1}^N \left( - \frac{y_i}{\lambda^2} \right) - 0
&& \text{use } \frac{d}{d\lambda}\left(\frac{1}{\lambda}\right) = -\frac{1}{\lambda^2} \text{, } N \text{ is constant} \\[6pt]
&= - \sum_{i=1}^N \frac{y_i}{\lambda^2}.
\end{aligned}
$$

The observed Fisher information is

$$
\mathcal I_{\text{obs}}(\lambda)
= -\frac{\partial^2 \ell}{\partial \lambda^2}
= \sum_{i=1}^N \frac{y_i}{\lambda^2}.
$$

Asymptotically, $\operatorname{Var}(\hat\lambda) \approx \mathcal I_{\text{obs}}(\lambda)^{-1}$.^[Rather than "approximately equals" ($\approx$), it's more technically correct to write "asymptotically equals." However, we treat "asymptotically equals" to mean "approximately" in practice. Given the practical orientation of the course, it seems better to write "approximately" here.]

To find the estimate of the SE, replace $\lambda$ with $\hat\lambda$:

$$
\widehat{\operatorname{SE}}(\hat\lambda)
= \sqrt{\mathcal I_{\text{obs}}(\hat\lambda)^{-1}}
= \sqrt{\frac{\hat\lambda^2}{\sum_{i=1}^N y_i}}
= \sqrt{\frac{\hat\lambda^2}{N \hat\lambda}}
= \sqrt{\frac{\hat\lambda}{N}}
= \sqrt{\frac{\operatorname{avg}(y)}{N}},
$$

since $\sum_{i=1}^N y_i = N \hat\lambda$.

:::

## `operations` {ref="operations"}

The code below loads the number of enforcement operations for each district in Bogota from Holland (2015) from the {crdata} package. See `?crdata::holland2015` for details. 

```{r}
# load holland2015 data set from crdata package
holland2015 <- crdata::holland2015 

# get operations in bogota only
ops <- holland2015$operations[holland2015$city == "bogota"]

# print
ops
```

**A Poisson model.**

**Task:** Using the Poisson estimators from [poisson]{.ex-ref}, find the ML estimate and SE estimate of $\lambda$ for a Poisson model. 

But before moving on, compare the SD of the predictive distribution (no need to simulate, it depends only and directly on $\lambda$) and the SD of the data. Why might this be a problem for the SE estimate?

::: hint

See [Section 3.1](https://pos5747.github.io/notes/wk02/03-predictive-distribution.html#example-poisson-distribution) of the notes for the predictive distribution of a Poisson model. [Section A.5](https://pos5747.github.io/notes/appendicies/999-distributions.html#poisson-distribution) gives its SD.

:::

::: sol

**Solution.**

First, fit the Poisson model and compute the ML estimate and SE of $\lambda$ without using `optim()`, following the observed-information derivation above. 

```{r}
#| echo: true
#| message: false
#| warning: false

# load packages
library(tidyverse)
library(tinytable)

# load holland2015 data set from crdata package
holland2015 <- crdata::holland2015 

# get operations in bogota only
ops <- holland2015$operations[holland2015$city == "bogota"]

# estimate lambda with ml
N <- length(ops)
lambda_hat_pois <- mean(ops); lambda_hat_pois
se_lambda_hat_pois  <- sqrt(lambda_hat_pois / N); se_lambda_hat_pois

# compare predictive SD to data SD
sd_pred_pois <- sqrt(lambda_hat_pois)   # predictive SD under Poisson
sd_data      <- sd(ops)                 # empirical SD

# summarize results in a small table
res_pois <- data.frame(
  Statistic = c("Predictive SD ", "Data SD"),
  Value     = c(sd_pred_pois, sd_data)
)

tt(res_pois, digits = 3, align = "lr")
```

This poses a big problem for our SE estimate. The SE estimate depends on the variance. If the variance is wildly underestimated, the the SE will be wildly underestimated as well.

:::

## Exponential {ref="exponential"}

Suppose an exponential model of $y$. We know that the ML estimate of the rate $\lambda$ is $\hat \lambda = \frac{1}{\operatorname{avg}(y)}$. By the invariance property, the ML estimate of the mean $\mu = \frac{1}{\lambda}$ is $\hat \mu = \frac{1}{\hat \lambda}$. 

1. Use the observed Fisher information to find a closed-form estimate of the SE of $\hat \lambda$. 
1. Then use the delta method to find a closed-form estimate of the SE of $\hat \mu$.

::: hint

See [Section 6.6](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#example-exponential-model) of the notes for the first part and [Section 7.2](https://pos5747.github.io/notes/wk03/04-delta-method.html#poisson-from-lambda-to-sd) for the delta method with one parameter.

:::

::: hint

For an exponential model with rate $\lambda$, the density is
$f(y; \lambda) = \lambda e^{-\lambda y}$, so
$\ell(\lambda) = \sum \big(\log \lambda - \lambda y_i\big)$.

:::

::: sol

**Solution**

The log-likelihood is
$$
\ell(\lambda) \;=\; \sum_{i=1}^N \big[\log \lambda \;-\; \lambda y_i\big].
$$

First, find the second derivative with respect to $\lambda$.

$$
\begin{aligned}
\ell(\lambda)
&= \sum_{i=1}^N \big[\log \lambda - \lambda y_i\big]
&& \text{Exponential log-likelihood} \\[6pt]
\frac{\partial \ell}{\partial \lambda}
&= \sum_{i=1}^N \left(\frac{1}{\lambda} - y_i\right)
&& \text{use } \frac{d}{d\lambda}\log \lambda = \frac{1}{\lambda},\; \frac{d}{d\lambda}(-\lambda y_i) = -y_i \\[6pt]
&= \frac{N}{\lambda} \;-\; \sum_{i=1}^N y_i
&& \text{collect terms} \\[10pt]
\frac{\partial^2 \ell}{\partial \lambda^2}
&= \frac{\partial}{\partial \lambda}\!\left(\frac{N}{\lambda} - \sum_{i=1}^N y_i\right)
&& \text{differentiate again} \\[6pt]
&= -\,\frac{N}{\lambda^2} \;-\; 0
&& \text{use } \frac{d}{d\lambda}\!\left(\frac{1}{\lambda}\right) = -\frac{1}{\lambda^2},\; \sum y_i \text{ is constant}.
\end{aligned}
$$

The observed Fisher information is
$$
\mathcal I_{\text{obs}}(\lambda)
= -\frac{\partial^2 \ell}{\partial \lambda^2}
= \frac{N}{\lambda^2}.
$$

To find the estimate of the SE, replace $\lambda$ with $\hat\lambda$:
$$
\widehat{\operatorname{SE}}(\hat\lambda)
= \sqrt{\mathcal I_{\text{obs}}(\hat\lambda)^{-1}}
= \sqrt{\frac{\hat\lambda^{2}}{N}}
= \frac{\hat\lambda}{\sqrt{N}}
= \frac{1}{\sqrt{N}\,\operatorname{avg}(y)}.
$$

Now suppose we use the invariance property to obtain $\hat\mu = \dfrac{1}{\hat\lambda}$. Then we can use the delta method to estimate the SE with $\tau(\lambda) = \dfrac{1}{\lambda}$.

The first derivative is $\tau'(\lambda) = -\frac{1}{\lambda^2}$.

Plugging in, we have
$$
\begin{aligned}
\widehat{\operatorname{Var}}(\hat\mu)
  &\approx \big[\tau'(\hat\lambda)\big]^2 \cdot \widehat{\operatorname{Var}}(\hat\lambda)
   \;=\; \left(\frac{1}{\hat\lambda^2}\right)^2 \cdot \frac{\hat\lambda^{2}}{N} \\[6pt]
  &= \frac{1}{N\,\hat\lambda^{2}}.
\end{aligned}
$$

And the standard error is
$$
\widehat{\mathrm{SE}}(\hat\mu)
\;\approx\; \sqrt{\frac{1}{N\,\hat\lambda^{2}}}
\;=\; \frac{1}{\sqrt{N}\,\hat\lambda}
\;=\; \frac{\operatorname{avg}(y)}{\sqrt{N}}.
$$

:::

## Delta method, by hand {ref="delta-by-hand"}

Suppose you obtain the ML estimate $\hat\theta=(\hat a,\hat b)=(2,\,1)$ with an estimated covariance matrix of
  $$
  \widehat{\mathrm{Var}}(\hat\theta)
  =
  \widehat{\mathrm{Var}}(\hat a,\hat b)
  =
  \begin{bmatrix}
  1 & \tfrac{1}{2}\\[2pt]
  \tfrac{1}{2} & 1
  \end{bmatrix}.
  $$ 
  
However, you want to estimate the quantity of interest $q = \tau(a,b)=\dfrac{a}{b}$. Find $\hat q$ using the invariance property and estimate the SE of $\hat q$ using the delta method. (Find the gradient analytically.) Confirm your work with R.

::: hint

See [Section 7.3.3](https://pos5747.github.io/notes/wk03/04-delta-method.html#the-matrix-algebra) of the notes.

:::

::: hint

[exponential]{.ex-ref} is a one-parameter version of this multi-parameter calculation.

:::

::: sol

**Solution**

First, find the gradient. For $\tau(a,b)=a/b$,

$$
\nabla\tau(a,b)=
\begin{bmatrix}
\frac{\partial}{\partial a}(a/b)\\[6pt]
\frac{\partial}{\partial b}(a/b)
\end{bmatrix}
=
\begin{bmatrix}
1/b\\[4pt]
-\,a/b^2
\end{bmatrix}.
$$
Plugging in $(\hat a,\hat b)=(2,1)$, we have $\nabla\tau(\hat a,\hat b)=\begin{bmatrix} 1\\[2pt] -2 \end{bmatrix}$.

Now do the required matrix multiplication.

$$
\begin{aligned}
\widehat{\mathrm{Var}}[\tau(\hat\theta)]
&\approx
\overbrace{\begin{bmatrix}1 & -2\end{bmatrix}}^{\nabla\tau(\hat\theta)^{\!\top}}
\Biggl(
\underbrace{\begin{bmatrix}
1 & \tfrac{1}{2}\\[2pt]
\tfrac{1}{2} & 1
\end{bmatrix}}_{\widehat{\mathrm{Var}}(\hat\theta)}
\underbrace{\begin{bmatrix}1\\ -2\end{bmatrix}}_{\nabla\tau(\hat\theta)}
\Biggr) \\[10pt]
&=
\begin{bmatrix}1 & -2\end{bmatrix}
\begin{bmatrix}
1\cdot 1 + \tfrac{1}{2}\cdot(-2)\\[4pt]
\tfrac{1}{2}\cdot 1 + 1\cdot(-2)
\end{bmatrix}
=
\begin{bmatrix}1 & -2\end{bmatrix}
\begin{bmatrix}
0\\[2pt]
-1.5
\end{bmatrix}
= 0 + 3 = 3.
\end{aligned}
$$
Thus $\widehat{\mathrm{Var}}[\widehat{\tau}]=3$ and $\widehat{\mathrm{SE}}(\widehat{\tau})=\sqrt{3}$.

We can confirm our work with R.

```{r}
#| echo: true
#| message: false
#| warning: false

# given values
a_hat <- 2
b_hat <- 1
V_hat <- matrix(c(1, 1/2,
                  1/2, 1), nrow = 2, byrow = TRUE)

# gradient at (a_hat, b_hat) for tau(a,b) = a/b
grad <- c(1 / b_hat, -a_hat / b_hat^2)  # c(1, -2)

# delta method
var_tau_hat <- grad %*% V_hat %*% grad
se_tau_hat  <- sqrt(var_tau_hat)

var_tau_hat  # 3
se_tau_hat   # sqrt(3)
```

:::

## Turnout rates, part 1 {ref="turnout1"}

Load a [tibble version of the 2024 turnout data](https://gist.github.com/carlislerainey/fda3b4eb7e95d50949a445d9ae7f02c8) from the [UF Election Lab](https://election.lab.ufl.edu/2024-general-election-turnout/) into R. 


Model the proportion of the voting-eligible population that voted using a beta distribution. Estimate the parameters of the beta distribution $\alpha$ and $\beta$ as well as the mean $\mu = \frac{\alpha}{\alpha + \beta}$ and the SD $\sigma = \sqrt{\frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}}$. Use a parametric bootstrap to estimate the SE of each.

::: hint

See [Section 5.2](https://pos5747.github.io/notes/wk03/02-parametric-bootstrap.html#example-beta-distribution) of the notes.

:::

::: hint

Fitting a beta model with `optim()` 1,000s of times might lead to a convergence error in a couple of fits. These might be important, but for the sake of *this exercise* we can ignore them.

:::

::: sol

**Solution**

```{r}
#| warning: false
#| message: false

# load packages
library(tidyverse)
library(tinytable)

# load data as tibble 
turnout_2024 <- tribble(
  ~state,             ~vep_turnout,
  "Alabama",          0.5893,
  "Alaska",           0.6378,
  "Arizona",          0.6360,
  "Arkansas",         0.5348,
  "California",       0.6206,
  "Colorado",         0.7314,
  "Connecticut",      0.6708,
  "Delaware",         0.6702,
  "District of Columbia", 0.6357,
  "Florida",          0.6671,
  "Georgia",          0.6826,
  "Hawaii",           0.5027,
  "Idaho",            0.6344,
  "Illinois",         0.6325,
  "Indiana",          0.5869,
  "Iowa",             0.7078,
  "Kansas",           0.6318,
  "Kentucky",         0.6219,
  "Louisiana",        0.6077,
  "Maine",            0.7424,
  "Maryland",         0.6930,
  "Massachusetts",    0.6803,
  "Michigan",         0.7464,
  "Minnesota",        0.7635,
  "Mississippi",      0.5745,
  "Missouri",         0.6434,
  "Montana",          0.6820,
  "Nebraska",         0.6796,
  "Nevada",           0.6580,
  "New Hampshire",    0.7405,
  "New Jersey",       0.6724,
  "New Mexico",       0.5957,
  "New York",         0.6044,
  "North Carolina",   0.7032,
  "North Dakota",     0.6310,
  "Ohio",             0.6539,
  "Oklahoma",         0.5328,
  "Oregon",           0.7194,
  "Pennsylvania",     0.7143,
  "Rhode Island",     0.6328,
  "South Carolina",   0.6214,
  "South Dakota",     0.6400,
  "Tennessee",        0.5761,
  "Texas",            0.5657,
  "Utah",             0.6415,
  "Vermont",          0.7089,
  "Virginia",         0.7119,
  "Washington",       0.7017,
  "West Virginia",    0.5546,
  "Wisconsin",        0.7664,
  "Wyoming",          0.6128
)

# beta log-likelihood
beta_ll_fn <- function(theta, y) { 
  alpha <- theta[1] 
  beta <- theta[2] 
  ll <- sum(dbeta(y, shape1 = alpha, shape2 = beta, log = TRUE))
  return(ll)
}

# create a function to fit the beta model
est_beta <- function(y) {
  est <- optim(par = c(3, 3), fn = beta_ll_fn, y = y,
               control = list(fnscale = -1),
               method = "BFGS") # for >1d problems
  if (est$convergence != 0) print("Model did not converge!")
  res <- list(est = est$par)
  return(res)
}

# ml estimates
ml_est <- est_beta(turnout_2024$vep_turnout)
a <- ml_est$est[1]; a  # alpha
b <- ml_est$est[2]; b  # beta
a / (a + b)  # mean; using a and b for readability
sqrt( (a * b) / ((a + b)^2 * (a + b + 1)) ) # SD

# parametric bootstrap
n_bs <- 100
bs_alpha_hat <- 
  bs_beta_hat <- 
  bs_mu_hat <-
  bs_sigma_hat <-
  numeric(n_bs)  # containers for the estimates

for (i in 1:n_bs) {
  # bootstrap sample
  bs_y <- rbeta(length(turnout_2024$vep_turnout), shape1 = ml_est$est[1], shape2 = ml_est$est[2])
  
  # fit model to bootstrapped sample
  bs_est <- est_beta(bs_y)
  
  # store ml estimates of alpha and beta for readability below
  bs_a <- bs_est$est[1]
  bs_b <- bs_est$est[2]

  # compute and store all qis
  bs_alpha_hat[i] <- bs_a
  bs_beta_hat[i]  <- bs_b
  bs_mu_hat[i]    <- bs_a / (bs_a + bs_b)
  bs_sigma_hat[i] <- sqrt( (bs_a * bs_b) / ((bs_a + bs_b)^2 * (bs_a + bs_b + 1)) )
}

# bootstrap SEs
bs_se_alpha_hat <- sd(bs_alpha_hat)
bs_se_beta_hat  <- sd(bs_beta_hat)
bs_se_mu_hat    <- sd(bs_mu_hat)
bs_se_sigma_hat <- sd(bs_sigma_hat)

# tinytable (avoid underscores in visible labels)
tab_df <- tibble::tibble(
  Parameter     = c("$\\alpha$", "$\\beta$", "$\\mu$", "$\\sigma$"),
  `ML estimate` = c(a_hat = a, 
                    b_hat = b, 
                    mu_hat = a / (a + b), 
                    sigma_hat = sqrt( (a * b) / ((a + b)^2 * (a + b + 1)) )),
  `Bootstrap SE`= c(bs_se_alpha_hat, bs_se_beta_hat, bs_se_mu_hat, bs_se_sigma_hat)
)
tt(
  tab_df,
  digits = 3,
  caption = "Beta model: ML estimates and parametric-bootstrap standard errors.",
)


```

:::

## Turnout rates, part 2 {ref="turnout2"}

Estimate the SEs of $\hat{\mu}$ numerically (i.e., using `optim()` with `hessian = TRUE` combined with the delta method with `numDeriv::grad()`). Compare to the results from the parametric bootstrap in part 1.

::: hint

See [Section 6.7](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#beta-model-and-optim) of the notes for $\widehat{\operatorname{Var}}(\hat\theta)$ from `optim()`, and [Section 7.3.4](https://pos5747.github.io/notes/wk03/04-delta-method.html#numerical-gradient) for the numerical gradient.

:::

::: hint

This is like [delta-by-hand]{.ex-ref}, except that $\widehat{\operatorname{Var}}(\hat\theta)$ is computed by `optim()` rather than given in the text of the question.. 

In [poisson]{.ex-ref}, you found the observed information for a single parameter. Here, the observed information is a $2 \times 2$ matrix.

:::

::: sol

**Solution**

First, find the variance matrix estimate with `optim()`.

```{r}
#| warning: false
#| message: false

library(numDeriv)  # for numerical gradients

# refit the model, this time keeping the hessian
# (turnout_2024 and beta_ll_fn come from part 1)
est <- optim(par = c(3, 3), fn = beta_ll_fn, y = turnout_2024$vep_turnout,
             control = list(fnscale = -1),
             method = "BFGS",
             hessian = TRUE)

theta_hat <- est$par        # c(alpha_hat, beta_hat)
info_obs  <- -est$hessian   # observed information (note the negative)
var_hat   <- solve(info_obs)  # estimated covariance matrix of theta_hat

theta_hat
var_hat
sqrt(diag(var_hat))  # ses of alpha_hat and beta_hat
```

Now use the delta method to convert the variance (matrix) estimates of $(\hat{alpha}, \hat{\beta})$ to an estimate of the variance of $\hat{\mu}$. 

- The quantity of interest is $\mu = \tau(\alpha, \beta) = \frac{\alpha}{\alpha + \beta}$. 
- The delta method is $\widehat{\operatorname{Var}}[\tau(\hat\theta)] \approx \nabla\tau(\hat\theta)^\top \widehat{\operatorname{Var}}(\hat\theta) \nabla\tau(\hat\theta)$. 

We could differentiate $\tau()$ by hand here, but `numDeriv::grad()` can do it numerically and that's what the question asks us to do!

```{r}
# the quantity of interest as a function of theta
mu_fn <- function(theta) {
  a <- theta[1]
  b <- theta[2]
  a / (a + b)
}

# numerical gradient of tau() evaluated at theta_hat
grad_mu <- grad(func = mu_fn, x = theta_hat)
grad_mu

# delta method
var_mu_hat <- t(grad_mu) %*% var_hat %*% grad_mu
se_mu_hat  <- sqrt(var_mu_hat)

se_mu_hat
```

Now compare the two of SEs for $\hat{\mu}$. (`bs_se_mu_hat` is the bootstrap SE  computed in part 1.)

```{r}
# tinytable (avoid underscores in visible labels)
compare_df <- tibble::tibble(
  Parameter         = c("$\\mu$"),
  `Bootstrap SE`    = c(bs_se_mu_hat),
  `Delta method SE` = c(se_mu_hat)
)
tt(
  compare_df,
  digits = 3,
  caption = "Beta model: parametric-bootstrap and delta-method SEs compared",
)
```

:::

## Two CIs {ref="two-cis"}

Consider a simple Bernoulli model for $y \in \{0,1\}$ with parameter $\pi \in [0,1]$. Create a small sample with a rare event: take $N = 20$ trials with exactly one success.

**Part 1.** *Wald confidence interval.*

Compute the ML estimate $\hat \pi = \operatorname{avg}(y)$, the Wald SE $\widehat{\mathrm{SE}}(\hat \pi) = \sqrt{\hat \pi(1-\hat \pi)/N}$, and the usual $95\%$ Wald confidence interval $\hat \pi \pm 1.96 \cdot\widehat{\mathrm{SE}}(\hat \pi)$. What do you notice about the interval relative to the parameter space $[0,1]$?

**Part 2.** *Parametric bootstrap confidence interval.*

Use a *parametric bootstrap* to obtain a $95\%$ percentile confidence interval for $\pi$ based on the fitted model $\text{Bernoulli}(\hat \pi)$. Compare the bootstrap interval with the Wald interval. What stands out? Briefly explain.

::: hint

See [Section 6.5](https://pos5747.github.io/notes/wk03/03-fisher-information-matrix.html#from-curvature-to-wald-confidence-intervals) of the notes for the Wald interval and [Section 5.1](https://pos5747.github.io/notes/wk03/02-parametric-bootstrap.html#example-toothpaste-cap-problem) for a bootstrap percentile interval on a Bernoulli model.

:::

::: hint

This is like the parametric bootstrap from [turnout1]{.ex-ref}. Repeatedly simulate $y^{*(b)}_1,\dots,y^{*(b)}_N \stackrel{iid}{\sim} \text{Bernoulli}(\hat \pi)$, recompute $\hat \pi^{*(b)} = \operatorname{avg}(y^{*(b)})$, and then take the empirical $2.5^\text{th}$ and $97.5^\text{th}$ percentiles of $\{\hat \pi^{*(b)}\}$.

:::

::: sol

**Solution.**

```{r}
#| echo: true
#| message: false
#| warning: false

set.seed(1)

# data: N = 20 with exactly one success
N <- 20
y <- c(1, rep(0, N - 1))  # rare event sample
y <- sample(y)            # shuffle (irrelevant, but tidy)

# ml estimate and wald 95%ci
pi_hat   <- mean(y)
se <- sqrt(pi_hat * (1 - pi_hat) / N)
wald_ci <- pi_hat + c(-1, 1) * 1.96 * se

pi_hat
se
wald_ci
```

The Wald interval extends below 0, which is outside the parameter space $[0,1]$.

```{r}
#| echo: true
#| message: false
#| warning: false

# parametric bootstrap w/ percentile 95% ci
n_bs <- 2000
bs_est <- numeric(n_bs)  # a container for the estimates
for (i in 1:n_bs) {
  bs_y <- rbinom(20, size = 1, prob = pi_hat)
  bs_est[i] <- mean(bs_y)
}

print(quantile(bs_est, probs = c(0.025, 0.975)), digits = 2)  # 95% ci
```

The Wald interval uses a normal approximation centered at $\hat \pi$ with variance estimated by Fisher information. With extreme $\hat \pi$ near zero and small $N$, this approximation is poor and the lower bound of the CI can dip below 0. In contrast, the parametric bootstrap respects the model's support by construction (every $\hat \pi^{*(b)} \in [0,1]$), so the percentile interval stays within $[0,1]$ and typically provides more reasonable coverage in this setting.

:::
