---
title: "Randomized Experiments"
subtitle: "An Adventure in Nonparametric Inference"
#outertheme: "infolines"
#innertheme: "circles"
colortheme: "dove"
theme: "Goettingen"
#innertheme: metropolis
format: 
  beamer: 
    fontsize: "8pt"
    slide-level: 3
editor: source
header-includes:
  - "\\usepackage{booktabs}"
  - "\\usepackage{amssymb}"
  - "\\setbeamertemplate{itemize items}{\\ensuremath{\\circ}}"
  - "\\usepackage{hyperref}"
  - "\\hypersetup{colorlinks=true, linkcolor=red, urlcolor=red, citecolor=red, anchorcolor=red}"
echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
options(width = 60)
options(digits = 3)
```

Here's the packages we're using.

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(modelsummary)
```

# Opening Comments

This week, we continue applied modeling!

1. **engines**: design-based inference!
1. **distribution**: none!
1. **quantities of interest**: the ATE!
1. **evaluating models** frequentist properties

---

The notation here follows Imbens and Rubin (2015). An exception is that I use $\text{ATE}$ to represent their $\tau_{\text{fs}}$ and $\widehat{\text{ATE}}$ to represent their $\hat{\tau}_{\text{dif}}$.

# Review

---

- For discrete random variable $X$, $E(X) = \sum_{\text{all }x}x \cdot \Pr(X = x)$
- $\text{Var}(X) = E\left[ (X - E(X))^2\right] = E\left[ X^2\right] - E(X)^2$. 

---

Our goal is to obtain point estimates, variance estimates, and confidence intervals from randomized experiments **using only the assumptions of the design**.

# $\text{ATE}$

We have $N$ units in our experiment and we assign $N_t$ units to treatment and $N_c = N - N_t$ units to control. We require that $1 < N_t < N$, so that at least one unit is assigned to treatment and control.

We use the random variable $W_i$ to assign unit $i$ to treatment ($W_i = 1$) or control ($W_i = 0$).

For each unit $i$, we have two potential outcomes: $Y_i(1)$ if assigned to treatment and $Y_i(0)$ if assigned to control.

---

For unit $i \in \{1,...,N\}$, we have one observed potential outcome, denoted by $Y_i^{\text{obs}}$.

$$
Y_i^{\text{obs}} = Y_i(W_i) = 
\begin{cases} 
Y_i(0) & \text{if } W_i = 0, \\
Y_i(1) & \text{if } W_i = 1.
\end{cases}
$$

For each unit, we also have one missing potential outcome, denoted by $Y_i^{\text{mis}}$:

$$
Y_i^{\text{mis}} = Y_i(1 - W_i) = 
\begin{cases} 
Y_i(1) & \text{if } W_i = 0, \\
Y_i(0) & \text{if } W_i = 1.
\end{cases}
$$

---

So what is this $Y_i(\cdot)$ thing?

### A Data Set of Fully Observed Potential Outcomes

\begin{table}[ht]
\centering
\begin{tabular}{cccc}
\hline
Unit & \(Y_i(0)\) & \(Y_i(1)\) & \(Y_i(1) - Y_i(0)\) \\ \hline
1 & 0 & 1 & 1 \\
2 & 1 & 2 & 1 \\
3 & 0 & 0 & 0 \\
4 & 1 & 3 & 2 \\
5 & 0 & 1 & 1 \\
6 & 1 & 2 & 1 \\
7 & 0 & 0 & 0 \\
8 & 1 & 3 & 2 \\
9 & 0 & 1 & 1 \\
10 & 1 & 2 & 1 \\ \hline
\end{tabular}
\caption{Mock Dataset with Potential Outcomes and Their Differences}
\end{table}

---


**Average Treatment Effect**: Lets use these potential outcomes to define the average treatment effect (ATE) as $\text{ATE} = \frac{1}{N}\sum_{i = 1}^N \left[ Y_i(1) - Y_i(0)\right] = \overline{Y}(1) - \overline{Y}(0)$, where $\overline{Y}(0)$ and $\overline{Y}(1)$ are the averages of the potential control and treated outcomes respectively, so that $\overline{Y}(0) = \frac{1}{N} \sum_{i=1}^{N} Y_i(0)$ and $\overline{Y}(1) = \frac{1}{N} \sum_{i=1}^{N} Y_i(1)$.


In our case, we are explicitly interested in what we call today the **sample average treatment effect** (SATE). That is, were are interested in the ATE among the units *in our sample* not an existing or imagined larger population.

### Two Potential Realization for the Experiment

\begin{table}
    \centering
    \begin{tabular}[t]{cccc}
    \hline
    Unit & \(Y_i(0)\) & \(Y_i(1)\) & \(W_i\) \\ \hline
    1 & ? & 1 & 1 \\
    2 & 1 & ? & 0 \\
    3 & ? & 0 & 1 \\
    4 & 1 & ? & 0 \\
    5 & 0 & ? & 0 \\
    6 & ? & 2 & 1 \\
    7 & 0 & ? & 0 \\
    8 & ? & 3 & 1 \\
    9 & 0 & ? & 0 \\
    10 & ? & 2 & 1 \\ \hline
    \end{tabular}
    \quad % Spacing between tables
    \begin{tabular}[t]{cccc}
    \hline
    Unit & \(Y_i(0)\) & \(Y_i(1)\) & \(W_i\) \\ \hline
    1 & 0 & ? & 0 \\
    2 & ? & 2 & 1 \\
    3 & 0 & ? & 0 \\
    4 & ? & 3 & 1 \\
    5 & ? & 1 & 1 \\
    6 & 1 & ? & 0 \\
    7 & ? & 0 & 1 \\
    8 & 1 & ? & 0 \\
    9 & ? & 1 & 1 \\
    10 & 1 & ? & 0 \\ \hline
    \end{tabular}
\end{table}

Notice that we cannot compute the ATE from these data because we cannot never obtain $Y_i(1) - Y_i(0)$. (This is the fundamental problem of causal inference).

# $\widehat{\text{ATE}}$

### An Estimator for the Average Treatment Effect

Suppose that we observe data from a completely randomized experiment in which $N_t = \sum_{i=1}^{N} W_i$ units are randomly selected to be assigned to treatment and the remaining $N_c = \sum_{i=1}^{N} (1 - W_i)$ are assigned to control. Because of the randomization, a natural^[See next slide] estimator for the average treatment effect is the difference in the average outcomes between those assigned to treatment and those assigned to control, so that

$$
\widehat{\text{ATE}} = \overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c},
$$

where


$$
\overline{Y}^{\text{obs}}_{c} = \frac{1}{N_c} \sum_{i:W_i=0} Y^{\text{obs}}_{i} \quad \text{and} \quad\overline{Y}^{\text{obs}}_{t} =\frac{1}{N_t} \sum_{i:W_i=1} Y^{\text{obs}}_{i}.
$$

Recall that $Y_i^{\text{obs}} = Y_i(1)$ if $W_i = 1$ and $Y_i^{\text{obs}} = Y_i(0)$ if $W_i = 0$. Thus we can write $\widehat{\text{ATE}}$ as

$$
\widehat{\text{ATE}} = \frac{1}{N} \sum_{i=1}^{N} \left( \frac{W_i \cdot Y_i(1)}{N_t/N} - \frac{(1 - W_i) \cdot Y_i(0)}{N_c/N} \right).
$$

### Why $\widehat{\text{ATE}}$?

Why is $\widehat{\text{ATE}} = \overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}$ a ``natural'' estimator of the ATE? 

For both intuitive and rigorous reasons, simply replacing population moments (e.g., the population mean) with sample moments (e.g., the sample mean) produces a good estimator. 

Here's a sketch of the theory. For distributions $F$ that are ``not too weird'' (i.e., meet fairly weak regularity conditions), then the empirical cdf $\hat{F}$ is a "good substitute" for $F$. It's a good substitute in the sense that $\hat{F}_N \rightarrow F$ as $N \rightarrow \infty$. This is an important "brute force" approach to nonparametric estimation.

In our case, we don't use it to defend the estimator, but to obtain a starting point.

### What makes $\widehat{\text{ATE}}$ uncertain?

From our usual parametric modeling perspective, we *imagine* that the data are draw from a distribution $y_i \sim f(\theta_i)$. 

But this approach to randomized experiments is not like that; we're not imagining the $Y_i(1)$ and $Y_i(0)$ are random. Instead, they are fixed values. 

The only thing that's random in the equation below is $W_i$, which is the assignment to treatment and control. Because we designed that mechanism, we know its distribution. Thus we have *design-based inference* rather than model-based inference.

$$
\widehat{\text{ATE}} = \frac{1}{N} \sum_{i=1}^{N} \left( \frac{W_i \cdot Y_i(1)}{N_t/N} - \frac{(1 - W_i) \cdot Y_i(0)}{N_c/N} \right).
$$

### Is $\widehat{\text{ATE}}$ a good estimator?



**Theorem**: $\widehat{\text{ATE}} = \overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}$ is an unbiased estimator of $\text{ATE}$.

*Proof.*

We're trying to show that $E_W \left[ \widehat{\text{ATE}} \right] = \text{ATE}$. *Walk through $(2) \rightarrow (3)$ together.*

\small

\begin{align}
E_W \left[ \widehat{\text{ATE}} \right] &= E_W \left[ \overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c} \right] \\
&= E_W \left[ \frac{1}{N} \sum_{i=1}^{N} \left( \frac{W_i \cdot Y_i(1)}{N_t/N} - \frac{(1 - W_i) \cdot Y_i(0)}{N_c/N} \right) \right] \\
&= \frac{1}{N} \sum_{i=1}^{N} \left( E_W[W_i] \cdot Y_i(1) \frac{1}{N_t/N} - E_W[1 - W_i] \cdot Y_i(0) \frac{1}{N_c/N} \right) \\
&= \frac{1}{N} \sum_{i=1}^{N} (Y_i(1) - Y_i(0)) \\
&= \text{ATE}
\end{align}

\normalsize

Done! $\widehat{\text{ATE}}$ is an unbiased estimator of $\text{ATE}$ under the assumptions of the design. No parametric model of the data needed, just a (known) model of the the design.

# $\text{Var} \left( \widehat{\text{ATE}} \right)$

### Some Initial Results

To calculate the variance of $\widehat{\text{ATE}} = \overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}$, we need $E_W[W_i^2]$ and $\text{Var}_W(W_i)$.

We have

$$
E_W[W_i^2] = E_W[W_i] = \frac{N_t}{N} \quad \text{and} \quad \text{Var}_W(W_i) = \frac{N_t}{N} \left( 1 - \frac{N_t}{N} \right).
$$

*Proof.* Homework Exercise.

---

We also need $E_W[W_i \cdot W_{i'}]$.

$$
E_W[W_i \cdot W_{i'}] = P_{W}(W_i = 1) \cdot P_{W}(W_{i'} = 1|W_i = 1) = \frac{N_t}{N} \cdot \frac{N_t - 1}{N - 1}, \quad \text{for } i \neq j,
$$

---

**Theorem** The variance of $\hat{\text{ATE}}$ is

$$
\text{Var}_W (\overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}) = \frac{S_c^2}{N_c} + \frac{S_t^2}{N_t} - \frac{S_{tc}^2}{N},
$$

where $S_c^2$ and $S_t^2$ are the variances of $Y_i(0)$ and $Y_i(1)$:

$$
S_c^2 = \frac{1}{N - 1} \sum_{i=1}^{N} (Y_i(0) - \overline{Y}(0))^2, \quad \text{and} \quad S_t^2 = \frac{1}{N - 1} \sum_{i=1}^{N} (Y_i(1) - \overline{Y}(1))^2,
$$

and $S_{tc}^2$ is the variance of the unit-level treatment effects:

$$
S_{tc}^2 = \frac{1}{N - 1} \sum_{i=1}^{N} \left(Y_i(1) - Y_i(0) - (\overline{Y}(1) - \overline{Y}(0))\right)^2.
$$
*Proof.* Long and tedious! So here we go...

---


Remember,

$$
\widehat{\text{ATE}} = \frac{1}{N_t} \sum_{i=1}^{N} W_i \cdot Y^{\text{obs}}_i - \frac{1}{N_c} \sum_{i=1}^{N} (1 - W_i) \cdot Y^{\text{obs}}_i.
$$
Re-arranging, we have

$$
= \frac{1}{N} \sum_{i=1}^{N} \left( \frac{N}{N_t} \cdot W_i \cdot Y_i(1) - \frac{N}{N_c} \cdot (1 - W_i) \cdot Y_i(0) \right).
$$

It happens to be helpful to replace $W_i$ with $D_i$.

$$
D_i = W_i - \frac{N_t}{N} = \begin{cases}
\frac{N_c}{N} & \text{if } W_i = 1, \\
-\frac{N_t}{N} & \text{if } W_i = 0.
\end{cases}
$$

Facts:

- $E(D_i) = 0$.
- $\text{Var}_W(D_i) = E[D_i^2] = \frac{N_tN_c}{N^2}$. 

*Proof.* Homework exercise.

---

Another fact: For $i \neq j$ the distribution of $D_i \cdot D_j$ is

$$
P_{W}(D_i \cdot D_j = d) = \begin{cases}
\frac{N_t \cdot (N_t - 1)}{N \cdot (N - 1)} & \text{if } d = \frac{N_c^2}{N^2}, \\
\frac{2 \cdot N_t \cdot N_c}{N \cdot (N - 1)} & \text{if } d = -\frac{N_tN_c}{N^2}, \\
\frac{N_c \cdot (N_c - 1)}{N \cdot (N - 1)} & \text{if } d = \frac{N_t^2}{N^2}, \\
0 & \text{otherwise}.
\end{cases}
$$

Thus,

$$
E_W[D_i \cdot D_j] = \begin{cases}
\frac{N_c \cdot N_t}{N^2} & \text{if } i = j, \\
-\frac{N_t \cdot N_c}{N^2 \cdot (N - 1)} & \text{if } i \neq j.
\end{cases}
$$

*Proof.* Homework exercise.

---

Written using $D_i$ rather than $W_i$, the estimate is:

$$
\widehat{\text{ATE}} = \frac{1}{N} \sum_{i=1}^{N} \left( \frac{N}{N_t} \cdot D_i + \frac{N_t}{N} \right) \cdot Y_i(1) - \frac{N}{N_c} \left( \frac{N_c}{N} - D_i \right) \cdot Y_i(0).
$$
Re-arranging, this becomes

$$
= \frac{1}{N} \sum_{i=1}^{N} (Y_i(1) - Y_i(0)) + \frac{1}{N} \sum_{i=1}^{N} D_i \cdot \left( \frac{N}{N_t} \cdot Y_i(1) + \frac{N}{N_c} \cdot Y_i(0) \right).
$$

Exercise: Using the equation above, show that $\widehat{\text{ATE}}$ is an unbiased estimator of $\text{ATE}$.

---

$$
\widehat{\text{ATE}} = \overbrace{\frac{1}{N} \sum_{i=1}^{N} (Y_i(1) - Y_i(0))}^{\text{fixed}} + \underbrace{\frac{1}{N} \sum_{i=1}^{N} D_i \cdot  \overbrace{ \left( \frac{N}{N_t} \cdot Y_i(1) + \frac{N}{N_c} \cdot Y_i(0) \right)}^{Y_i^+}}_{\text{random variable}}
$$

Let $Y_i^+ = (N/N_t)Y_i(1) + (N/N_c)Y_i(0)$. Then

$$
\text{Var}_W (\widehat{\text{ATE}}) = \text{Var}_W \left( \frac{1}{N} \sum_{i=1}^{N} D_i \cdot Y_i^+ \right) = \frac{1}{N^2} E_W \left[ \left( \sum_{i=1}^{N} D_i \cdot Y_i^+ \right)^2 \right].
$$

Expanding, we get:

$$
\text{Var}_W (\widehat{\text{ATE}}) = E_W \left[ \frac{1}{N^2} \sum_{i=1}^{N} \sum_{j=1}^{N} D_iD_jY_i^+Y_j^+ \right]
$$
$$
= \frac{1}{N^2} \sum_{i=1}^{N} (Y_i^+)^2 E_W [D_i^2] + \frac{1}{N^2} \sum_{i=1}^{N} \sum_{j \neq i} E_W [D_i \cdot D_j] \cdot Y_i^+ \cdot Y_j^+
$$
$$
= \frac{N_c \cdot N_t}{N^4} \sum_{i=1}^{N} (Y_i^+)^2 - \frac{N_c \cdot N_t}{N^4 \cdot (N - 1)} \sum_{i=1}^{N} \sum_{j \neq i} Y_i^+ \cdot Y_j^+
$$

$$
= \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} (Y_i^+)^2 - \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \sum_{j \neq i} Y_i^+ \cdot Y_j^+
$$

$$
= \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \left( Y_i^+ - \overline{Y}^+ \right)^2
$$
$$
= \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \left( \frac{N}{N_t} \cdot Y_i(1) + \frac{N}{N_c} \cdot Y_i(0) - \left( \frac{N}{N_t} \cdot \overline{Y}(1) + \frac{N}{N_c} \cdot \overline{Y}(0) \right) \right)^2
$$

---

\begin{align*}
&= \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \left( \frac{N}{N_t} \cdot Y_i(1) - \frac{N}{N_t} \cdot \overline{Y}(1) \right)^2 \\
&\quad + \frac{N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \left( \frac{N}{N_c} \cdot Y_i(0) - \frac{N}{N_c} \cdot \overline{Y}(0) \right)^2 \\
&\quad + \frac{2 \cdot N_c \cdot N_t}{N^3 \cdot (N - 1)} \sum_{i=1}^{N} \left( \frac{N}{N_t} \cdot Y_i(1) - \frac{N}{N_t} \cdot \overline{Y}(1) \right) \cdot \left( \frac{N}{N_c} \cdot Y_i(0) - \frac{N}{N_c} \cdot \overline{Y}(0) \right) \\
&= \frac{N_c}{N \cdot N_t \cdot (N - 1)} \sum_{i=1}^{N} (Y_i(1) - \overline{Y}(1))^2 + \frac{N_t}{N \cdot N_c \cdot (N - 1)} \sum_{i=1}^{N} (Y_i(0) - \overline{Y}(0))^2 \\
&\quad + \frac{2}{N \cdot (N - 1)} \sum_{i=1}^{N} (Y_i(1) - \overline{Y}(1)) \cdot (Y_i(0) - \overline{Y}(0)). 
\end{align*}

---

Recall the definition of $S_{tc}^2$, which implies that
\begin{align*}
S_{tc}^2 &= \frac{1}{N - 1} \sum_{i=1}^{N} \left(Y_i(1) - \overline{Y}(1) - (Y_i(0) - \overline{Y}(0))\right)^2 \\
&= \frac{1}{N - 1} \sum_{i=1}^{N} \left(Y_i(1) - \overline{Y}(1)\right)^2 + \frac{1}{N - 1} \sum_{i=1}^{N} \left(Y_i(0) - \overline{Y}(0)\right)^2 \\
&\quad - \frac{2}{N - 1} \sum_{i=1}^{N} \left(Y_i(1) - \overline{Y}(1)\right) \cdot \left(Y_i(0) - \overline{Y}(0)\right) \\
&= S_t^2 + S_c^2 - \frac{2}{N - 1} \sum_{i=1}^{N} \left(Y_i(1) - \overline{Y}(1)\right) \cdot \left(Y_i(0) - \overline{Y}(0)\right).
\end{align*}

Hence, the expression in (Result 1) is equal to
\begin{align*}
\text{Var}_W (\overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}) &= \frac{N_c}{N \cdot N_t} \cdot S_t^2 + \frac{N_t}{N \cdot N_c} \cdot S_c^2 \\
&\qquad + \frac{1}{N} \left( S_t^2 + S_c^2 - S_{tc}^2 \right) \\
&= \frac{S_t^2}{N_t} + \frac{S_c^2}{N_c} - \frac{S_{tc}^2}{N}.
\end{align*}

# $\widehat{\text{Var}} \left( \widehat{\text{ATE}} \right)$

For a given set of potential outcomes, we now know the variance. But to make inferences, we need to be able to *estimate* this variance using the observed data $Y^\text{obs}_i$. 

An unbiased estimator for $S_c^2$ is

$$
s_c^2 = \frac{1}{N_c - 1} \sum_{i:W_i=0} \left( Y_i(0) - \overline{Y}^{\text{obs}}_c \right)^2 = \frac{1}{N_c - 1} \sum_{i:W_i=0} \left( Y^{\text{obs}}_i - \overline{Y}^{\text{obs}}_c \right)^2.
$$

Similarly, an unbiased estimator for $S_c^2$ is

$$
s_t^2 = \frac{1}{N_t - 1} \sum_{i:W_i=1} \left( Y_i(1) - \overline{Y}^{\text{obs}}_t \right)^2 = \frac{1}{N_t - 1} \sum_{i:W_i=1} \left( Y^{\text{obs}}_i - \overline{Y}^{\text{obs}}_t \right)^2.
$$

---

The third term, $S_{tc}^2$, is the tricky one. This is population variance of the *unobservable* unit-level treatment effects---we cannot estimate this quantity.

We can only make assumptions about this quantity.

**Assumption 1 (Sharp Null)**:  If the treatment effect $Y_i(1) - Y_i(0)$ equals zero, then an unbiased estimator of the variance of $\widehat{ATE}$ is

$$
\widehat{\text{Var}}(\widehat{ATE}) = \frac{s_c^2}{N_c} + \frac{s_t^2}{N_t}. \tag{Neyman Variance Estimator}
$$
**Assumption 2 (Constant Effects)**:  If the treatment effect $Y_i(1) - Y_i(0)$ is constant, then an unbiased estimator of the variance of $\widehat{ATE}$ is also Neyman's variance estimator.

---

**Without an Assumption**:  For any $Y_i(1)$ and $Y_i(0)$, then expectation of Neyman's variance estimator is *at least* as large as the actual variance.

$$
\text{Var}_W (\overline{Y}^{\text{obs}}_{t} - \overline{Y}^{\text{obs}}_{c}) = \frac{S_t^2}{N_t} + \frac{S_c^2}{N_c} - \frac{S_{tc}^2}{N}.
$$

# Asymptotics

---

### A CI with Conservative Coverage Asymptotically

In Section 3 (pp. 1763-1763), Li and Deng (2017) provide a finite version of the CLT, so long as the potential outcomes are ``not too weird.''

A *conservative* $(1 - \alpha) \times 100\%$ CI is $\widehat{\text{ATE}} \pm \Phi^{-1}\left( 1 - \frac{\alpha}{2} \right) \cdot \sqrt{\widehat{\text{Var}} \left( \widehat{\text{ATE}} \right)}$. 

For a 90% confidence interval, this would be $\widehat{\text{ATE}} \pm 1.96 \cdot \sqrt{\widehat{\text{Var}} \left( \widehat{\text{ATE}} \right)}$.

Asymptotically, this interval has coverage of *at least* $(1 - \alpha) \times 100\%$.

# Best Practices

### Best Practices

**Estimation**

- It turns out that OLS with robust standard errors (HC2) is equivalent to the different-in-means and Neyman variance estimator we proposed above. See Samii and Aronow (2016). 
- In practice, we obtain estimates with something like the following:

```{r, eval = FALSE}
fit <- lm(y ~ treatment_indicator, data = df)
V_hat <- sandwich::vcovHC(fit, type = "HC2")
```

**Adjustment**

- Some authors are critical of regression adjustment for experimental data (Freedman 2008).
- However, in my view the dangers are mostly theoretical. In real world problems, regression adjust will almost always give you better estimates, sometimes *much* better estimates
   - Simply include controls.
   - Include mean-centered controls fully interacted with treatment.
   - Select controls with LASSO.
   
# Example

The Coversight Experiment

\footnotesize

```{r}
# load packages
library(tidyverse)

# read cleaned data from dropbox
coversight <- read_rds("https://www.dropbox.com/s/9grn8kkb5yzwagx/data.rds?raw=1")%>%
  select(pres_overall, amplify, failure, pid_strength, passed_mvc1) %>%
  filter(passed_mvc1) %>%  # remove respondents who failed MVC 1
  mutate(pid_strength_rs = arm::rescale(pid_strength)) # center partisan strength

# quick look
glimpse(coversight)
```

---

```{r}
# use only data from amplify condition (failure/success varies)
cs_failure <- coversight %>%
  filter(amplify == "Amplify")

table(cs_failure$failure)
```

---

```{r}
# make y for convenience
y <- cs_failure$pres_overall
treat <- as.numeric(cs_failure$failure == "Failure")

# compute point estimate
mean(y[treat == 1]) - mean(y[treat == 0])

# compute variance estimate
var(y[treat == 1])/sum(treat) + var(y[treat == 0])/sum(1 - treat)

# ols with robust ses
fit <- lm(y ~ treat)
coef(fit)["treat"]
diag(sandwich::vcovHC(fit, type = "HC2"))["treat"]
```

---

\small

```{r}
fit1 <- lm(pres_overall ~ failure, data = cs_failure)
fit2 <- lm(pres_overall ~ failure + I(pid_strength_rs + 100), data = cs_failure)
fit3 <- lm(pres_overall ~ failure*pid_strength_rs, data = cs_failure)

modelsummary(list("No Controls" = fit1, 
                  "One Control" = fit2, 
                  "One Control, Interacted" = fit3), 
             vcov = "HC2", 
             gof_map = NA, 
             escape = TRUE, 
             output = "latex")

```


