---
title: "Interaction"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
bibliography: references.bib
citations-hover: true
reference-location: document
warning: false
message: false
cache: false
---

```{r}
#| include: false
library(tidyverse)
```

## Interaction

Interaction means: the effect\* of $X$ on $Y$ depends on the value of some moderating variable $Z$.

<div style="height: 4rem;"></div>


The problem: 

- we can have interaction on the scale of the linear predictor $\eta = \beta_0 + \beta_1 x_1 + ... + \beta_k x_k$.
- we can have interaction on the scale of the outcome $E(Y) = g^{-1}(\eta)$.

<div style="height: 4rem;"></div>


1. It isn't clear which we should care about.
1. One does not imply the other.
1. They can even go in opposite directions.

::: {.aside}

The same sorts of arguments apply to any model with a non-linear link function, but logit serves as a good example.

:::


# Why the Distinction Matters

---

## Latent Scale $\eta_i$

- $\eta_i$ is **unbounded** and **linear** in the parameters.  
- Only the **product term** can create interaction on this scale.  
- For the model $\eta_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \beta_3 (X_i Z_i) + \dots$, interaction only exists when $\frac{\partial^2 \eta_i}{\partial X \partial Z} = \beta_3 \neq 0$.

---

## Probability Scale $\Pr(Y_i=1)$

- The inverse link $g^{-1}(\cdot)$ is **nonlinear** and S-shaped.  
- Even if $\beta_3 = 0$ (or no product term is included), the effect of $X$ on $\Pr(Y)$ varies with $Z$ because $\frac{\partial \Pr(Y_i)}{\partial X} = \frac{d g^{-1}(\eta_i)}{d\eta_i} \cdot \frac{\partial \eta_i}{\partial X}$.
- Since $\frac{d g^{-1}(\eta)}{d\eta}$ depends on $\eta$, any variable that shifts $\eta$ affects instantaneous marginal effects.  
- This follows from **compression**, where effects shrink as $\Pr(Y)$ approaches $0$ or $1$.
- Thus, interaction on the probability scale may occur **with or without** a product term. It can even have the opposite effect as the product term.

---

## Interaction on the Latent Scale

This scale corresponds to hypotheses explicitly about the latent index (e.g., linear effects on an underlying utility or propensity).

This works *just like* linear regression, where $\eta_i = \beta_0 + \beta_1 X_i + \beta_2 Z_i + \beta_3 (X_i Z_i) + \dots$

### Definition

Interaction on the latent scale exists when $\frac{\partial^2 \eta_i}{\partial X \partial Z} = \beta_3 \neq 0$.

### Instantaneous Marginal Effect of $X$ on $\eta_i$

$$
\frac{\partial \eta_i}{\partial X}
  = \beta_1 + \beta_3 Z_i.
$$

### First Difference

For a discrete change in $X$, 

$$
\Delta \eta_i = \beta_3(Z_{hi} - Z_{lo}).
$$

---

## Interaction on the Probability Scale 

We can measure interaction on the probability scale with 

1. second derivatives (cross-partials)
2. second differences 

---

## Approach I: Using Derivatives

### Definition of Interaction I (Derivatives)

Interaction on the probability scale is present when

$$
\frac{\partial^2 \Pr(Y_i)}{\partial X \partial Z} \neq 0.
$$

Depends on the mixture of:

1. Latent interaction ($\beta_3 \neq 0$) 
2. Nonlinearity of the inverse link ($g^{-1}$).

**I don't find these instantaneous marginal effects useful, except when they are constant.**

---

## Approach II: Using Differences

For a discrete change in $X$ at a fixed value of $Z$, define the **first difference** as

$$
\Delta_X(Z) =
\Pr(Y=1 \mid X_{hi}, Z) -
\Pr(Y=1 \mid X_{lo}, Z).
$$

And define the **second difference** as

$$
\Delta\Delta =
\big[ \Pr(Y \mid X_{hi}, Z_{hi}) - \Pr(Y \mid X_{lo}, Z_{hi}) \big]
-
\big[ \Pr(Y \mid X_{hi}, Z_{lo}) - \Pr(Y \mid X_{lo}, Z_{lo}) \big].
$$

### Definition of Interaction II (Differences)**

If $\Delta\Delta \neq 0$, the effect of $X$ changes when $Z$ changes.

---

## Theoretical Rationale

### When to Theorize on the Latent Scale

The theory concerns an **unobserved continuous index** (e.g., utility, evaluation, propensity).  

### When to Theorize on the Probability Scale

The theory concerns **changes in probability**, not the latent index.  

Important: the model's compression *must be part of the theory*---you must theorize interaction *because* of ceiling and floor effects.

---

## My Take {.smaller}

We are left with no good default.

- Theories are rarely precise enough to distinguish between these in any obvious way.
- To the extent they do, they either: 
   - Build in floor or ceiling effects in not-super-interesting ways (e.g., contiguity and democracy).
   - Have an explicit latent utility interpretation (e.g., spatial voting).
   
   

**Two Claims**

These pull us in opposite directions.
   
   
- **Most theories** (in my view) vaguely refer to unbounded notions of the outcome.
- **Most readers** expect estimates on the scale of the outcome (probability scale, in this case).


**Suggestion**

- When you are computing quantities of interest, work on the scale of the outcome *unless you have a great reason to do otherwise* (e.g., spatial models).
- When you are theorizing, remain mindful of compression. Go out of your way to explicitly think about what happens as events become very likely or unlikely.

# Example

In order to turn out to vote, you need to feel like you have a stake in society. 

There are two ways to obtain this feeling:

1. A **financial stake**.
2. A **civic stake**, mainly developed from education.

If you lack a stake in society, then you tend not to vote. If you have a stake in society, then you tend to vote. Importantly, these two stakes are substitutes---you just need one.

---

Load our data.

```{r}
turnout <- ZeligData::turnout |>
  glimpse()
```

And fit two models.

```{r}
# no product term; scale sets mean = 0 and SD = 1
f <- vote ~ educate + income
fit <- glm(f, data = turnout, family = binomial)

# product term
f2 <- vote ~ educate*income
fit2 <- glm(f2, data = turnout, family = binomial)
```

---

```{r}
library(marginaleffects)

grid <- crossing(educate = unique(turnout$educate), 
                 income = unique(turnout$income)) |>
  glimpse()
```

---

```{r}
p <- predictions(fit2, newdata = grid) |>
  glimpse()
```

--- 

```{r}
ggplot(p, aes(x = income, y = estimate, color = educate, group = educate)) + 
  geom_line()
```

---

```{r}
# effect of educate moving from 25th to 75th percentile
c <- comparisons(fit2, 
                 variables = list(educate = "iqr"),
                 newdata = datagrid(income = unique)) |>
  glimpse()
```

---

```{r}
ggplot(c, aes(x = income, y = estimate, ymin = conf.low, ymax = conf.high)) + 
  geom_ribbon(fill = "grey80") + 
  geom_line()
```
---

## A Formal Test

We need to actually compute the second difference and test whether it's difference from zero.

- Confidence interval.
- *p*-value

```{r}
p <- predictions(fit2, 
                 variables = list(educate = "iqr", income = "iqr"),
                 newdata = datagrid()) |>
  glimpse()
```

---

```{r}
p |> 
  select(educate, income, estimate, conf.low, conf.high) |>
  tinytable::tt()
```


```{r}
# compute the ci and p-value for the 2nd difference
# this is a formal test for interaction on the probability scale
hypotheses(p, hypothesis = "(b4 - b2) - (b3 - b1) = 0")
```

::: {.callout-warning}
`"b4"` isn't the hi-hi-scenario. It's just the 4th row in `p`. If the rows in `p` change, so to does the meaning of `"b4"`. Be careful!
:::


## But...

Here is the test for the model *without* a product term.

```{r}
p <- predictions(fit, 
                 variables = list(educate = "iqr", income = "iqr"),
                 newdata = datagrid())

hypotheses(p, hypothesis = "(b4 - b2) - (b3 - b1) = 0")
```

**How can this happen?!?**

## Also but...

```{r}

modelsummary::modelsummary(list("No Prod." = fit, "Prod" = fit2))
```