---
title: "Hierarchical Models"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    mathjax: true  
execute:
  echo: true
fig-align: center
fig-width: 8
fig-asp: 0.5
---


# Opening Comments

- Exam
- Workshop
- Paper

```{r echo=FALSE}
#| message: false
#| warning: false
options(width = 90)
options(digits = 3)

library(tidyverse)
theme_set(theme_minimal())
```

---

Here are the packages we're using:

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(marginaleffects)
library(modelsummary)

# for fitting hierarchical models
library(lme4)
library(brms)  # see also rstanarm::stan_lmer()

# for post-processing simulations
library(tidybayes)
```


---

This week, I introduce the idea of a hierarchical model. 

::: {.fragment .custom .blur}
These fall squarely within our probability modeling framework---but with an exciting new twist.
:::

::: {.fragment .custom .blur}
Before: $x_i \sim f(x_i; \theta)$
:::


::: {.fragment .custom .blur}
Now: $x_i \sim f(x_i; \theta_{j[i]})$, where $\theta_j \sim f(\theta_j; \phi)$
:::

# Contact in Finland

In my dissertation, I was interested in how survey self-reports of **being contacted** by a political party varies across electoral districts. 


You can see [here](http://www.carlislerainey.com/papers/stratmob.pdf) for the details. 

---

For now, let's just estimate the proportion of people in each district that are contacted by a party.

::: {.fragment .custom .blur}
Let's start with Finland.
:::

::: {.fragment .custom .blur}
```{r}
# load data
finland <- crdata::finland |>
  glimpse()
```
:::

::: {.fragment .custom .blur}
It will be convenient in later steps to have `contacted` be an indicator and `district_id` be a nice factor, so let's do that now.

1. Remove "District" from the strings.
1. `reorder()` the levels by the average of `contacted`. (Important sneaky trick).
:::

---

```{r}
# load data
finland <- crdata::finland |>
  # make contacted an indicator
  mutate(contacted = as.numeric(contacted == "Contacted")) |>
  # mutate district_id to factor ordered by contact rate
  mutate(district_id = str_remove(district_id, "District "),
         district_id = reorder(district_id, contacted)) |>
  glimpse()
```

---

A scatterplot is a safe and potentially informative way to start.

```{r}
gg1 <- ggplot(finland, aes(x = district_id, y = contacted)) + 
  geom_jitter(shape = 21, alpha = 0.5, size = 1)
print(gg1)
```

---

Now let's estimate the probability of being contacted by a political party in each district.

::: {.fragment .custom .blur}
Start by writing down a probability model for being contacted. This is a binary outcome, so if we're going to assume (conditional) independence, it must be Bernoulli.
:::

::: {.fragment .custom .blur}
$$
\text{Contacted}_i \sim \text{Bernoulli}(\pi_j),
$$

where $i$ indexes the respondent and $j$ indexes the electoral district.
:::

::: {.fragment .custom .blur}
Remember that the ML estimate of $\pi$ is just the sample mean (or sample proportion). 
:::

---

We can do this easily for each district with `group_by()` and `summarize()`. 

::: {.fragment .custom .blur}
Let's also add an estimate of the standard error using $\widehat{\text{SE}}(\hat{\pi}_j) = \frac{\sqrt{\hat{\pi}_j(1 - \hat{\pi}_j)}}{\sqrt{n}}$.
:::

::: {.fragment .custom .blur}
```{r}
estimates <- finland %>%
  group_by(district_id) %>%
  summarize(pi_hat = mean(contacted),
            n = n(),
            se_hat = sqrt(pi_hat*(1 - pi_hat))/sqrt(n)) %>%
  glimpse()
```
:::

---

Now let's add these estimates to the scatterplot.

```{r}
gg2 <- gg1 + 
  geom_errorbar(data = estimates, aes(ymin = pi_hat - 1.64*se_hat,
                                      ymax = pi_hat + 1.64*se_hat, 
                                      y = NULL),
                width = .2, color = "#1b9e77") + 
  geom_point(data = estimates, aes(y = pi_hat), 
             color = "#1b9e77", size = 2)
print(gg2)
```


---

You can see that the confidence intervals (and sample size) varies by district, so let's add that information to the plot.

::: {.fragment .custom .blur}
```{r}
#| fig-asp: 0.5
gg3 <- gg2 + 
  geom_label(data = estimates, aes(y = pi_hat + 1.64*se_hat,
                                   label = n),
                color = "#1b9e77",
             vjust = -0.2)
print(gg3)
```
:::


---

Great! Now we're done. We've got ML estimates of the contact rate for all districts. 

::: {.fragment .custom .blur}
But wait... can we do better?
:::

---

Let's look at district 309 on the far right. We our estimate of the contact rate in this district is the highest.

```{r}
#| fig-asp: 0.5
#| echo: false
print(gg3)
```

::: {.fragment .custom .blur}
Our estimator $\hat\pi$ is unbiased, but **if you had to say**, would you say that the turnout rate in district 309 was higher or lower than $\hat{\pi}_{309} = 0.40$?
:::

::: {.fragment .custom .blur}
If your intuition matches mine, you will say *lower*. Why? This "why?" is an important question.
:::

---

If you think that $\hat{\pi}$ is a good estimator, but you suddenly want to adjust this particular estimate, you must have additional information coming from somewhere. Where?

::: {.fragment .custom .blur}
The answer: the *other districts*. We admit that the contact rate will be different across districts, but it also seems reasonable to assume they will be similar (that's "different, but similar"). Given that 309 has a small sample and looks relatively unlike the other estimates, we want to pull it down *slightly* toward the other estimates.
:::

::: {.fragment .custom .blur}
This is exactly the intuition of hierarchical models. Let's write down this "different, but similar" intuition. 
:::

::: {.fragment .custom .blur}
To do this, it will be convenient to have our $\pi_j$s on the logit scale.
:::

---

Start with the likelihood.

$$
\text{Contacted}_i \sim \text{Bernoulli}(\pi_j)
$$

::: {.fragment .custom .blur}
Now translate the $\pi$s to the logit scale.

$$
\pi_j = \text{logit}^{-1}(\alpha_j)
$$
:::

::: {.fragment .custom .blur}
Each of these $\alpha_j$s is like a constant term in a logit model with no covariates. (We could add covariates, but that makes the example less clear.)
:::

---

Now for the hierarchical part: we want to capture the notion that these $\alpha_j$s are different *but similar*. 

::: {.fragment .custom .blur}
We can do that by modeling them as draws from normal distribution.

$$
\alpha_j \sim N(\mu, \sigma^2)
$$
:::

::: {.fragment .custom .blur}
- We can estimate $\mu$ (the "typical" $\alpha$) and $\sigma$ (the degree of similarity) from the data. This is "hierarchical piece."
:::

::: {.fragment .custom .blur}
- We sometimes refer to batches of parameters modeled as draws from a distribution as "random effects."
:::

---

To estimate the hierarchical model: 

- use ML via the `lme4::lmer()`/`lme4::glmer()` function. Fast, fragile, approximate.
- use MCMC via the `brm()` function. Slow(er), robust, exact.^[Also `rstanarm::stan_lmer()`.]

::: {.fragment .custom .blur}
We're going to start with `lmer()` because I don't want to juggle posterior simulation and hierarchical models simultaneously, but you should always use `brm()` always unless you simply need a quick fit.
:::

---

We can fit the model with `glmer()`.

```{r}
#| code-line-numbers: "|2"
library(lme4)
f <- contacted ~ (1 | district_id)
fit <- glmer(f, data = finland, family = binomial)
```

::: {.fragment .custom .blur}

And we can summarize the results.

```{r}
arm::display(fit)
```

From the output above, we estimate $\mu$ as -1.37 and $\sigma$ as 0.31. That is, the $\alpha_j$s are about -1.37 give-or-take 0.31 or so.
:::

---

But here's the useful part: we can take this $\hat{\mu}$ and $\hat{\sigma}$ and use them as a Bayesian prior for the individual $\alpha_j$s. 

::: {.fragment .custom .blur}
Each $\alpha_j$ has a likelihood (using the data from that district) and a prior (using $\hat{\mu}$ and $\hat{\sigma}$).
:::

::: {.fragment .custom .blur}
```{r}
re <- ranef(fit)$district_id[, "(Intercept)"] # this is centered at zero (i.e., alpha - mu); random part
fe <- fixef(fit)["(Intercept)"]  # mu; fixed part

alpha_hat <- fe + re  # district-level linear predictors
pi_hat <- plogis(fe + re)  # district-level probabilities
```
:::

---

In this case, it's easy to use `predict()` to get this same estimates. (The `predict()` approach is a little safer because it's easier to keep the `pi_hat` and the `district_id` label together.)

```{r}
hier_estimates <- tibble(district_id = unique(finland$district_id)) %>%
  mutate(pi_hat = predict(fit, newdata = ., type = "response")) %>%
  arrange(district_id) %>%
  glimpse()
```

---

We can also do this with \{marginaleffects\}.

```{r}
predictions(fit,
            newdata = datagrid(district_id = unique)) |>
  select(district_id, estimate) |>
  arrange(district_id) |>  # comparisons() arranges predictions differently
  glimpse()
```

---

We can add this to our plot to see how the estimates changed. 

```{r}
gg4 <- gg3 + 
  geom_point(data = hier_estimates,
             aes(y = pi_hat), 
             color = "#d95f02",
             shape = 10, size = 3)
print(gg4)
```

::: {.aside}
- Notice that our new estimate for 309 is now at the very bottom of our initial confidence interval.
- Notice the differences between the original ML estimates and CIs and the new estimates from the hierarchical model.
:::

---

We can compare the initial ML estimates to the hierarchical model estimates.

```{r}
comb <- bind_rows(list("initial ML" = estimates,
                       "hierarchical" = hier_estimates), 
                  .id = "type") %>%
  glimpse()
```

---

```{r}
gg_comb <- ggplot(comb, aes(x = pi_hat, y = type)) + 
  geom_point(alpha = 0.5) + 
  geom_line(aes(group = district_id), alpha = 0.5)
print(gg_comb)
```

---

It's helpful to think of these approaches in terms of the amount of information they borrow across groups.

::: {.fragment .fade-up}
1. **No Pooling.** Assume the groups are different, and that nothing can be learned from the other groups. Estimate each group separately. 
:::
::: {.fragment .fade-up}
2. **Complete Pooling**. Assume the groups are identical. Collapse all the groups together to estimate the single, common estimate.
:::
::: {.fragment .fade-up}
3. **Partial Pooling**. A compromise between the two extremes above. Assume the groups are different, *but similar*. Estimate the amount of similarity using the data, and pool the estimates toward a common estimate accordingly.
:::

---

We did not implement complete pooling above, but we can do it here to compare the three approaches. 

```{r}
# use the overall rate for each district
common_estimate <- tibble(district_id = unique(finland$district_id)) %>%
  mutate(pi_hat = mean(finland$contacted)) %>%
  glimpse()
```

---

```{r}
# combine the three types
comb <- bind_rows(list("No Pooling" = estimates,
                       "Partial Pooling" = hier_estimates,
                       "Complete Pooling" = common_estimate), 
                  .id = "type") %>%
  mutate(type = reorder(type, pi_hat, var)) %>%
  glimpse()
```

---

```{r}
gg_comb2 <- ggplot(comb, aes(x = pi_hat, y = type)) + 
  geom_point(alpha = 0.5) + 
  geom_line(aes(group = district_id), alpha = 0.5)
print(gg_comb2)  
```

# Key things

::: {.fragment .custom .blur}
First, the 309 problem.
:::

::: {.fragment .custom .blur}
Second, $\mu_i = \alpha_{j[i]} + \beta x_i$, where $\alpha_j \sim N(0, \sigma^2_\alpha)$.
:::

::: {.fragment .custom .blur}
Third, the figure below.

```{r}
#| fig-asp: 0.5
#| echo: false
print(gg_comb2)  
```
:::


# We stopped here on 10/21; we'll pick up here next time.

# Contact in the UK

```{r}
uk <- crdata::uk |>
  # make contacted an indicator
  mutate(contacted = as.numeric(contacted == "Contacted")) |>
  # mutate district_id to factor ordered by contact rate
  mutate(district_id = str_remove(district_id, "District "), 
         district_id = reorder(district_id, contacted)) |>
  glimpse()
```

---

```{r}
gg1_uk <- ggplot(uk, aes(x = district_id, y = contacted)) + 
  geom_jitter(shape = 21, alpha = 0.5, size = 1)
print(gg1_uk)
```

---

```{r}
estimates <- uk %>%
  group_by(district_id) %>%
  summarize(pi_hat = mean(contacted),
            n = n(),
            se_hat = sqrt(pi_hat*(1 - pi_hat))/sqrt(n)) %>%
  glimpse()
```

---


```{r}
#| fig-asp: 0.5
gg2_uk <- gg1_uk + 
  geom_errorbar(data = estimates, aes(ymin = pi_hat - 1.64*se_hat,
                                      ymax = pi_hat + 1.64*se_hat, 
                                      y = NULL),
                width = .2, color = "#1b9e77") + 
  geom_point(data = estimates, aes(y = pi_hat), 
             color = "#1b9e77", size = 2)
print(gg2_uk)
```

---

```{r}
no_pooling <-  uk %>%
  group_by(district_id) %>%
  summarize(pi_hat = mean(contacted))  %>%
  glimpse()
```

---


```{r}
complete_pooling <- tibble(district_id = unique(uk$district_id)) %>%
  mutate(pi_hat = mean(uk$contacted)) %>%
  glimpse()
```

---

```{r}
#| code-line-numbers: "|1"
f <- contacted ~ (1 | district_id)
fit <- glmer(f, data = uk, family = binomial)

arm::display(fit)
```

---

```{r}
partial_pooling <- tibble(district_id = unique(uk$district_id)) %>%
  mutate(pi_hat = predict(fit, newdata = ., type = "response")) %>%
  arrange(district_id) %>%
  glimpse()
```

---

```{r}
comb <- bind_rows(list("No Pooling" = no_pooling,
                       "Partial Pooling" = partial_pooling,
                       "Complete Pooling" = complete_pooling), 
                  .id = "type") %>%
  mutate(type = reorder(type, pi_hat, var)) %>%
  arrange(district_id) %>%
  glimpse()
```

---

```{r}
#| fig-asp: 0.5
ggplot(comb, aes(x = pi_hat, y = type)) + 
  geom_point(alpha = 0.5) + 
  geom_line(aes(group = district_id), alpha = 0.5)
```

# Contact in the UK (w/ predictors!)

We can also add predictors of the random parameters, such as the competitiveness of the district.

```{r}
glimpse(uk)
```

---

```{r}
uk_districts <- uk %>% 
  select(district_id, district_competitiveness) %>%
  distinct() %>%
  glimpse()

w_comp <- partial_pooling |>
  left_join(uk_districts) |>
  glimpse()
```

---

```{r}
ggplot(w_comp, aes(x = district_competitiveness, y = pi_hat)) + 
  geom_point() 
```

---

```{r}
ggplot(w_comp, aes(x = district_competitiveness, y = pi_hat)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE, color = "#1b9e77")
```

---

Hey! We should build this directly into our model.

```{r}
#| code-line-numbers: "|1"
contacted ~ district_competitiveness + (1 | district_id)
fit <- glmer(f, data = uk, family = binomial)

arm::display(fit)
```

---

```{r}
partial_pooling <- uk %>%
  select(district_id, district_competitiveness) %>%
  distinct() %>%
  mutate(pi_hat = predict(fit, newdata = ., type = "response")) %>%
  glimpse()
```

---

```{r}
ggplot(partial_pooling, aes(x = district_competitiveness, y = pi_hat)) + 
  geom_point()
```

---

```{r}
comb <- bind_rows(list("No Pooling" = no_pooling,
                       "Partial Pooling" = partial_pooling,
                       "Complete Pooling" = complete_pooling), 
                  .id = "type") %>%
  mutate(type = reorder(type, pi_hat, var)) %>%
  glimpse()
```
  
---  

```{r}
ggplot(comb, aes(x = pi_hat, y = type)) + 
  geom_point(alpha = 0.5) + 
  geom_line(aes(group = district_id), alpha = 0.5)
```

# A simulation exercise in R

# Contact in Several Countries

$$
\text{Pr}(\text{Contacted}_{i} = 1) = \text{logit}^{-1}(\alpha_{jk} + {\bf X}_i\beta)
$$
$$
\alpha_{jk} \sim N(\gamma_{0k} + \gamma_{1k}\text{Competitiveness}_j, \sigma^2_{\alpha}) \text{ for } j = 1, 2,..., J
$$

$$
\begin{pmatrix} \gamma_{0k} \\ \gamma_{1k} \end{pmatrix} 
\sim 
{\large} N\begin{pmatrix}  \begin{pmatrix} \mu_{\gamma_{0}} \\ \mu_{\gamma_{1}} \end{pmatrix}, 
\begin{pmatrix}         \sigma^2_{\gamma_{0}}                                           &                 \rho\sigma_{\gamma_{0}}\sigma_{\gamma_{1}} \\
\rho\sigma_{\gamma_{0}}\sigma_{\gamma_{1}}         &                \sigma^2_{\gamma_{1}}
\end{pmatrix}\end{pmatrix} \text{ , for }k = 1, 2, ..., K
$$

$$
\mu_{\gamma_{0}} = \delta_{00} + \delta_{01}\text{Proportional Rules}_k
$$

$$
\mu_{\gamma_{1}} = \delta_{10} + \delta_{11}\text{Proportional Rules}_k
$$

---
  
```{r}
rainey <- crdata::rainey2015 |>
  glimpse()
```

---

We can fit this model with `lmer()`, but this *relatively small* model is too big or complex. 

::: {.fragment .custom .blur}

```{r}
#| code-line-numbers: "|2,3,4"
# model is fragile
f <- contacted ~ district_competitiveness*electoral_rules +
               (1 | district_id) +
               (1 + district_competitiveness | election_id)
fit <- glmer(f, data = rainey, family = binomial)
```
:::

::: {.fragment .custom .blur}
Stan would have no problems with this one...
:::

---

And show the estimates.

```{r}
arm::display(fit, detail = TRUE)
```

---

Using the `lmer()`-like formula, we can fit the hiearchical model with Stan.
  
```{r, results='hide', cache = TRUE}
#| label: "fit-brm-model"
start_time <- Sys.time()
stan_fit <- brm(contacted ~ district_competitiveness*electoral_rules +
                  (1 | district_id) + 
                  (1 + district_competitiveness | election_id),
                data = rainey, 
                family = bernoulli, 
                chains = 3,
                cores = 3,
                iter = 1000, 
                backend = "cmdstanr")
end_time <- Sys.time()
```

::: {.fragment .custom .blur}

And it fits in less than 10 minutes. This took hours or days in 2012 with JAGS.

```{r}
end_time - start_time
```

:::

---

```{r}
print(stan_fit)
```

---
  
```{r}
summary(stan_fit)
```

---

```{r}
#| warning: false
#| message: false

# create grid
grid <- rainey |>
  select(district_id, 
         election_id, 
         district_competitiveness, 
         electoral_rules) |>
  distinct() |>
  glimpse()
```

::: {.fragment .custom .blur}
```{r}  
# compute posterior average for each district
pred_district <- grid |>  
  add_epred_draws(stan_fit) %>%
  summarize(post_avg = mean(.epred)) %>%
  glimpse()
```
:::

---

```{r}
#| warning: false
#| message: false

# compute posterior average for each country (in a 'typical' district)
pred_country <- rainey %>%
  select(district_id, 
         election_id, 
         district_competitiveness, 
         electoral_rules) %>%
  distinct() %>%
  add_epred_draws(stan_fit, 
                  re_formula = ~ (1 + district_competitiveness | election_id)) %>%
  summarize(post_avg = mean(.epred)) %>%
  glimpse()
```

---
  
```{r}
#| warning: false
#| message: false

ggplot(pred_district, aes(x = district_competitiveness, y = post_avg)) +
  facet_wrap(vars(election_id)) +
  geom_point() +
  geom_line(data = pred_country, color = "red", size = 1.5)
```

---

```{r}
grid <- rainey %>%
  select(district_id, 
         election_id, 
         district_competitiveness, 
         electoral_rules)

p <- predictions(stan_fit,
                 newdata = grid) |>
  glimpse()
```

---

```{r}
p_typical <- predictions(
  stan_fit,
  newdata = grid, 
  re_formula = ~ (1 + district_competitiveness | election_id)
  ) |>
  glimpse()
```

---

```{r}
ggplot(p, aes(x = district_competitiveness, y = estimate)) +
  facet_wrap(vars(election_id)) +
  geom_point() +
  geom_line(data = p_typical, color = "red", size = 1.5)
```
