---
title: "Ideal Points and MRP"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
execute:
  echo: true
fig-align: center
fig-width: 8
fig-asp: 0.5
---

# Opening Comments

- Exam
- Workshop
- Paper

```{r echo=FALSE}
#| message: false
#| warning: false
options(width = 90)
options(digits = 3)

library(tidyverse)
theme_set(theme_minimal())
```

---

Here are the packages we're using:

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(marginaleffects)
library(modelsummary)
library(ggrepel)

# for fitting hierarchical models
library(lme4)
library(brms)  # see also rstanarm::stan_lmer()
library(rstan)

# for post-processing simulations
library(tidybayes)

# handle future conflicts
# - MASS v. dplyr for filter()
conflicted::conflict_prefer("filter", "dplyr")
# - MASS v. dplyr for select()
conflicted::conflict_prefer("select", "dplyr")

```


# Ideal Point Models

## `SupremeCourt` data

```{r}
#| message: false
#| warning: false
sc <- read_csv("https://pos5747.github.io/data/sc2000.csv") |>
glimpse()
```

## Summarize

```{r}
sc |> 
  group_by(justice, appointed_by) |>
  summarize(prop_cons_votes = mean(cons_vote, na.rm = TRUE)) |>
  arrange(prop_cons_votes)
```

## Plot 1

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.5

sc_summary <- sc |> 
  group_by(justice, appointed_by) |>
  summarize(prop_cons_votes = mean(cons_vote, na.rm = TRUE)) |>
  ungroup() |>
  mutate(justice = reorder(justice, prop_cons_votes)) |>
  arrange(prop_cons_votes)

ggplot(sc_summary, aes(x = justice, y = prop_cons_votes)) + 
  geom_col(width = 0.5) + 
  geom_label(aes(label = scales::percent(prop_cons_votes, accuracy = 1)), hjust = 1.2) + 
  coord_flip()
```

## Plot 2

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4
library(ggrepel)

set.seed(1234)
ggplot(sc_summary, aes(x = prop_cons_votes, y = 0)) +
  geom_hline(yintercept = 0) +
  geom_point(size = 3) +
  geom_label_repel(
    aes(label = justice),
    nudge_y = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "y",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Proportion of Conserative Votes"  
    ) + 
  scale_y_continuous(limits = c(-.1, .1))
```

## Estimate "abilities" of justices and "difficulties" of cases

-   ability is analagous to a justice's ideology.
-   difficulty is analogous to a case's ideology.

```{r}
f <- cons_vote ~ (1 | justice) + (1 | case_id)
fit <- glmer(f, data = sc, family = binomial)
```

## Print the "abilities"

```{r}
ranef(fit)$justice |>
  as_tibble(rownames = "justice") |>
  rename(ideal_point = `(Intercept)`) |>
  arrange(ideal_point)
```

## Plot the abilities

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4

ip <- ranef(fit)$justice |>
  as_tibble(rownames = "justice") |>
  rename(ideal_point = `(Intercept)`) |>
  arrange(ideal_point)

set.seed(1234)
ggplot(ip, aes(x = ideal_point, y = 0)) +
  geom_hline(yintercept = 0) +
  geom_point(size = 3) +
  geom_label_repel(
    aes(label = justice),
    nudge_y = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "y",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Ideal Point"  
    ) + 
  scale_y_continuous(limits = c(-.1, .1))
```

## Add cases

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4

library(ggbeeswarm)

cases <- data.frame(case = as.numeric(ranef(fit)$case_id[, "(Intercept)"])) 

set.seed(1234)
ggplot(ip, aes(y = ideal_point, x = 0)) +
  geom_vline(xintercept = 0) +
  geom_beeswarm(data = cases, aes( y = case, x = 0), shape = 18, alpha = 0.3, color = "red") + 
  geom_beeswarm(size = 2, alpha = 0.5) +
  geom_label_repel(
    aes(label = justice),
    nudge_x = rep_len(c(-1, 1)*0.035, 9),
    segment.color = "grey50",
    direction = "x",
    point.padding = 0.5, 
    min.segment.length = 0.0
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  labs(
    x = "Ideal Point"  
    ) + 
  scale_x_continuous(limits = c(-.1, .1)) + 
  coord_flip()
```

## Plot the logit curve

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| fig-asp: 0.4
mu <- as.numeric(fixef(fit))

beta <- ranef(fit)$case_id |>
  as_tibble(rownames = "case_id") |>
  rename(beta = `(Intercept)`)

cases_to_keep <- c("Case 20", "Case 5", "Case 2", "Case 42", "Case 22", "Case 33")

pred_lines <- crossing(justice_ideology = seq(-2, 2, length.out = 100), 
                      case_id = unique(sc$case_id)) |>
  left_join(beta) |>
  mutate(pr_cons_vote = plogis(mu + justice_ideology - beta)) |>
  mutate(case_id = reorder(case_id, pr_cons_vote)) |>
  filter(case_id %in% cases_to_keep)

pred_points <- sc |>
  left_join(beta) |> 
  left_join(ip) |>
  mutate(pr_cons_vote = plogis(mu + ideal_point - beta)) |>
  mutate(case_id = reorder(case_id, pr_cons_vote)) |>
    filter(case_id %in% cases_to_keep)

ggplot(pred_lines, aes(x = justice_ideology, y = pr_cons_vote)) + 
  geom_vline(xintercept = ip$ideal_point, color = "grey50", linewidth = 0.2) + 
  geom_line() + 
  geom_point(data = pred_points, aes(x = ideal_point, y = pr_cons_vote, shape = factor(cons_vote)), fill = "white") + 
  facet_wrap(vars(case_id)) + 
  scale_shape_manual(values = c(19, 21)) + 
  labs(subtitle = "Selected cases", 
       shape = "Conservative Vote", 
       y = "Probability of a Conservative Vote", 
       x = "Justice Ideal Point")
```

## Exercise

Fit a 2-PL model to these data.


# MRP

## MRP

-   "Multilevel regression with poststratification" or "Mister P"
-   We have sufficiently large polls to estimate national opinion well, but how can we use these to estimate state-level opinion well across the fifty states?

## Two-step process

1.  Use a **national survey data** (e.g., N = 2,000) and **a hierarchical model** to estimate public opinion within small subgroups defined by state and demographic variables. *Example: Men age 18-29 without military experience and no HS degree living in Alabama.*
2.  Look up the fraction of each state that falls into each demographic category. Weight the predictions for each state-demographic groups by these fractions to obtain an estimate of state-level opion.

## Two-step process: Step 2(a)

Find the populations of demographic subgroups in each state.

This might come from the Census or CPS.

Here's an example [packaged with](https://marginaleffects.com/chapters/mrp.html) {marginaleffects}.

```{r}
demographics <- get_dataset("ces_demographics") |>
  glimpse()
```

## Two-step process: Step 1

Use a **national survey data** (e.g., N = 2,000) and a hierarchical model to estimate public opinion within each each subgroup.

```{r}
survey <- get_dataset("ces_survey") |>
  glimpse()
```

## Two-step process: Step 2(b)

Fit the model.

```{r}
#| label: mrp-model
#| message: false
#| warning: false
#| cache: true
#| results: hide
# fit model
f <- defund ~ age + education + military + gender + (1 + gender | state)
model <- brm(f, family = bernoulli, data = survey, chains = 4, cores = 4)
```

## Two-step process: Step 2(c)

Create the predictions.

```{r}
p <- predictions(model, newdata = demographics) |>
  glimpse()
```

## Two-step process: Step 2(d)

Weight the predictions.

```{r}
ests <- p |>
  mutate(w_ests = estimate * percent) |>
  group_by(state) |>
  summarize(mrp_est = sum(w_ests)) |>
  ungroup() |>
  glimpse()
```

## Plot

```{r}
gg_ests <- ests |>
  mutate(state = reorder(state, mrp_est))
ggplot(gg_ests, aes(x = mrp_est, y = state)) + 
  geom_point()
```

## Question

We used this model. What is it? What others might we have used?

```         
defund ~ age + education + military + gender + (1 + gender | state)
```
