---
title: "IRT Models"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
execute:
  echo: true
fig-align: center
fig-width: 8
fig-asp: 0.5
---


# Opening Comments

- Exam
- Workshop
- Paper

```{r echo=FALSE}
#| message: false
#| warning: false
options(width = 90)
options(digits = 3)

library(tidyverse)
theme_set(theme_minimal())
```

---

Here are the packages we're using:

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(marginaleffects)
library(modelsummary)

# for fitting hierarchical models
library(lme4)
library(brms)  # see also rstanarm::stan_lmer()
library(rstan)

# for post-processing simulations
library(tidybayes)
```


---

This week, I introduce the idea of a IRT model. 

This is just *one specific type* of hierarchical model.

---

The `answers` dataset has answers from hypothetical students to real questions from a recent exam.

- `question_id_num` indexes the 55 questions on the exam.
- `student_id_num` idexes the 100 students who took the exam.
- `correct_answer` indicates (0/1) where the student answered the question correctly.

```{r}
answers <- read_csv("https://pos5747.github.io/data/example-answers.csv") |>
  glimpse()
```

---

```{r}
grades <- answers %>% 
  group_by(student_id_num) %>%
  summarize(proportion_correct = mean(correct_answer)) %>%
  mutate(
    letter_grade = cut(
      proportion_correct,
      breaks = c(-Inf, 0.60, 0.63, 0.67, 0.70, 0.73, 0.77, 0.80, 0.83, 0.87, 0.90, 0.93, Inf),
      labels = c("F",  "D-", "D",  "D+", "C-", "C",  "C+", "B-", "B",  "B+", "A-", "A"),
      ordered_result = TRUE
    )
  )

grades |> filter(student_id_num <= 10)
```

---

## Rasch / 1-PL model {.smaller}

*Indices*

- Answers: $n = 1, \dots, N$. <span class="gray-text">($N = 55 \times 100 = 5{,}500$ in our data.)</span>  
- Students: $j = 1, \dots, J$. <span class="gray-text">($J = 100$ in our data.)</span> 
- Questions: $k = 1, \dots, K$. <span class="gray-text">($K = 55$ in our data.)</span> 


*Outcome*: $y_{n} \in \{0, 1\}$ indicates whether the $n$th answer is correct.

$$
y_n \sim \mathrm{Bernoulli}(\pi_n) \\
\pi_n = \text{logit}^{-1} \left( \mu + \alpha_{j[n]} - \beta_{k[n]} \right).
$$

*Parameters*

- $\pi_n$: the probability that answer $n$ is correct
- $\alpha_{j[n]}$: the **ability** of student $j$ [who is giving the $n$th answer]. 
   - Higher $\alpha_j$ ⇒ more likely to answer correctly; more knowledgeable student.
- $\beta_{k[n]}$: the **difficulty** of question $k$ [which is associated with the $n$th answer]. 
   - Higher $\beta_k$ ⇒ less likely to answer correctly; harder item.
- $\mu$: average difficulty across questions

---

## Hierarchical Model

$$
\alpha_j \sim N(0,\sigma_\alpha^2)\\
\beta_k \sim N(0,\sigma_\beta^2)
$$

## Priors 

$$
\sigma_\alpha \sim \mathrm{half\text{-}Cauchy}(0,1)\\
\sigma_\beta \sim \mathrm{half\text{-}Cauchy}(0,1)
$$

$$
\mu \sim \mathrm{Cauchy}(0,1).
$$

::: {.aside}
- A change from 0 to 1 on the logit scale increase the chance of success from 50% to 73%. A change from 1 to 2 increases the chance of success from 73% to 88%.
```{r}
plogis(0); plogis(1); plogis(2)
```
- $\sigma_\alpha = 1$ means that the "give or take" on the abilities is about 1. 
- Suppose an average student has a 73% change of answering a given question correctly. If $\sigma_\alpha = 1$, then "most" (68%) students have between a 50% and 88% chance of answering correctly. This seems "about right" to me, so I expect $\sigma_\alpha$ to be about 1 (i.e., not 0.1 or 10). 
- The median of a half-Cauchy is the scale, so I set the scale to the $\sigma_\alpha$ that seems "about right." 
- This is a somewhat aggressive prior that polls the variances toward zero---I want to treat bias the model toward treating students and questions as "similar". Setting the scale to 2.5 or 5 is more conservative.
:::


---

```{stan}
#| output.var: irt
#| eval: false
data {
  int<lower=1> N;  // number of answers
  int<lower=1> J;  // number of students
  int<lower=1> K;  // number of questions
  int<lower=1, upper=J> jj[N];  // student index for answer n
  int<lower=1, upper=K> kk[N];  // question index for answer n
  int<lower=0, upper=1> y[N];  // answer n is correct (0/1)
}

parameters {
  real mu;  // mean difficulty
  vector[J] alpha;  // abilities; one per student
  vector[K] beta;  // difficulties; one per question
  real<lower=0> sigma_beta;  // SD of difficulties
  real<lower=0> sigma_alpha;  // SD of abilities
}

model {
  sigma_beta ~ cauchy(0, 1);  // prior for SD of difficulties
  sigma_alpha ~ cauchy(0, 1);  // prior for SD of abilities
  mu ~ cauchy(0, 1);  // prior for mean difficulty
  alpha ~ normal(0, sigma_alpha);  // abilities
  beta ~ normal(0, sigma_beta);  // difficulties
  for (n in 1:N) {
    y[n] ~ bernoulli_logit(mu + alpha[jj[n]] - beta[kk[n]]);  // likelihood
  }
}
```

---

```{r}
# make data for stan
data_list <- list(
  # upper bounds of indices
  N = nrow(answers),                 # - n in {1,...,N} indexes answers
  J = max(answers$student_id_num),   # - j in {1,...,J} indexes students
  K = max(answers$question_id_num),  # - k in {1,...,K} indexes questions
  # indices for each respondent
  jj = answers$student_id_num,   # jj is a vector of student indices for each answer
  kk = answers$question_id_num,  # kk is a vector of question indices for each answer
  # observed answers
  y = answers$correct_answer
)
```

`jj` and `kk` are subtle!

- Because all students answer all questions, we have $N = 100 \times 55 = 5,500$ answers.
- `jj` and `kk` are both length-$N$ vectors.
- `jj[n]` tells us the student who gave the $n$th answer.
- `kk[n]` tells us the question that goes with the $n$th answer.

---

```{r}
n <- 1:5500
look_at_n <- c(1, 10, 55, 100, 1000, 2000, 5500) # answers to look at
cbind("n" = n, "jj" = data_list$jj, "kk" = data_list$kk, "y" = data_list$y)[look_at_n, ] 
```

---

I like to create a character string that has the Stan model, so that it's right in my R script. (Or you can create a file.)

```{r}
stan_code <- "
data {
  int<lower=1> N;  // number of answers
  int<lower=1> J;  // number of students
  int<lower=1> K;  // number of questions
  int<lower=1, upper=J> jj[N];  // student index for answer n
  int<lower=1, upper=K> kk[N];  // question index for answer n
  int<lower=0, upper=1> y[N];  // answer n is correct (0/1)
}

parameters {
  real mu;  // mean difficulty
  vector[J] alpha;  // abilities; one per student
  vector[K] beta;  // difficulties; one per question
  real<lower=0> sigma_beta;  // SD of difficulties
  real<lower=0> sigma_alpha;  // SD of abilities
}

model {
  sigma_beta ~ cauchy(0, 1);  // prior for SD of difficulties
  sigma_alpha ~ cauchy(0, 1);  // prior for SD of abilities
  mu ~ cauchy(0, 1);  // prior for mean difficulty
  alpha ~ normal(0, sigma_alpha);  // abilities
  beta ~ normal(0, sigma_beta);  // difficulties
  for (n in 1:N) {
    y[n] ~ bernoulli_logit(mu + alpha[jj[n]] - beta[kk[n]]);  // likelihood
  }
}
"
```

---

```{r}
#| message: false
#| warning: false
#| results: hide
#| cache: true

start_time <- Sys.time()

fit <- stan(model_code = stan_code, 
            data = data_list,
            iter = 1000, 
            warmup = 500,
            chains = 3, 
            cores = 3)
            
end_time <- Sys.time()
```


::: {.aside}
This model took `r format(difftime(end_time, start_time, units = "auto"), digits = 1)` to fit.
:::
---

```{r}
print(fit)
```

---

```{r}
fit %>%
  spread_draws(alpha[student_id_num])
```

---

```{r}
fit %>%
  spread_draws(alpha[student_id_num]) %>%
  median_qi()  # works like group_by() |> summarize()
```

---

```{r}
#| code-fold: true
fit %>%
  spread_draws(alpha[student_id_num], ndraws = 100) %>%
  median_qi() %>%
  left_join(grades) %>%
  slice_sample(n = 25) %>%
  ggplot(aes(x = alpha, xmin = .lower, xmax = .upper, 
             y = reorder(student_id_num, alpha), 
             color = letter_grade)) +
  #geom_label(aes(label = scales::percent(proportion_correct, accuracy = 1)), vjust = -0.3) + 
  geom_pointinterval(linewidth = 0.5, size = 0.5) + 
  theme_minimal() + 
  labs(y = "random sample of 25 students", 
       color = "letter grade")
```

--- 

```{r}
spread_draws(fit, beta[question_id_num]) 
```

--- 

```{r}
spread_draws(fit, beta[question_id_num]) |>
  median_qi()
```

--- 

```{r}
spread_draws(fit, beta[question_id_num]) |>
  median_qi() |>
  arrange(desc(beta)) 
```

---

```{r}
#| code-fold: true
#| fig-asp: .35
#| fig-width: 12
fit %>%
  spread_draws(beta[question_id_num]) %>%
  median_qi() %>%
  ggplot(aes(x = beta, xmin = .lower, xmax = .upper, 
             y = reorder(question_id_num, beta))) +
  geom_pointinterval(linewidth = 0.5, size = 0.5) + 
  theme_minimal() + 
  coord_flip() + 
  labs(x = "question")
```

---

## Fitting with `brm()`

```{r}
#| results: hide
#| cache: true
f <- correct_answer ~ (1 | question_id_num) + (1 | student_id_num)
brm_fit <- brm(f, data = answers, family = bernoulli, chains = 4, cores = 4)
```

The list of output from `brm()` includes a fit that is just the output from `stan()` above. We can use this element if we want, but the parameters have less nice names.

```{r}
print(brm_fit$fit)
```

--- 

```{r}
# brm()'s "r_question_id_num" is our "beta"
spread_draws(brm_fit, r_question_id_num[question_id_num, ]) |>
  median_qi() |> 
  arrange(desc(-r_question_id_num)) 
```

Notice anything different?

---

### And \{marginaleffects\}!

```{r}
#| fig-asp: .35
#| fig-width: 12
p <- predictions(brm_fit)
ggplot(p, aes(y = estimate, x = reorder(question_id_num, estimate))) + 
  geom_point(alpha = 0.2) + 
  geom_line(aes(group = student_id_num), alpha = 0.2) + 
  labs(x = "question")
```

---

## Fitting with `lmer()`

```{r}
# fit model
f <- correct_answer ~ (1 | question_id_num) + (1 | student_id_num)
lmer_fit <- glmer(f, data = answers, family = binomial)

# grab "difficulties"
ranef(lmer_fit)$question_id_num |>
  as_tibble(rownames = "question_id_num") |> 
  rename(beta = "(Intercept)") |>
  arrange(beta)
```

---

### And \{marginaleffects\}!

::: {.aside}
- We talked about how `predictions()` and `comparisons()` **unify** diverse statistical models.
- But it also unifies diverse estimation routines. 
- Here it unifies `brm()` and `lmer()` output. The code below only changes the fit I give `predictions()`.
:::

```{r}
#| fig-asp: .35
#| fig-width: 12
p <- predictions(lmer_fit)
ggplot(p, aes(y = estimate, x = reorder(question_id_num, estimate))) + 
  geom_point(alpha = 0.2) + 
  geom_line(aes(group = student_id_num), alpha = 0.2) + 
  labs(x = "question")
```

---

### What if no partial pooling?

```{r}
#| fig-asp: .35
#| fig-width: 12
# fit model
f <- correct_answer ~ factor(question_id_num) + factor(student_id_num)
glm_fit <- glm(f, data = answers, family = binomial)

# marginaleffects
p1 <- predictions(glm_fit)
ggplot(p1, aes(y = estimate, x = reorder(question_id_num, estimate))) + 
  geom_point(alpha = 0.2) + 
  geom_line(aes(group = student_id_num), alpha = 0.2) + 
  labs(x = "question")
```

---

```{r}
#| code-fold: true
#| fig-asp: .75
#| fig-width: 12
comb <- bind_rows(
  mutate(p, model = "lmer"), 
  mutate(p1, model = "glm")) |>
mutate(question_id_num = reorder(factor(question_id_num), estimate))

ggplot(comb, aes(x = estimate, y = model)) + 
  geom_point(alpha = 0.2) + 
  geom_line(aes(group = rowid), alpha = 0.2) + 
  facet_wrap(vars(question_id_num))
```

## What if the exam had Students 1 through 10 had Questions 7, 28, 13, 49, and 43?

```{r}
#| code-fold: true
#| fig-asp: .75
#| fig-width: 8
sub <- filter(answers, question_id_num %in% c(27, 28, 13, 49, 43), student_id_num <= 10)

# lmer
f <- correct_answer ~ (1 | question_id_num) + (1 | student_id_num)
lmer_fit <- glmer(f, data = sub, family = binomial)
p <- predictions(lmer_fit)

# glm
f <- correct_answer ~ factor(question_id_num) + factor(student_id_num)
glm_fit <- glm(f, data = sub, family = binomial)
p1 <- predictions(glm_fit)

comb <- bind_rows(
  mutate(p, model = "lmer"), 
  mutate(p1, model = "glm")) |>
mutate(question_id_num = reorder(factor(question_id_num), estimate))

ggplot(comb, aes(x = estimate, y = model)) + 
  geom_point(alpha = 0.2) + 
  geom_line(aes(group = rowid), alpha = 0.2) + 
  facet_wrap(vars(question_id_num))
```



