---
title: "MRP"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
execute:
  echo: true
fig-align: center
fig-width: 8
fig-asp: 0.5
---


```{r echo=FALSE}
#| message: false
#| warning: false
options(width = 90)
options(digits = 3)

library(tidyverse)
theme_set(theme_minimal())
```

Here are the packages we're using:

```{r}
#| message: false
#| warning: false

# generally useful packages
library(tidyverse)
library(marginaleffects)
library(modelsummary)
library(ggrepel)

# for fitting hierarchical models
library(lme4)
library(brms)  # see also rstanarm::stan_lmer()
library(rstan)

# for post-processing simulations
library(tidybayes)
```

## MRP

-   "Multilevel regression with poststratification" or "Mister P"
-   We have sufficiently large polls to estimate national opinion well, but how can we use these to estimate state-level opinion well across the fifty states?

## Two-step process

1.  Use a **national survey data** (e.g., N = 2,000) and **a hierarchical model** to estimate public opinion within small subgroups defined by state and demographic variables. *Example: Men age 18-29 without military experience and no HS degree living in Alabama.*
2.  Look up the fraction of each state that falls into each demographic category. Weight the predictions for each state-demographic groups by these fractions to obtain an estimate of state-level opion.

## Two-step process: Step 2(a)

Find the populations of demographic subgroups in each state.

This might come from the Census or CPS.

Here's an example [packaged with](https://marginaleffects.com/chapters/mrp.html) {marginaleffects}.

```{r}
demographics <- get_dataset("ces_demographics") |>
  glimpse()
```

## Two-step process: Step 1

Use a **national survey data** (e.g., N = 2,000) and a hierarchical model to estimate public opinion within each each subgroup.

```{r}
survey <- get_dataset("ces_survey") |>
  glimpse()
```

## Two-step process: Step 2(b)

Fit the model.

```{r}
#| label: mrp-model
#| message: false
#| warning: false
#| cache: true
#| results: hide
# fit model
f <- defund ~ age + education + military + gender + (1 + gender | state)
model <- brm(f, family = bernoulli, data = survey, chains = 4, cores = 4)
```

## Two-step process: Step 2(c)

Create the predictions.

```{r}
p <- predictions(model, newdata = demographics) |>
  glimpse()
```

## Two-step process: Step 2(d)

Weight the predictions.

```{r}
ests <- p |>
  mutate(w_ests = estimate * percent) |>
  group_by(state) |>
  summarize(mrp_est = sum(w_ests)) |>
  ungroup() |>
  glimpse()
```

## Plot

```{r}
gg_ests <- ests |>
  mutate(state = reorder(state, mrp_est))
ggplot(gg_ests, aes(x = mrp_est, y = state)) + 
  geom_point()
```

## Question

We used this model. What is it? What others might we have used?

```         
defund ~ age + education + military + gender + (1 + gender | state)
```
