---
title: "Unordered Logit"
format:
  revealjs:
    css: styles.css
    transition: fade
    slide-number: true
    width: 1600
    height: 900
    code-block-height:  900px
    mathjax: true  
bibliography: references.bib
citations-hover: true
reference-location: document
warning: false
message: false
---

```{r}
#| include: false
library(tidyverse)
```

## The Inverse-Logit Function

Suppose $y_i \in \{0, 1\}$. Using logistic regression, we model $\Pr(y_i = 1)$ as a function of covariates, so that

$$
\Pr(Y_i = 1) = \frac{\exp(X_i \beta)}{1 + \exp(X_i \beta)}.
$$

The function $f(z) = \frac{e^z}{1 + e^z}$ is called the inverse-logit function and maps any real number $X_i \beta$ into a probability between 0 and 1.

Because the two probabilities must sum to one:

$$
\Pr(Y_i = 0) = 1 - \Pr(Y_i = 1)
             = \frac{1}{1 + \exp(X_i \beta)}.
$$

## The Softmax Function

The softmax generalizes the inverse-logit--it takes $J$ real input (i.e., from $\mathbb{R}^J$ and scales them to sum to one.

$$
\text{softmax}(z_j) = \frac{\exp(z_j)}{\sum_{k=1}^{J} \exp(z_k)}, \qquad j = 1,\ldots,J. 
$$
Because the $J$ outputs must sum to one, it is natural to interpret the outputs as *probabilities*.

---

The OJS widget below allows you to experiment with $J = 4$ inputs and see how the outputted probabilities change.

```{ojs}
//| echo: false

// Sliders (left column)
viewof z1 = Inputs.range([-5, 5], {step: 0.1, value: 1,   label: "z1"});
viewof z2 = Inputs.range([-5, 5], {step: 0.1, value: 0.5, label: "z2"});
viewof z3 = Inputs.range([-5, 5], {step: 0.1, value: -0.5,label: "z3"});
viewof z4 = Inputs.range([-5, 5], {step: 0.1, value: -1,  label: "z4"});

// Softmax pieces
labels = ["Pr(A)","Pr(B)","Pr(C)","Pr(D)"];
zs = [z1, z2, z3, z4];
exps = zs.map(Math.exp);
sumexp = exps.reduce((a, b) => a + b, 0);
ps = exps.map(e => e / sumexp);

// Data for stacked bar (one row, four segments)
data = ps.map((p, i) => ({ label: labels[i], value: p, row: "softmax" }));

// === Two-column layout: sliders (left) + live table (right) ===
{
const container = html`<div style="
  display:flex;
  gap:24px;
  align-items:flex-start;
  padding-top:8px;
"></div>`;

// Left column: sliders
const left = html`<div style="display:grid; gap:10px; min-width:220px;"></div>`;
left.append(viewof z1, viewof z2, viewof z3, viewof z4);

// Right column: table
const rows = labels.map((cls, i) => ({
Class: cls,
z: zs[i],
"exp(z)": exps[i],
Probability: ps[i]
}));

const right = html`<div style="
  min-width:240px;
  max-width:660px;
  padding:4px 0;
  font-size:2.00rem;
  line-height:1.1;
"></div>`;

right.append(
Inputs.table(rows, {
columns: ["Class", "z", "exp(z)", "Probability"],
format: {
z: d => d.toFixed(2),
"exp(z)": d => d.toFixed(3),
Probability: d => (d * 100).toFixed(1) + "%"
}
})
);

container.append(left, right);
return container;
}

// Stacked bar chart — bigger everything
Plot.plot({
width: 900,              // wider
height: 120,             // taller
marginLeft: 20,
marginRight: 20,
marginBottom: 50,
x: {
domain: [0, 1],
tickFormat: d => (d * 100).toFixed(0) + "%",
label: "Probability",
labelArrow: false,
labelAnchor: "center",
tickSize: 10,
tickPadding: 8
},
y: { axis: null },

// Bigger fonts everywhere
style: {
fontSize: "18px",        // main text size
fontFamily: "sans-serif"
},

color: {
legend: true,
label: "Class",
tickSize: 12
},

marks: [
Plot.barX(
data,
Plot.stackX({ x: "value", y: "row", fill: "label", sort: { fill: null } })
),
Plot.ruleX([0, 1])
]
})


```


## Adding covariates 

Now suppose the outcome takes $J$ possible values:

$$
y_i \in \{1, 2, \ldots, J\}.
$$

**Example 1**: We might label *vote choice* in the US as (1) *abstain*, (2) *Republican*, (3) *Democrat*, or (4) *other*.

**Example 2**: We might label *coup attempts* in a given country-year as (0) *none attempted*, (1) *failed attempt*, or (2) *successful attempt*.

---

For each category $j$, define a linear predictor $\eta_{ij} = X_i \beta_j$. Here, each $\beta_j$ is a *vector* of coefficients.

We can use the *softmax* function to convert these linear predictors into probabilities that sum to one. 

$$
\Pr(Y_i = j) = \frac{\exp(\eta_{ij})}{\sum_{k=1}^{J} \exp(\eta_{ik})},
\qquad j = 1, \ldots, J.
$$

This is a generalization of the inverse-logit to $J$ categories. If $J = 2$, the softmax reduces to the inverse-logit.

---

However, this model is not identified because adding any constant to each of the $\eta_{ij}$ produces the same probabilities.

::: columns
::: column

**Example Inputs**

| Input $z_j$ | Softmax $p_j$ |
|------------:|--------------:|
| $-1$        | 0.0303 |
| $1$         | 0.2242 |
| $2$         | 0.6095 |
| $0.5$       | 0.1360 |

:::
::: column

**Adding a Constant ($+2$)**

| New Input $z_j + 2$        | Softmax $p_j$ |
|---------------------------:|--------------:|
| $-1 + 2 = 1$               | 0.0303 |
| $1 + 2 = 3$                | 0.2242 |
| $2 + 2 = 4$                | 0.6095 |
| $0.5 + 2 = 2.5$            | 0.1360 |

:::
:::

Adding a constant to all inputs leaves the softmax unchanged. This is why the multinomial logit model requires an identification constraint.

---

To identify the model, we set $\eta_{iJ} = 1$. Then for $j = 1, \ldots, J-1$, we have

$$
\Pr(Y_i = j) =
\frac{\exp(X_i \beta_j)}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$

And for the "baseline" category $J$, we have

$$
\Pr(Y_i = J) =
\frac{1}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$
---

```{r}
# load data
anes <- read_csv("https://pos5747.github.io/data/red-state-pid.csv") |>
  mutate(pid = factor(pid, ordered = TRUE)) |>
  mutate(pid3 = case_when(
    pid %in% c("1. Strong Democrat",
               "2. Not very strong Democrat") ~ "D",
    pid %in% c("6. Not very strong Republican",
               "7. Strong Republican") ~ "R",
    pid %in% c("3. Independent (leans Democrat)",
               "4. Independent",
               "5. Independent (leans Republican)") ~ "I"
  )) |>
  glimpse()
```
---

```{r}
library(nnet)
fit <- multinom(pid3 ~ income, data = anes)

# summarize fit
summary(fit)
```

---

```{r}
# use {marginaleffects}
library(marginaleffects)
p <- predictions(fit, 
            newdata = datagrid(income = unique)) |>
  glimpse()
```

---

```{r}
ggplot(p, aes(x = income, y = estimate, color = group)) + 
  geom_point() + 
  geom_line()
```

---

```{r}
#| message: false
#| warning: false
#| cache: true
#| results: hide
# load packages
library(brms)

# fit same model, but coefs vary across years
fit <- brm(pid3 ~ income*rs_state_median_income + (1 + income | state_name), 
           data = anes, 
           family = categorical(),
           cores = 4, 
           chains = 4, 
           warmup = 2000, 
           iter = 4000,
           backend = "rstan")
```

---

```{r}
#| fig-asp: 0.65
#| fig-width: 8
#| code-fold: true

# make grid 
grid <- anes %>%
  dplyr::select(state_name, state_median_income, rs_state_median_income) %>%
  distinct() %>%
  mutate(income = NA)

# make predictions
p <- predictions(fit, 
                 variables = list(income = unique),
                 newdata = grid) |>
  mutate(state_name = reorder(state_name, rs_state_median_income))

# make plot
gg <- ggplot(p, aes(x = income, y = estimate, color = group)) +
  geom_line() +
  facet_wrap(vars(state_name))
print(gg)
```

---

```{r}
#| echo: false
#| fig-asp: 0.65
#| fig-width: 8

print(gg)
```

---

```{r}
#| echo: false
#| fig-asp: 0.5
#| fig-width: 8

# make plot
p <- filter(p, state_name %in% c("Mississippi", "District of Columbia"))
gg <- ggplot(p, aes(x = income, y = estimate, color = group)) +
  geom_line() +
  facet_wrap(vars(state_name))
print(gg)
```

## An observation worth remembering

```{r}
glimpse(anes)

f <- pid ~ income
fit1 <- MASS::polr(f, data = anes)
fit2 <- multinom(f, data = anes)

BIC(fit1, fit2)
```
