# Metropolis

## MCMC

The Metropolis algorithm belongs to a larger class of Markov chain Monte Carlo (MCMC) algorithms, which includes Metropolis-Hastings, Gibbs sampling, and the Hamiltonian Monte Carlo algorithm used by Stan.

::: {.defbox}
MCMC refers to a family of algorithms for generating samples from a target probability distribution by constructing a Markov chain whose stationary distribution is that target. Over many iterations, the samples produced by the chain approximate (dependent) draws from the posterior distribution, allowing us to estimate quantities such as means, variances, and credible intervals even when the posterior cannot be computed analytically.
:::

## Metropolis

::: {.algorithm}
**Algorithm: Metropolis**

*Inputs:*

- The unnormalized log-posterior $\log f(\theta\mid y)$.
- Desired number of iterations $S$. *Note: the samples are dependent so choose large $S$ (e.g., $S = 20{,}000$), perhaps very large (e.g., $S = 200{,}000$).*
- Tuning parameter $\tau$, which controls the width of the uniform proposal density.

*Algorithm:*

- **Initialize:** Choose an initial $\theta^{(1)}$. Set $s=1$.
- **Repeat until $s=S$:**
    - **Propose:** Draw $z$ from a uniform distribution from $[\theta^{(s)} - \tau]$ to $[\theta^{(s)} + \tau]$.
    - **Compute log-acceptance:** $\Delta = \log f(z)-\log f \big(\theta^{(s)}\big)$
    - **Accept/reject:** Draw $u \sim \text{Uniform}(0,1)$. Accept iff $\log u \le \Delta$. If accept, set $\theta^{(s+1)}=z$; otherwise $\theta^{(s+1)}=\theta^{(s)}$.
    - **Iterate:** $s\leftarrow s+1$.

*Output:* A Markov chain $\{\theta^{(s)}\}_{s=1}^S$ with stationary distribution $f(\theta\mid y)$. Consecutive draws are not independent; in practice, discard a burn-in and tune $\tau$ to maintain a reasonable acceptance rate (perhaps 0.2-0.5, problem-dependent).
:::

## Accept/reject

::: {.defbox}
Importantly, this accepts with probability equal to $\dfrac{f(\text{proposal})}{f(\text{current})}$ if the ratio is between zero and one and always if the ratio is greater than one.
:::

::: {.defbox}
This step is similar to the rejection algorithm, except we keep the current value in place of the rejected proposal in Metropolis.
:::

::: {.notes}
Both sentences are the footnote to the accept/reject step in the notes.
[CTK: contrast with last week: rejection throws the proposal away and tries again; Metropolis writes down the current value again.]
:::

<!-- walker-clicks: begin (generated by wk07-lecture-material/walker-clicks/make-walker-segments.R; do not edit by hand) -->
## The algorithm, one iteration at a time

::: {.r-stack}
```{=html}
<video data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-01.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-02.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-03.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-04.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-05.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-06.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-07.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-08.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-09.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-10.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-11.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-12.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-13.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-14.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-15.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-16.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-17.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-18.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-19.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-20.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-21.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-22.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-23.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-24.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-25.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-26.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-27.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-28.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-29.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-30.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-31.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-32.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-33.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-34.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-35.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-36.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-37.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-38.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-39.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-40.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-41.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-42.mp4"></video>
<video class="fragment" data-autoplay muted playsinline preload="auto" width="1280" src="img/walker/seg-43.mp4"></video>
<script>
(function () {
  function media(el) {
    var list = el.matches && el.matches('video[data-autoplay]') ? [el] : [];
    return list.concat(Array.prototype.slice.call(el.querySelectorAll('video[data-autoplay]')));
  }
  function hook() {
    Reveal.on('fragmentshown', function (ev) {
      (ev.fragments || [ev.fragment]).forEach(function (f) {
        media(f).forEach(function (v) {
          // a click during a segment leaves that segment playing to its end
          // underneath the new one (seeking a hidden video to its end does not
          // take in Chrome), so stepping back shows its last frame
          v.currentTime = 0; var p = v.play(); if (p && p.catch) p.catch(function () {});
        });
      });
    });
    Reveal.on('fragmenthidden', function (ev) {
      (ev.fragments || [ev.fragment]).forEach(function (f) {
        media(f).forEach(function (v) { v.pause(); v.currentTime = 0; });
      });
    });
  }
  function init() { if (!window.Reveal) return; if (Reveal.isReady()) hook(); else Reveal.on('ready', hook); }
  if (document.readyState === 'complete') init(); else window.addEventListener('load', init);
})();
</script>
```
:::

::: {.notes}
The walker animation (`wk07-slides-material/figs/fig-metropolis-walker.gif`) cut into 43 segments that advance on click: the first plays on arrival, each click plays the next and holds its last frame. 42 clicks in all. Segment ends (iteration-step) and lengths:

1. 1-intro (1.0 s)
2. 2-propose (0.8 s)
3. 2-compute (0.8 s)
4. 2-draw_u (0.5 s)
5. 2-decide (0.8 s)
6. 2-record (0.6 s)
7. 3-propose (0.8 s)
8. 3-compute (0.8 s)
9. 3-draw_u (0.5 s)
10. 3-decide (0.8 s)
11. 3-record (0.6 s)
12. 4-propose (0.5 s)
13. 4-compute (0.5 s)
14. 4-draw_u (0.5 s)
15. 4-decide (0.5 s)
16. 4-record (0.5 s)
17. 5-propose (0.5 s)
18. 5-compute (0.5 s)
19. 5-draw_u (0.5 s)
20. 5-decide (0.5 s)
21. 5-record (0.5 s)
22. 18-zoom (13.8 s)
23. 19-propose (0.8 s)
24. 19-compute (0.8 s)
25. 19-draw_u (0.8 s)
26. 19-decide (0.8 s)
27. 19-record (0.6 s)
28. 20-propose (0.8 s)
29. 20-compute (0.8 s)
30. 20-draw_u (0.8 s)
31. 20-decide (0.8 s)
32. 20-record (0.6 s)
33. 21-propose (0.8 s)
34. 21-compute (0.8 s)
35. 21-draw_u (0.8 s)
36. 21-decide (0.8 s)
37. 21-record (0.6 s)
38. 22-propose (0.6 s)
39. 22-compute (0.6 s)
40. 22-draw_u (0.8 s)
41. 22-decide (0.8 s)
42. 22-record (0.6 s)
43. 36-hold (16.0 s)

A clicker's next button sends Right or Page Down, which reveal maps to the next fragment, so nothing needs configuring. Clicking while a segment is still playing skips to the next one; Left steps back to the previous segment's last frame.

[CTK: the pause points are the `stops` table at the top of make-walker-segments.R; edit it and re-run the script to reshape the clicks.]
:::
<!-- walker-clicks: end -->

## `metrop()` {.smaller}

::: {.columns}
::: {.column width="50%"}
```r
metrop <- function(logf, theta_start, S = 10000, tau = 0.1, ...) {
  # initialize matrix of samples with starting values
  k <- length(theta_start)
  samples <- matrix(NA_real_, nrow = S, ncol = k)
  samples[1, ] <- theta_start

  # proceed with algorithm
  for (s in 2:S) {
    # extract current location
    current <- samples[s - 1, ]

    # generate symmetric random-walk proposal
    proposed_move <- runif(k, -tau, tau)
    proposal <- current + proposed_move
```
:::

::: {.column width="50%"}
```r
    # acceptance step
    delta <- logf(proposal, ...) - logf(current, ...)
    if (delta > 0) {
      accept <- TRUE
    } else {
      accept <- (log(runif(1)) <= delta)
    }

    # update samples
    if (accept) {
      samples[s, ] <- proposal
    } else {
      samples[s, ] <- current
    }
  }

  # return a list, so the samples are m$samples
  list(samples = samples)
}
```
:::
:::

```{r}
#| label: s2-metrop-features
#| echo: false
# the "some features added" version from the notes and the in-class scripts:
# same algorithm and same random draws as the minimal version above, but it
# returns list(prop_accepted, samples) and reports a message (suppressed here)
metrop <- function(logf, theta_start, S = 10000, tau = 0.1, progress_bar = FALSE, ...) {
  start_time <- Sys.time()
  k <- length(theta_start)
  samples <- matrix(NA_real_, nrow = S, ncol = k)
  samples[1, ] <- theta_start
  n_accepted <- 0
  if (progress_bar) {
    pb <- txtProgressBar(min = 0, max = S, style = 3)
  }
  for (s in 2:S) {
    current <- samples[s - 1, ]
    proposed_move <- runif(k, -tau, tau)
    proposal <- current + proposed_move
    delta  <- logf(proposal, ...) - logf(current, ...)
    if (delta > 0) {
      accept <- TRUE
    } else {
      accept <- (log(runif(1)) <= delta)
    }
    if (accept) {
      samples[s, ] <- proposal
      n_accepted <- n_accepted + 1
    } else {
      samples[s, ] <- current
    }
    if (progress_bar && (s %% (S / 100) == 0 || s == S)) {
      setTxtProgressBar(pb, s)
    }
  }
  if (progress_bar) close(pb)
  message(
    paste0(
      "Successfully generated ", scales::comma(S), " dependent samples!\n",
      "Accepted moves: ", scales::comma(n_accepted), "\n",
      "Acceptance rate: ", scales::percent(n_accepted / (S - 1), accuracy = 1)
    )
  )
  list(
    prop_accepted = n_accepted / (S - 1),
    samples = samples
  )
}
```

::: {.notes}
This is the minimal version from the notes, except that it returns `list(samples = samples)` rather than the bare matrix, so `m$samples` works on every later slide. The in-class scripts (`wk07-toothpaste-metropolis.R`, `wk07-logit-metropolis.R`) use the "some features added" version: same algorithm, plus a progress bar, the acceptance rate, and a reminder message. It returns a list, so the code from here on uses `m$samples`. The deck runs that version too (hidden chunk).
:::

## Toothpaste cap

For $k=8$ and $N=150$ with a $\text{beta}(3, 15)$ prior, the posterior is $\text{beta}(11,157)$.

::: {.columns}
::: {.column width="55%"}
```{r}
#| label: s2-toothpaste-run
# create log-target for Beta posterior; -Inf outside (0,1)
logf <- function(p) {
  ifelse(p <= 0 | p >= 1,
         -Inf,
         dbeta(p, 11, 157, log = TRUE))
}

# run algorithm starting in the middle of (0,1)
set.seed(1234)
m <- metrop(
  logf,            # log-target
  theta_start = 0.5
)
```
:::

::: {.column width="45%"}
```{r}
#| label: s2-toothpaste-means
# closed-form posterior mean
11 / (11 + 157)

# simulations (samples is S x k; here k = 1)
mean(m$samples)
```
:::
:::

## Toothpaste cap: the samples

```{r}
#| label: s2-toothpaste-panel
#| echo: false
#| fig-width: 12
#| fig-asp: 0.5
par(mfrow = c(2, 2), family = "Source Sans 3", cex = 1.3, las = 1, mar = c(4, 5, 2, 1))
hist(m$samples, main = "All 10,000 samples", xlab = "p")
plot(m$samples, type = "l", main = "All 10,000 samples", xlab = "Iteration", ylab = "p")
plot(m$samples[1:100], type = "l", main = "First 100 samples", xlab = "Iteration", ylab = "p")
plot(m$samples[1001:1100], type = "l", main = "Samples 1,001 through 1,100", xlab = "Iteration", ylab = "p")
```

## Two things worth noting

1. The samples are **not independent**. Many consecutive draws are identical (rejections) or move only a little; the chain **explores the target slowly**.
2. It takes a few iterations for the chain to reach the high-probability region (burn-in). Starting at $\pi=0.5$ is far from the posterior mass; it takes roughly a few dozen iterations to reach $\approx 0.1$.

## Burn-in

The MCMC samples are heavily dependent on the starting values, so discard early samples that depend strongly on the starting value (the "burn-in"). Discarding the first 10% to 50% of the samples is typical.

```{r}
#| label: s2-burnin
# simulations, keep only samples 5,001 through 10,000 (discard first half)
mean(m$samples[5001:10000, 1])
```

::: {.notes}
Aside from the notes: Stan and {brms} call this period the "warmup" rather than the burn-in. During warmup, Stan's sampler also tunes itself (e.g., its step size), so the warmup samples are not draws from a fixed algorithm. Our Metropolis sampler keeps tau fixed, so "burn-in" fits here.
:::

## R-hat

But how do we know how long to run our algorithms? How do we know if the samples have converged to the target distribution? This is still a difficult problem.

::: {.defbox}
No diagnostic can show that the chains have converged, but R-hat can show that they have *not*. We check that R-hat is less than 1.01 (Vehtari et al. 2021).
:::

To compute R-hat, run several chains from overdispersed starting values. It is important that these be high and low relative to the target distribution. If chains have converged, within- and between-chain variation align and $\hat{R}\approx 1$.

::: {.notes}
Notes continue: If R-hat is larger than 1.01, the chains have not converged and we need more iterations. If R-hat is less than 1.01, we have found no sign of a problem.
Aside from the notes: {brms} warns only when some R-hat is larger than 1.05. We use the stricter threshold of 1.01 from Vehtari et al. (2021), so check R-hat yourself even when {brms} prints no warning.
:::

## R-hat: 25 iterations

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: s2-rhat-short
# run algorithm four times, making four "chains"
set.seed(1234)
m1 <- metrop(logf, theta_start = 0.01, S = 25)
m2 <- metrop(logf, theta_start = 0.25, S = 25)
m3 <- metrop(logf, theta_start = 0.75, S = 25)
m4 <- metrop(logf, theta_start = 0.99, S = 25)

# a matrix of chains for first (and only) dimension
matrix_of_chains <- cbind(
  m1$samples[, 1],
  m2$samples[, 1],
  m3$samples[, 1],
  m4$samples[, 1]
)

# compute r-hat using the rhat() function in {posterior}
posterior::rhat(matrix_of_chains)
```
:::

::: {.column width="50%"}
```{r}
#| label: s2-rhat-short-plot
#| echo: false
#| fig-width: 7
#| fig-asp: 0.85
colnames(matrix_of_chains) <- paste("Chain", 1:ncol(matrix_of_chains))
gg_df <- matrix_of_chains |>
  as_tibble() |>
  mutate(Iteration = 1:n()) |>
  pivot_longer(cols = `Chain 1`:`Chain 4`, names_to = "Chain", values_to = "Sample")

ggplot(gg_df, aes(x = Iteration, y = Sample, color = Chain)) +
  geom_line(linewidth = 1) +
  scale_color_lecture() +
  labs(color = NULL)
```
:::
:::

## R-hat: 25,000 iterations, first 10,000 discarded

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: s2-rhat-long
# run algorithm four times, making four "chains"
set.seed(1234)
m1 <- metrop(logf, theta_start = 0.01, S = 25000)
m2 <- metrop(logf, theta_start = 0.25, S = 25000)
m3 <- metrop(logf, theta_start = 0.75, S = 25000)
m4 <- metrop(logf, theta_start = 0.99, S = 25000)

# a matrix of chains for first (and only) dimension
matrix_of_chains <- cbind(
  m1$samples[10001:25000, 1],
  m2$samples[10001:25000, 1],
  m3$samples[10001:25000, 1],
  m4$samples[10001:25000, 1]
)

# compute r-hat using the rhat() function in {posterior}
posterior::rhat(matrix_of_chains)
```
:::

::: {.column width="50%"}
```{r}
#| label: s2-rhat-long-plot
#| echo: false
#| fig-width: 7
#| fig-asp: 0.85
colnames(matrix_of_chains) <- paste("Chain", 1:ncol(matrix_of_chains))
gg_df <- matrix_of_chains |>
  as_tibble() |>
  mutate(Iteration = 10000 + 1:n()) |>
  pivot_longer(cols = `Chain 1`:`Chain 4`, names_to = "Chain", values_to = "Sample")

ggplot(gg_df, aes(x = Iteration, y = Sample, color = Chain)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  scale_color_lecture() +
  scale_x_continuous(labels = scales::comma) +
  guides(color = guide_legend(override.aes = list(linewidth = 1.5, alpha = 1))) +
  labs(color = NULL)
```
:::
:::

::: {.notes}
If we increase S to a larger value and discard the initial samples, then R-hat falls below 1.01.
:::

## Parallel chains {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
#| label: s2-parallel
# overdispersed starting values
starting_values <- seq(0.05, 0.95, length.out = 10)

# one core per chain, but no more than the machine has
n_cores <- min(10, parallel::detectCores(logical = FALSE))
cl <- makeCluster(n_cores)
clusterSetRNGStream(cl, iseed = 1234)  # reproducible draws on the workers
registerDoParallel(cl)

# run 10 chains in parallel; combine samples with cbind()
matrix_of_chains <- foreach(s = starting_values,
                            .combine = cbind) %dopar% {
  m <- metrop(logf, theta_start = s, S = 25)
  m$samples
}
stopCluster(cl)

# compute R-hat using {posterior}
posterior::rhat(matrix_of_chains)
```
:::

::: {.column width="42%"}
```{r}
#| label: s2-parallel-plot
#| echo: false
#| fig-width: 6.5
#| fig-asp: 1
colnames(matrix_of_chains) <- paste("Chain", 1:ncol(matrix_of_chains))
gg_df <- matrix_of_chains |>
  as_tibble() |>
  mutate(Iteration = 1:n()) |>
  pivot_longer(cols = `Chain 1`:`Chain 10`, names_to = "Chain", values_to = "Sample") |>
  mutate(Chain = fct_inorder(Chain))

# exception to the deck palette: 10 chains need 10 colors, so viridis here
ggplot(gg_df, aes(x = Iteration, y = Sample, color = Chain)) +
  geom_line(linewidth = 0.9) +
  scale_color_viridis_d(end = 0.95) +
  labs(color = NULL) +
  theme(legend.position = "none")
```
:::
:::

::: {.notes}
Notes: Multiple chains are naturally parallelizable. In fact, these are called "embarrassingly parallel," because the chains are run entirely independently. Nothing from the first chain is needed for the second, and so on.
The figure shows the 25 samples from each of the 10 chains. Notice the strong dependence on starting values! In practice, we'd need to generate many more samples, discard the first samples, and check that R-hat is less than 1.01.
The in-class script uses S = 250 here; the notes and this slide use S = 25.
:::

## ESS

Dependent Metropolis samples carry less information than the same number of independent samples.

::: {.defbox}
Effective Sample Size (ESS): "our dependent samples are like ______ independent samples."
:::

```{r}
#| label: s2-ess
# the four long chains from above, minus the first 10,000 samples of each
matrix_of_chains <- cbind(
  m1$samples[10001:25000, 1],
  m2$samples[10001:25000, 1],
  m3$samples[10001:25000, 1],
  m4$samples[10001:25000, 1]
)
dim(matrix_of_chains)  # 15,000 x 4 = 60,000 samples

# are like
posterior::ess_bulk(matrix_of_chains)  # ESS for the center of the distribution
posterior::ess_tail(matrix_of_chains)  # ESS for tail estimands
```

::: {.notes}
Notes: The Effective Sample Size (ESS) summarizes the amount of information in our dependent samples. The literature and practices on ESS are evolving rapidly; see `?posterior::ess_bulk` for the latest references.
:::

## Tuning $\tau$

```{r}
#| label: s2-tuning
# varying the tuning parameter tau
set.seed(1234)
m1 <- metrop(logf, theta_start = 0.99, S = 1000, tau = 0.1)
m2 <- metrop(logf, theta_start = 0.99, S = 1000, tau = 0.01)
m3 <- metrop(logf, theta_start = 0.99, S = 1000, tau = 0.001)
m4 <- metrop(logf, theta_start = 0.99, S = 1000, tau = 3)
```

```{r}
#| label: s2-tuning-panel
#| echo: false
#| fig-width: 12
#| fig-asp: 0.32
par(mfrow = c(2, 2), family = "Source Sans 3", cex = 1.25, cex.main = 1, las = 1, mar = c(2.2, 4, 2, 1))
plot(m1$samples, type = "l", main = "tau = 0.1: Nice!", ylim = c(0, 1), xlab = "", ylab = "p")
plot(m2$samples, type = "l", main = "tau = 0.01: Too small! (Accept too often.)", ylim = c(0, 1), xlab = "", ylab = "p")
plot(m3$samples, type = "l", main = "tau = 0.001: Much too small! (Accept much too often.)", ylim = c(0, 1), xlab = "", ylab = "p")
plot(m4$samples, type = "l", main = "tau = 3: Too wide! (Accept too rarely.)", ylim = c(0, 1), xlab = "", ylab = "p")
```

::: {.key}
the value of `tau` should be something like the SD of the posterior
:::

::: {.smaller}
"Something like" means within an order of magnitude. For the toothpaste cap problem, about 5 SDs (`tau = 0.1`) is about right.
:::

::: {.notes}
Notes: The tuning parameter matters a lot. I've chosen a smart default for many problems, but the value of `tau` should be something like the SD of the posterior.
Measured 2026-10-06 (grid of tau, 20,000 draws, ESS per draw): the posterior SD of beta(11, 157) is 0.019; efficiency peaks near tau = 4 SD (acceptance about 40%), and tau = 0.1 (5.3 SD) is within about 10% of the peak; tau = SD gives one fifth the effective sample; tau = 3 is wider than the whole support. Theory (Roberts, Gelman & Gilks 1997): optimal proposal SD is about 2.4 posterior SDs in 1-d, shrinking as 1/sqrt(d); a uniform on +/- tau has SD tau/sqrt(3), so the optimal half-width is about 4 SD in 1-d and about 1.8 SD for the 5-parameter logit (measured optimum 1.5-2 SD, acceptance about 20%).
:::

## Why the Metropolis algorithm works

**Symmetric proposals give equal opportunity.**

- From any point $x$, the chance of proposing $z$ is the same as proposing $x$ from $z$.
- The proposal distribution itself doesn't favor any direction.

**Asymmetric acceptance adds the right bias.**

- Moves to higher density are always accepted.
- Moves to lower density are accepted with probability $f(z)/f(x)$.
- This rule makes the chain linger in high-density regions.

## Why the Metropolis algorithm works

**Balanced flows keep the target intact.**

- Although proposals are symmetric, the acceptance rule ensures that the expected number of transitions $A \rightarrow B$ equals those from $B \rightarrow A$.
    - **From higher-density $A$ to lower-density $B$:** the chain is at $A$ often (every proposal toward $A$ is accepted, so it is easy to get there), but it leaves $A$ for $B$ only sometimes, with probability $f(B)/f(A)$.
    - **From lower-density $B$ to higher-density $A$:** the chain is at $B$ less often (proposals toward $B$ are often rejected), but whenever it is at $B$ and proposes $A$, it always moves.
    - Many visits times a fraction of exits equals few visits times certain exits: $f(A) \cdot \frac{f(B)}{f(A)} = f(B) \cdot 1$. The two flows match.
- This condition guarantees that the samples are stationary.

**Long-run behavior reflects probability mass.**

- Because the chain moves through the space according to these balanced transition rules, it spends time in each region in proportion to its probability under the target.
- In the long run, the sample frequencies mirror the target distribution.

## A weird posterior {.smaller}

::: {.columns}
::: {.column width="50%"}
```{r}
#| label: s2-sawtooth
# sawtooth log-prior (unnormalized):
# log of ((n_teeth*p) %% 1) on (0,1)
log_prior_saw <- function(p, n_teeth = 5) {
  ifelse(p <= 0 | p >= 1, -Inf, log((n_teeth * p) %% 1))
}

# likelihood for 10 tosses w/ one success
log_lik <- function(p) {
  ifelse(p <= 0 | p >= 1, -Inf,
         dbinom(1, size = 10, prob = p, log = TRUE))
}

# log-posterior
log_post <- function(p) log_lik(p) + log_prior_saw(p)

# run metropolis algorithm
set.seed(1234)
m <- metrop(log_post, theta_start = 0.5)
```
:::

::: {.column width="50%"}
```{r}
#| label: s2-sawtooth-plot
#| echo: false
#| fig-width: 7
#| fig-asp: 1
par(mfrow = c(2, 1), family = "Source Sans 3", cex = 1.3, las = 1, mar = c(4, 5, 1, 1))
plot(m$samples, type = "l", xlab = "Iteration", ylab = "p")
hist(m$samples, breaks = 100, main = "", xlab = "p")
```
:::
:::

::: {.notes}
Notes: We can use the Metropolis algorithm for our "weird" sawtooth posterior. We'll use the log-posterior in this case.
:::

## Logistic regression: data {.smaller}

```{r}
#| label: s2-logit-data
#| output-location: column
# load only the turnout data frame
turnout <- ZeligData::turnout

# alias for arm::rescale(): mean 0, SD 0.5
rs <- function(x) { arm::rescale(x) }

# create the rescaled variables as columns
turnout <- turnout |>
  mutate(across(age:income, rs, .names = "rs_{.col}"))
f <- vote ~ rs_age + rs_educate + rs_income + race

# fit logit model
fit <- glm(f, family = binomial, data = turnout)

# print estimates
arm::display(fit, digits = 4)
```

::: {.notes}
Notes: Recall the logit model from a previous chapter. This time, let's rescale the numeric variables to have mean of zero and SD of 0.5. This keeps coefficients on a similar scale, which helps us with the computation. For more on this rescaling strategy and its advantages, see Gelman (2008).
Code comment in the notes: rescaling inside the formula (e.g., rs(age)) breaks predictions on new data, because rescale() would then use the new data's mean and SD.
:::

## Log-posterior: [unstable]{.red} {.smaller}

For this example, we imagine an improper, constant prior.

```{r}
#| label: s2-logit-logpost-bad
# make X and y
mf <- model.frame(f, data = turnout)
X <- model.matrix(f, data = mf)
y <- model.response(mf)

# ❌ correct, but unstable, log unnormalized posterior (flat prior, so just the log-likelihood)
log_posterior <- function(beta, y, X) {
  linpred <- X %*% beta
  # for large |linpred|, plogis() rounds to exactly 0 or 1, so dbinom() returns -Inf;
  #   if the current and proposed values are both -Inf, delta = -Inf - (-Inf) = NaN
  sum(dbinom(y, size = 1, prob = plogis(linpred), log = TRUE))
}
```

## Log-posterior: [stable]{.green} {.smaller}

**Why the previous version is unstable.** `plogis(linpred)` rounds to exactly 1 when `linpred` > 37 and to exactly 0 when `linpred` < −745. For an observation on the wrong side, `dbinom(..., log = TRUE)` returns `log(0)` = −Inf, so the whole sum is −Inf. A −Inf *proposal* is harmless (rejected), but if the *current* value is also −Inf (a wild start), `delta` = −Inf − (−Inf) = NaN and `if (delta > 0)` errors.

**The fix: never compute the probability.** Since $\log \pi = \eta - \log(1 + e^\eta)$ and $\log(1 - \pi) = -\log(1 + e^\eta)$, the log-likelihood is $\sum_i [y_i \eta_i - \log(1 + e^{\eta_i})]$. Compute $\log(1 + e^\eta)$ so that `exp()` never overflows and `log1p()` keeps precision when its argument is tiny.

```{r}
#| label: s2-logit-logpost
# ✅ same unnormalized log posterior, but computed on the log scale to avoid rounding
log_posterior <- function(beta, y, X) {
  linpred <- drop(X %*% beta)                        # eta; drop() makes it a vector
  log1pexp <- ifelse(linpred > 0,
                     linpred + log1p(exp(-linpred)),  # eta > 0: exp(-eta) <= 1, cannot overflow
                     log1p(exp(linpred)))             # eta <= 0: exp(eta) <= 1; log1p() exact for tiny x
  sum(y * linpred - log1pexp)                        # = sum(y*log(pi) + (1 - y)*log(1 - pi)), finite for every eta
}
```

At $\beta = (40, 0, 0, 0, 0)$, the unstable version returns −Inf; this one returns −20,320.

::: {.notes}
Checked 2026-10-06: from the deck's four starting values (−2, −1, 1, 2) the unstable version never hits the rounding, so both versions give identical chains here; from a start of 40 the unstable chain errors on its first step ("missing value where TRUE/FALSE needed") and the stable one runs. The two versions agree to 2e-5 at eta = 20. The notes chapter's comment on the unstable version says "0*Inf"; the mechanism is log(0) = -Inf (dbinom with prob exactly 0 or 1), then -Inf - (-Inf) = NaN in the acceptance step.
:::

## Running the algorithm

```{r}
#| label: s2-logit-run
# sample with metropolis
set.seed(1234)
S <- 20000
m1 <- metrop(log_posterior, S = S, tau = 0.1,
             theta_start = rep(-2, ncol(X)), y = y, X = X)
m2 <- metrop(log_posterior, S = S, tau = 0.1,
             theta_start = rep(-1, ncol(X)), y = y, X = X)
m3 <- metrop(log_posterior, S = S, tau = 0.1,
             theta_start = rep(1, ncol(X)), y = y, X = X)
m4 <- metrop(log_posterior, S = S, tau = 0.1,
             theta_start = rep(2, ncol(X)), y = y, X = X)
```

::: {.notes}
Same seed, S, tau, and starting values as `wk07-logit-metropolis.R`, so the numbers on the next slides match what students get (the script adds `progress_bar = TRUE`, which does not change the draws).
:::

## R-hat and ESS: intercept {.smaller}

::: {.columns}
::: {.column width="62%"}
```{r}
#| label: s2-logit-rhat
# first parameter (intercept); discard the first 20% of each chain as burn-in
start <- .2*S + 1
end <-   S
matrix_of_chains <- cbind(
  m1$samples[start:end, 1],
  m2$samples[start:end, 1],
  m3$samples[start:end, 1],
  m4$samples[start:end, 1]
)

# compute r-hat (for intercept)
posterior::rhat(matrix_of_chains)
```
:::

::: {.column width="38%"}
```{r}
#| label: s2-logit-ess
# compute ess (for intercept)
posterior::ess_bulk(matrix_of_chains)
posterior::ess_tail(matrix_of_chains)
```
:::
:::

::: {.notes}
Notes: R-hat less than about 1.01; ESS more than about 2,000.
[CTK: at S = 20,000 the bulk ESS for the intercept is below the notes' 2,000 heuristic (see the output); say whether that is enough here or whether students should run longer, as the notes' 100,000-iteration version does.]
:::

## Intercept: four chains

```{r}
#| label: s2-logit-trace
#| echo: false
#| fig-width: 14
#| fig-asp: 0.45
colnames(matrix_of_chains) <- paste("Chain", 1:ncol(matrix_of_chains))
gg_df <- matrix_of_chains |>
  as_tibble() |>
  mutate(Iteration = start - 1 + 1:n()) |>
  pivot_longer(cols = `Chain 1`:`Chain 4`, names_to = "Chain", values_to = "Sample")

gg1 <- ggplot(gg_df, aes(x = Iteration, y = Sample, color = Chain)) +
  geom_line(linewidth = 0.2) +
  facet_wrap(vars(Chain)) +
  scale_color_lecture() +
  scale_x_continuous(labels = scales::comma, breaks = c(5000, 12500, 20000)) +
  theme(legend.position = "none")

gg2 <- ggplot(gg_df, aes(x = Sample, color = Chain)) +
  geom_density(linewidth = 1) +
  scale_color_lecture() +
  labs(color = NULL)

gg1 + gg2
```

## All parameters

```{r}
#| label: s2-logit-diag-table
#| echo: false
chains_list <- list(
  m1$samples[start:end, ],
  m2$samples[start:end, ],
  m3$samples[start:end, ],
  m4$samples[start:end, ]
)

# helper: build a (iterations x chains) matrix for parameter j
make_mat <- function(j) cbind(
  chains_list[[1]][, j],
  chains_list[[2]][, j],
  chains_list[[3]][, j],
  chains_list[[4]][, j]
)

p <- ncol(X)

# compute diagnostics for all parameters
rhat_all <- vapply(seq_len(p), function(j) posterior::rhat(     make_mat(j)), numeric(1))
ess_bk   <- vapply(seq_len(p), function(j) posterior::ess_bulk( make_mat(j)), numeric(1))
ess_tl   <- vapply(seq_len(p), function(j) posterior::ess_tail( make_mat(j)), numeric(1))

# assemble table (clean names + emoji check for R-hat < 1.01)
tbl <- data.frame(
  Coefficient = colnames(X),
  `R-hat` = sprintf("%.3f", rhat_all),
  `R-hat < 1.01` = ifelse(rhat_all < 1.01, "✅", "❌"),
  `ESS (bulk)` = scales::comma(ess_bk),
  `ESS (tail)` = scales::comma(ess_tl),
  check.names = FALSE
)

tt(tbl)
```

::: {.notes}
Notes: The code above computes the R-hat and ESS for only the first parameter (intercept); we should check the rest as well.
Notes: The table below shows that R-hat is less than 1.01 for every parameter, so we have found no sign that the chains have failed to converge. (True at S = 20,000 too; check the rendered column.)
:::

## `glm()` vs. `metrop()`: point estimates and 95% CIs

```{r}
#| label: s2-logit-compare
#| echo: false
#| fig-width: 12
#| fig-asp: 0.5

# helper to convert a chain matrix slice to a tibble with correct column names
as_draws <- function(mat, start, end, label, coef_names) {
  stopifnot(ncol(mat) == length(coef_names))
  tb <- tibble::as_tibble(mat[start:end, , drop = FALSE], .name_repair = ~coef_names)
  dplyr::mutate(tb, .chain = label)
}

coef_names <- names(coef(fit))

draws <- dplyr::bind_rows(
  as_draws(m1$samples, start, end, "m1", coef_names),
  as_draws(m2$samples, start, end, "m2", coef_names),
  as_draws(m3$samples, start, end, "m3", coef_names),
  as_draws(m4$samples, start, end, "m4", coef_names)
)

# pivot to long for summarizing
draws_long <- draws |>
  tidyr::pivot_longer(cols = dplyr::all_of(coef_names),
                      names_to = "term", values_to = "value")

# posterior summaries
post_summ <- draws_long |>
  dplyr::group_by(term) |>
  dplyr::summarize(
    est   = median(value),
    lo95  = quantile(value, 0.025),
    hi95  = quantile(value, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(source = "metrop()")

# Wald CI summaries from glm()
glm_tidy <- broom::tidy(fit) |>
  dplyr::mutate(
    lo95 = estimate - 1.96*std.error,
    hi95 = estimate + 1.96*std.error,
    est  = estimate,
    source = "glm()"
  ) |>
  dplyr::select(term, est, lo95, hi95, source)

# combine & order
plot_df <- dplyr::bind_rows(post_summ, glm_tidy) |>
  dplyr::mutate(term = factor(term, levels = rev(coef_names)))

cols <- c("glm()" = lecture_colors[["red"]],
          "metrop()" = lecture_colors[["blue"]])

ggplot(plot_df, aes(x = est, y = term, color = source)) +
  geom_errorbar(aes(xmin = lo95, xmax = hi95),
                width = 0.15, position = position_dodge(width = 0.6), alpha = 0.75) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_vline(xintercept = 0, linewidth = 0.5, linetype = 2) +
  labs(
    x = "Coefficient",
    y = NULL,
    color = NULL
  ) +
  scale_color_manual(values = cols) +
  theme(axis.text.y = element_text(hjust = 0))
```

::: {.notes}
Notes: Finally, we can compare the posterior mean and 95% credible intervals to the ML estimate and 95% Wald confidence intervals. They are almost identical.
(The notes' plotting code uses the posterior median for the point; so does this figure.)
:::

## Quantities of interest

To compute a quantity of interest, we need to use our Bayesian invariance property. We need to transform, then summarize, the simulations.

```{r}
#| label: s2-qi-setup
# make X_lo
X_lo <- cbind(
  "constant" = 1, # intercept
  "rs_age" = -0.5, # 1 SD below avg -- see ?arm::rescale
  "rs_educate" = 0,
  "rs_income" = 0,
  "white" = 1 # white indicator = 1
)

# make X_hi by modifying the relevant value of X_lo
X_hi <- X_lo
X_hi[, "rs_age"] <- 0.5  # 1 SD above avg

# function to compute first difference
fd_fn <- function(beta, hi, lo) {
  plogis(hi%*%beta) - plogis(lo%*%beta)
}
```

## Transform, then summarize {.smaller}

```{r}
#| label: s2-qi-loop
# each row is one beta-tilde; discard burn-in; stack the chains
beta_tilde <- rbind(
  m1$samples[start:end, ], # chain 1, minus burn-in
  m2$samples[start:end, ], # chain 2, minus burn-in
  m3$samples[start:end, ], # chain 3, minus burn-in
  m4$samples[start:end, ]  # chain 4, minus burn-in
)

# transform simulations of coefficients into simulations of first-difference
fd_tilde <- numeric(nrow(beta_tilde))  # container
for (i in 1:nrow(beta_tilde)) {
  fd_tilde[i] <- fd_fn(beta_tilde[i, ], hi = X_hi, lo = X_lo)
}

# summarize: posterior mean, posterior SD, and 95% credible interval
mean(fd_tilde)
sd(fd_tilde)
quantile(fd_tilde, probs = c(0.025, 0.975))
```

## With {marginaleffects}

```{r}
#| label: s2-qi-me
# compute qi with the rescaled columns
cmp <- comparisons(fit, variables = list(rs_age = c(-0.5, 0.5)),
                   newdata = datagrid(grid_type = "mean_or_mode"))

# ML estimate, delta-method SE, and 95% confidence interval
cmp |>
  as.data.frame() |>
  select(term, contrast, estimate, std.error, conf.low, conf.high)
```

::: {.notes}
[CTK: the two estimates agree; {marginaleffects} gives the ML point estimate with a delta-method interval, metrop() gives the posterior mean of the first difference.]
:::

## `metrop()` vs. `comparisons()`: the first difference

```{r}
#| label: s2-qi-table
#| echo: false
# one summary row per method; kept for the three-way table in the Stan section
fd_row_glm <- tibble(Method = "glm() + comparisons(): ML + delta method",
                     Estimate = cmp$estimate, SE = cmp$std.error,
                     `2.5%` = cmp$conf.low, `97.5%` = cmp$conf.high)
fd_row_metrop <- tibble(Method = "metrop(): posterior simulations",
                        Estimate = mean(fd_tilde), SE = sd(fd_tilde),
                        `2.5%` = quantile(fd_tilde, 0.025), `97.5%` = quantile(fd_tilde, 0.975))

# the two sets of summaries side by side
bind_rows(fd_row_metrop, fd_row_glm) |>
  tt(notes = "metrop(): posterior mean, posterior SD, and 2.5th and 97.5th percentiles of the simulated first differences. comparisons(): ML estimate, delta-method standard error, and Wald 95% confidence interval.") |>
  format_tt(j = 2:5, digits = 3, num_fmt = "decimal", num_zero = TRUE) |>
  style_tt(j = 2:5, align = "r")
```

::: {.notes}
The first difference in Pr(vote) for a 1 SD shift in age (rs_age from -0.5 to 0.5), with education and income at their means and race = white. Same quantity, two routes: the posterior mean and credible interval from the four Metropolis chains stacked, and the ML point estimate with a delta-method interval from comparisons(). They agree to the second decimal.
:::

## Summary



::: {.columns}
::: {.column width="33%"}
::: {.defbox}
**MCMC**

MCMC offers a powerful and general way to sample from an unnormalized (log-)posterior. The Metropolis algorithm is one such method. Hamiltonian Monte Carlo (HMC) via Stan is even better.
:::
:::

::: {.column width="33%"}
::: {.defbox}
**R-hat and burn-in**

The first MCMC samples are highly dependent on the starting values. Because of this, you need to:

1. Discard the samples from a burn-in period.
2. Run multiple chains and check that R-hat is less than 1.01.
:::
:::

::: {.column width="33%"}
::: {.defbox}
**ESS and large samples**

The MCMC samples are dependent on the previous samples. Because of this, you need to:

1. Generate more samples than you would need if they were independent.
2. Use ESS to understand your effective sample size.
:::
:::
:::
