# Stan

## Stan

- Stan offers an extremely efficient alternative to Metropolis and other MCMC algorithms.
- It uses a hyper-optimized version of Hamiltonian Monte Carlo.
- Stan and its universe of supporting software is extremely well-documented and widely used.

## Logistic regression model

$$
y_i \sim \text{Bernoulli}(\pi_i), \quad \pi_i = \operatorname{logit}^{-1} \left( X_i \beta \right).
$$

We'll assume weakly informative priors (i.e., variance = 25).

$$
\beta \sim \mathcal{N}(0, 5^2 I).
$$

Because the scale of the variables affects the interpretation of the priors, it's common to rescale variables to have a common scale (e.g., SD of 0.5, SD of 1, or range of 1).

## Three blocks

Each Stan model has three main blocks.

1. `data { ... }` Known inputs provided from R.
1. `parameters { ... }` Unknowns to infer (e.g., `vector[K] beta;`, `real<lower=0> sigma;`). Put **constraints** that reflect the parameter's support.
1. `model { ... }` Priors and likelihood live here. Use distribution statements (`theta ~ normal(0,1);`)

There are other blocks you can use as well.

## `logit.stan`

The Stan model mirrors the math closely.

```{.stan code-line-numbers="1-7|9-12|14-18"}
// data block
data {
  int<lower=0> N;  // rows in design matrix
  int<lower=1> K;  // columns in design matrix
  array[N] int<lower=0, upper=1> y;  // binary outcome
  matrix[N, K] X;  // design matrix
}

// parameters block
parameters {
  vector[K] beta;  // logit coefficients
}

// model block
model {
  beta ~ normal(0, 5);               // n(0, 5) prior for each beta
  y ~ bernoulli_logit(X * beta);     // logistic regression likelihood
}
```

## Data for Stan

```{r}
#| label: s3-turnout
# load only the turnout data frame
turnout <- ZeligData::turnout  |>
  # create rescaled columns (rescaling inside the formula breaks predictions on new data)
  mutate(across(age:income, arm::rescale, .names = "rs_{.col}")) |>
  glimpse()
```

## Data for Stan

```{r}
#| label: s3-stan-data
# build model frame
f <- vote ~ rs_age + rs_educate + rs_income + race
mf <- model.frame(f, data = turnout)
X  <- model.matrix(f, data = mf)
y  <- model.response(mf)

# bundle data for Stan
stan_data <- list(
  N = nrow(X),
  K = ncol(X),
  y = as.integer(y),
  X = X
)
```

## Writing `logit.stan` from R

```{r}
#| label: s3-write-stan
#| cache: false
stan_code <- "
data {
  int<lower=0> N;
  int<lower=1> K;
  array[N] int<lower=0, upper=1> y;
  matrix[N, K] X;
}
parameters {
  vector[K] beta;
}
model {
  beta ~ normal(0, 5);               // weakly informative prior
  y ~ bernoulli_logit(X * beta);     // logistic regression likelihood
}
"
# Write the Stan program so the document runs without external files.
writeLines(stan_code, con = 'logit.stan')
```

## Install {cmdstanr} and CmdStan

{cmdstanr} is not on CRAN, so we install it from the Stan r-universe and then use it to install CmdStan.

```{r}
#| label: s3-install
#| eval: false
install.packages("cmdstanr",
                 repos = c("https://stan-dev.r-universe.dev", getOption("repos")))
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

## {cmdstanr}

First, we compile the model with `cmdstan_model()`. Then we draw samples with the `$sample()` method.

```{r}
#| label: s3-sample
#| cache: true
#| results: hide
mod <- cmdstan_model("logit.stan")

fit_cmd <- mod$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,  # run the chains in parallel
  iter_warmup = 1000,
  iter_sampling = 2000,  # excluding warmup
  seed = 123
)
```

```{r}
#| label: s3-sample-load
#| include: false
#| cache: true
#| dependson: s3-sample
# A CmdStanMCMC object reads its draws lazily from CSV files in a temp directory,
# which are gone by the next render, so the cached copy from the chunk above is
# unusable on a warm render. Read the draws and sampler diagnostics into the
# object here and re-assign it, so knitr stores the loaded object in this
# chunk's cache (it is loaded after the chunk above's, and wins).
invisible(fit_cmd$draws())
invisible(fit_cmd$sampler_diagnostics())
fit_cmd <- fit_cmd
```

## `$summary()`

The `$summary()` method reports posterior summaries along with R-hat, bulk ESS, and tail ESS for each parameter.

```{r}
#| label: s3-summary
#| cache: false
fit_cmd$summary(variables = "beta")
```

## Diagnostics

- $\hat{R}$: an R-hat larger than 1.01 shows that the chains have not converged.
- Effective Sample Size (ESS): how much independent information is in the draws.

```{r}
#| label: s3-rhat-ess
#| cache: false
draws_beta <- fit_cmd$draws(variables = "beta")  # draws array
summarise_draws(draws_beta, "rhat", "ess_bulk", "ess_tail")
```

As a rule of thumb, check $\hat{R} < 1.01$ and ESS > 2,000.

## HMC diagnostics {.smaller}

HMC has three diagnostics of its own, reported for each chain. They signal two different kinds of trouble: a [**bad sample**]{.red} (the draws may be biased; do not use them) or an [**inefficient sampler**]{.blue} (the draws are fine; you just got fewer than you paid for).

```{r}
#| label: s3-hmc-diag
#| cache: false
#| message: true
fit_cmd$diagnostic_summary()
```

## HMC diagnostics: rules of thumb {.smaller}

| Diagnostic | Signals | No problem | Problem | What to do |
|:--|:--|:--|:--|:--|
| Divergent transitions | [bad sample]{.red} | 0 after warmup | **any**: the sampler could not follow the posterior's curvature, so it misses part of the posterior | raise `adapt_delta` (default 0.8; try 0.95 or 0.99); if they persist, rescale or reparameterize |
| E-BFMI | [bad sample]{.red} (the tails) | above 0.3 in every chain (about 1 is typical) | **below 0.3**: the sampler does not move well between energy levels, so the tails are explored poorly; longer chains do not fix it | reparameterize; look for heavy tails or funnel-shaped posteriors |
| Maximum treedepth | [inefficient sampler]{.blue} | 0 hits | many hits: trajectories are cut short, so the chain explores slowly; the draws are still valid | raise `max_treedepth` (default 10) or rescale the predictors |

## {bayesplot}: densities by chain

```{r}
#| label: s3-dens-overlay
#| cache: false
#| fig-width: 14
#| fig-asp: 0.42
# densities of parameters by chain
mcmc_dens_overlay(draws_beta)
```

## {bayesplot}: ridges

```{r}
#| label: s3-areas-ridges
#| cache: false
#| fig-width: 14
#| fig-asp: 0.42
# ridges plot of densities of parameters
mcmc_areas_ridges(draws_beta)
```

## {bayesplot}: trace plots

```{r}
#| label: s3-trace
#| cache: false
#| fig-width: 14
#| fig-asp: 0.42
# trace plots of parameters by chain
mcmc_trace(draws_beta)
```

## ShinyStan

The **ShinyStan** app provides an interactive dashboard for convergence diagnostics, traceplots, divergences, and posterior summaries.

```{r}
#| label: s3-shinystan
#| eval: false
library(shinystan)

# launch the app for the {cmdstanr} fit
launch_shinystan(fit_cmd)
```

You can inspect trace plots, divergences, and other visual diagnostics interactively without additional coding.

## Quantities of interest

```{r}
#| label: s3-beta-tilde
#| cache: false
# put the simulations of the coefficients into a matrix
beta_tilde <- fit_cmd$draws(variables = "beta", format = "draws_matrix")
head(beta_tilde)
```

Each row is one draw of $\beta$, and each column is one coefficient.

## First difference

To compute a first difference, transform draws using the invariance principle.

```{r}
#| label: s3-x-lo-hi
#| cache: false
# make X_lo
X_lo <- cbind(
  "constant" = 1, # intercept
  "rs_age" = -0.5, # 1 SD below avg -- see ?arm::rescale
  "rs_educate" = 0,
  "rs_income" = 0,
  "white" = 1 # white indicator = 1
)

# make X_hi by modifying the relevant value of X_lo
X_hi <- X_lo
X_hi[, "rs_age"] <- 0.5  # 1 SD above avg
```

## First difference

```{r}
#| label: s3-fd
#| cache: false
# function to compute first difference
fd_fn <- function(beta, hi, lo) {
  beta <- as.vector(beta)  # to prevent column/row confusion
  plogis(hi%*%beta) - plogis(lo%*%beta)
}

# transform simulations of coefficients into simulations of first-difference
fd_tilde <- numeric(nrow(beta_tilde))  # container
for (i in 1:nrow(beta_tilde)) {
  fd_tilde[i] <- fd_fn(beta_tilde[i, ], hi = X_hi, lo = X_lo)
}

# summarize: posterior mean, posterior SD, and 95% credible interval
mean(fd_tilde)
sd(fd_tilde)
quantile(fd_tilde, probs = c(0.025, 0.975))
```

```{r}
#| label: s3-fd-row
#| include: false
# the Stan row of the first-difference comparison table (shown in the {brms} section)
fd_row_stan <- tibble(Method = "Stan: posterior simulations",
                      Estimate = mean(fd_tilde), SE = sd(fd_tilde),
                      `2.5%` = quantile(fd_tilde, 0.025), `97.5%` = quantile(fd_tilde, 0.975))
```

```{r}
#| label: s3-tidy-compiled-model
#| include: false
#| cache: false
# cmdstan_model() leaves the compiled binary beside logit.stan; both would be
# copied into the published site by `make site`, so remove them after sampling
# (the writeLines chunk above recreates logit.stan on every render)
unlink(c("logit", "logit.stan"))
```

## Summary

**Stan as modern MCMC**\
Stan automates MCMC using a hyper-optimized version of HMC. It efficiently samples from posteriors that are otherwise difficult to explore.

**Interfaces**\
We run Stan from R with {cmdstanr}. The older {rstan} package runs the same Stan programs, but its summaries use older versions of R-hat and ESS.

**Diagnostics and Visualization**\
Check $\hat{R}$, ESS, and the HMC diagnostics (divergent transitions, maximum treedepth, E-BFMI). An R-hat larger than 1.01 or any divergent transitions after warmup mean that we cannot yet trust the draws. The `$summary()` and `$diagnostic_summary()` methods report these, and {bayesplot} and ShinyStan make it easy to explore them visually.

## Stan and Metropolis {.statement}

Stan generalizes what we learned with the Metropolis algorithm: the logic is the same, but the computation is vastly more efficient.
