# \{brms\}

## \{brms\}

**High-level interface to Stan**\
\{brms\} lets you specify Bayesian models using R's familiar formula syntax `y ~ x1 + x2`, then translates them automatically into efficient Stan code for sampling.

**Broad model support**\
It can fit all kinds of models. Much more general than any particular fitting function we've seen so far.

**Seamless post-processing and visualization**\
Built-in tools integrate with \{bayesplot\}, \{posterior\}, and \{tidybayes\} to summarize, diagnose, and visualize posterior draws without writing any Stan code directly.

## The `turnout` example

Here's the syntax for our usual logistic regression model.

```{r}
#| label: s4-fit-brm
#| results: hide
# load only the turnout data frame and hard-code rescaled variables
turnout <- ZeligData::turnout  |>
  mutate(across(age:income, arm::rescale, .names = "rs_{.col}"))

# formula
f  <- vote ~ rs_age + rs_educate + rs_income + race

# fit model with glm()
fit_glm <- glm(f, data = turnout, family = binomial)

# fit model with brm() via cmdstanr
fit_brm <- brm(f, data = turnout, family = bernoulli,
           chains = 4, cores = 4,
           backend = "cmdstanr",
           seed = 1234)
```

## `glm()` and `brm()` {.compare .smaller}

`glm()` and `brm()` obtain the same estimates.

```{r}
#| label: s4-compare
#| code-fold: true
normalize_terms <- function(x) {
  x <- sub("^b_", "", x)  # brms to glm names
  x <- ifelse(x == "(Intercept)", "Intercept", x)  # unify intercept
  x
}

modelsummary(
  list("glm()" = fit_glm, "brm()" = fit_brm),
  coef_rename = normalize_terms,
  statistic = "conf.int",
  notes = "95% confidence intervals for glm(); 95% credible intervals for brm().",
  gof_map = "nobs"
)
```

## Diagnostics

\{brms\} warns only when an R-hat is larger than 1.05, so we check R-hat, ESS, and divergent transitions ourselves.

```{r}
#| label: s4-diagnostics
# R-hat and ESS for each parameter
summarise_draws(fit_brm, "rhat", "ess_bulk", "ess_tail")

# number of divergent transitions after warmup
sum(nuts_params(fit_brm, pars = "divergent__")$Value)
```

## \{marginaleffects\}

And \{brms\} plays very nicely with \{marginaleffects\}.

```{r}
#| label: s4-comparisons
# compute first difference
comparisons(fit_brm, variables = list(rs_age = c(-0.5, 0.5)),
            newdata = datagrid(grid_type = "mean_or_mode"))
```

## Four routes to the same first difference

```{r}
#| label: s4-fd-table
#| echo: false
# brm() row: summaries of the posterior draws of the comparison
cmp_brm <- comparisons(fit_brm, variables = list(rs_age = c(-0.5, 0.5)),
                       newdata = datagrid(grid_type = "mean_or_mode"))
fd_brm <- get_draws(cmp_brm)$draw
fd_row_brm <- tibble(Method = "brm() + comparisons(): posterior simulations",
                     Estimate = mean(fd_brm), SE = sd(fd_brm),
                     `2.5%` = quantile(fd_brm, 0.025), `97.5%` = quantile(fd_brm, 0.975))

# the glm(), Metropolis, and Stan rows were saved in the earlier sections
bind_rows(fd_row_glm, fd_row_metrop, fd_row_stan, fd_row_brm) |>
  tt(notes = "First difference in Pr(vote) for rs_age from -0.5 to 0.5, other predictors at their means and race = white. glm() + comparisons(): ML estimate, delta-method standard error, and Wald 95% confidence interval. The other three rows: posterior mean, posterior SD, and 2.5th and 97.5th percentiles of the simulated first differences (for brm(), computed from get_draws(); comparisons() itself prints the posterior median and the same interval). metrop() uses a flat prior; Stan uses normal(0, 5) priors; brm() uses its defaults (flat on the coefficients).") |>
  format_tt(j = 2:5, digits = 3, num_fmt = "decimal", num_zero = TRUE) |>
  style_tt(j = 2:5, align = "r")
```

## Expected values

```{r}
#| label: s4-predictions
#| fig-width: 12
#| fig-asp: 0.36
# compute and plot expected values (i.e., predicted probabilities)
p <- predictions(fit_brm, variables = list(rs_age = seq(-1, 1, by = 0.1)),
            newdata = datagrid(grid_type = "mean_or_mode"))
ggplot(p, aes(x = rs_age, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(fill = "grey") +
  geom_line(color = lecture_colors[["blue"]])
```

## Modeling "smooth" relationships

To illustrate the power of MCMC and `brm()`, we can imagine fitting a smooth curve for age.

```{r}
#| label: s4-fit-smooth
#| results: hide
# smooth relationship between age and voting--notice the s()
f  <- vote ~ s(rs_age) + rs_educate + rs_income + race
smooth_fit <- brm(f, data = turnout, family = bernoulli,
           chains = 4, cores = 4,
           backend = "cmdstanr",
           control = list(adapt_delta = 0.99),  # avoids divergent transitions
           seed = 1234)
```

## Modeling "smooth" relationships

```{r}
#| label: s4-predictions-smooth
#| fig-width: 12
#| fig-asp: 0.36
# compute and plot expected values (i.e., predicted probabilities)
p <- predictions(smooth_fit, variables = list(rs_age = seq(-1, 1, by = 0.1)),
            newdata = datagrid(grid_type = "mean_or_mode"))
ggplot(p, aes(x = rs_age, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(fill = "grey") +
  geom_line(color = lecture_colors[["blue"]])
```

## Awkwardly scaled variables

Also, `brm()` has no trouble with the variables on their original scales. The original scales pose a serious hurdle for our `metrop()` sampler.

```{r}
#| label: s4-fit-smooth-raw
#| results: hide
# smooth relationship between age and voting--notice the s()
f  <- vote ~ s(age) + educate + income + race
smooth_fit <- brm(f, data = turnout, family = bernoulli,
           chains = 4, cores = 4,
           backend = "cmdstanr",
           control = list(adapt_delta = 0.99),  # avoids divergent transitions
           seed = 1234)
```

## Awkwardly scaled variables

```{r}
#| label: s4-predictions-smooth-raw
#| fig-width: 12
#| fig-asp: 0.36
# compute and plot expected values (i.e., predicted probabilities)
p <- predictions(smooth_fit, variables = list(age = seq(18, 90, by = 1)),
            newdata = datagrid(grid_type = "mean_or_mode"))
ggplot(p, aes(x = age, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_ribbon(fill = "grey") +
  geom_line(color = lecture_colors[["blue"]])
```

## Use `brm()` {.statement}

If `brm()` works for your application, you should much prefer it to writing your own Stan models.
