# LOOIC

```{r}
#| label: s5-helpers
#| echo: false
#| cache: false
# one decimal, as in the notes
f1 <- function(x) sub("^-", "−", sprintf("%.1f", x))
f2 <- function(x) sub("^-", "−", sprintf("%.2f", x))
```

## Why not the AIC? {.smaller}

In *Information Criteria*, we compared Poisson and NB for Holland's Santiago data with AIC and BIC. However, when we have posterior *simulations*, the AIC and BIC work poorly because we do not have a single point estimate. Instead, we have a *distribution*.

The AIC also works poorly with posterior simulations for two other reasons.

1. Plugging in a point estimate (e.g., the posterior mean) ignores the uncertainty in the posterior, while the LOOIC averages over all the draws.
2. The AIC's penalty counts parameters, but a prior that pulls the estimates toward zero makes a model less flexible than its count of parameters suggests. With a strong prior or a hierarchical model, $k$ overstates the model's effective number of parameters.

::: {.key}
When we have a sample from the posterior, we can instead use the LOOIC.
:::

## Example: Linear or cubic age? {.smaller}

First, the data and formulas.

```{r}
#| label: s5-turnout-data
# load only the turnout data frame and create rescaled columns
turnout <- ZeligData::turnout |>
  mutate(across(age:income, arm::rescale, .names = "rs_{.col}"))

# linear in age
f_linear <- vote ~ rs_age + rs_educate + rs_income + race

# cubic in age
f_cubic <- vote ~ rs_age + I(rs_age^2) + I(rs_age^3) +
  rs_educate + rs_income + race
```

Next fit the models.

```{r}
#| label: s5-turnout-fits
#| results: hide
fit_linear_brm <- brm(f_linear, data = turnout, family = bernoulli,
                      backend = "cmdstanr", chains = 4, cores = 4, seed = 1234)

fit_cubic_brm <- brm(f_cubic, data = turnout, family = bernoulli,
                     backend = "cmdstanr", chains = 4, cores = 4, seed = 1234)
```

## Compute the LOOIC {.evensmaller}

:::: {.columns}
::: {.column width="50%"}
```{r}
#| label: s5-loo-linear
# looic for linear model
loo_linear_brm <- loo(fit_linear_brm)
loo_linear_brm
```
:::
::: {.column width="50%"}
```{r}
#| label: s5-loo-cubic
# looic for cubic model
loo_cubic_brm <- loo(fit_cubic_brm)
loo_cubic_brm
```
:::
::::

- `elpd_loo` (higher is better)
- `p_loo` (effective number of parameters)
- `looic` = −2 $\times$ `elpd_loo` (lower is better)

## Compare the two models {.smaller}

```{r}
#| label: s5-loo-compare-turnout
# compare the two models
loo_compare(loo_linear_brm, loo_cubic_brm)
```

```{r}
#| label: s5-turnout-numbers
#| echo: false
#| cache: false
lc_turnout <- loo_compare(loo_linear_brm, loo_cubic_brm)
turnout_elpd_diff <- abs(lc_turnout$elpd_diff[2])
turnout_se_diff <- lc_turnout$se_diff[2]
```

- `loo_compare()` prints the best model in the top row, and `elpd_diff` is relative to that best model.
- `loo_compare()` reports differences on the ELPD scale. ELPD differences are half the size of LOOIC differences and have the opposite sign (higher ELPD is better; lower LOOIC is better). Multiply by (−2) for the LOOIC scale (e.g., `r f1(-turnout_elpd_diff)` elpd becomes `r f1(2 * turnout_elpd_diff)` LOOIC).
- `se_diff` is the SE of the *difference*.

Some rules of thumb:

- For |`elpd_diff`| < 4, the models predict about equally well, regardless of `se_diff`.
- For |`elpd_diff`| > 4 **AND** large compared to `se_diff`: a meaningful difference (see `?loo-glossary`).

::: {.key}
The cubic model is ahead by `r f1(turnout_elpd_diff)` `elpd` with an `se_diff` of `r f1(turnout_se_diff)`. This is modest evidence for the cubic.
:::

## The AIC and BIC, for comparison {.smaller}

```{r}
#| label: s5-aic-bic-turnout
# AIC and BIC from glm() fits, for comparison
fit_linear_glm <- glm(f_linear, family = binomial, data = turnout)
fit_cubic_glm  <- glm(f_cubic, family = binomial, data = turnout)
AIC(fit_linear_glm, fit_cubic_glm)
BIC(fit_linear_glm, fit_cubic_glm)
```

::: {.caption}
(The AIC is `r f1(AIC(fit_cubic_glm))` for the cubic and `r f1(AIC(fit_linear_glm))` for the linear, and the BIC is `r f1(BIC(fit_cubic_glm))` and `r f1(BIC(fit_linear_glm))`.)
:::

## AIC and LOO {.statement}

The AIC assumes the penalty, and LOO measures it.
