# Multinomial logit

## The inverse-logit function

Suppose $y_i \in \{0, 1\}$. Using logistic regression, we model $\Pr(y_i = 1)$ as a function of covariates, so that

$$
\Pr(Y_i = 1) = \frac{\exp(X_i \beta)}{1 + \exp(X_i \beta)}.
$$

The function $f(z) = \frac{e^z}{1 + e^z}$ is called the inverse-logit function and maps any real number $X_i \beta$ into a probability between 0 and 1.

Because the two probabilities must sum to one:

$$
\Pr(Y_i = 0) = 1 - \Pr(Y_i = 1)
             = \frac{1}{1 + \exp(X_i \beta)}.
$$

## The softmax function

The softmax generalizes the inverse-logit---it takes $J$ real inputs (i.e., from $\mathbb{R}^J$) and scales them to sum to one.

$$
\text{softmax}(z_j) = \frac{\exp(z_j)}{\sum_{k=1}^{J} \exp(z_k)}, \qquad j = 1,\ldots,J.
$$

Because the $J$ outputs must sum to one, it is natural to interpret the outputs as *probabilities*.

## The softmax function: $J = 4$

:::: {.columns}
::: {.column width="68%"}

```{ojs}
//| label: s6-softmax-widget
//| echo: false

// Sliders (left column)
viewof sm_z1 = Inputs.range([-5, 5], {step: 0.1, value: -1,  label: "z1"});
viewof sm_z2 = Inputs.range([-5, 5], {step: 0.1, value: 1,   label: "z2"});
viewof sm_z3 = Inputs.range([-5, 5], {step: 0.1, value: 2,   label: "z3"});
viewof sm_z4 = Inputs.range([-5, 5], {step: 0.1, value: 0.5, label: "z4"});

// Softmax pieces
sm_labels = ["Pr(A)","Pr(B)","Pr(C)","Pr(D)"];
sm_zs = [sm_z1, sm_z2, sm_z3, sm_z4];
sm_exps = sm_zs.map(Math.exp);
sm_sumexp = sm_exps.reduce((a, b) => a + b, 0);
sm_ps = sm_exps.map(e => e / sm_sumexp);

// Data for stacked bar (one row, four segments)
sm_data = sm_ps.map((p, i) => ({ label: sm_labels[i], value: p, row: "softmax" }));

// === Two-column layout: sliders (left) + live table (right) ===
{
const container = html`<div style="
  display:flex;
  gap:24px;
  align-items:flex-start;
  padding-top:8px;
"></div>`;

// Left column: sliders
const left = html`<div style="display:grid; gap:10px; min-width:220px;"></div>`;
left.append(viewof sm_z1, viewof sm_z2, viewof sm_z3, viewof sm_z4);

// Right column: table
const rows = sm_labels.map((cls, i) => ({
Class: cls,
z: sm_zs[i],
"exp(z)": sm_exps[i],
Probability: sm_ps[i]
}));

const right = html`<div style="
  min-width:240px;
  max-width:560px;
  padding:4px 0;
  font-size:1.6rem;
  line-height:1.1;
"></div>`;

right.append(
Inputs.table(rows, {
columns: ["Class", "z", "exp(z)", "Probability"],
format: {
z: d => d.toFixed(2),
"exp(z)": d => d.toFixed(3),
Probability: d => (d * 100).toFixed(1) + "%"
}
})
);

container.append(left, right);
return container;
}

// Stacked bar chart
Plot.plot({
width: 940,
height: 200,
marginLeft: 20,
marginRight: 36,
marginBottom: 46,
x: {
domain: [0, 1],
tickFormat: d => (d * 100).toFixed(0) + "%",
label: null,
tickSize: 10,
tickPadding: 8
},
y: { axis: null },
style: {
fontSize: "22px",
fontFamily: "Source Sans 3, sans-serif"
},
color: {
legend: true,
label: "Class",
range: ["#e41a1c", "#377eb8", "#4daf4a", "#984ea3"],
swatchSize: 22,
style: { fontSize: "22px" }
},
marks: [
Plot.barX(
sm_data,
Plot.stackX({ x: "value", y: "row", fill: "label", sort: { fill: null } })
),
Plot.ruleX([0, 1])
]
})
```

:::
::: {.column width="32%"}

**Example inputs**

```{r}
#| label: s6-softmax-fallback
#| echo: false
#| cache: false
# the chapter's four example inputs and their softmax values
softmax_ex <- tibble(z = c(-1, 1, 2, 0.5)) |>
  mutate(p = exp(z) / sum(exp(z)))

softmax_ex |>
  transmute(`Input z` = as.character(z),
            `Softmax p` = sprintf("%.4f", p)) |>
  tt() |>
  style_tt(j = 1:2, align = "r")
```

:::
::::

::: {.notes}
The OJS widget below allows you to experiment with $J = 4$ inputs and see how the outputted probabilities change. (The widget needs the page served over http; the table on the right is the static fallback.)
:::

## Adding covariates

Now suppose the outcome takes $J$ possible values:

$$
y_i \in \{1, 2, \ldots, J\}.
$$

**Example 1**: We might label *vote choice* in the US as (1) *abstain*, (2) *Republican*, (3) *Democrat*, or (4) *other*.

**Example 2**: We might label *coup attempts* in a given country-year as (0) *none attempted*, (1) *failed attempt*, or (2) *successful attempt*.

## Adding covariates

For each category $j$, define a linear predictor $\eta_{ij} = X_i \beta_j$. Here, each $\beta_j$ is a *vector* of coefficients.

We can use the *softmax* function to convert these linear predictors into probabilities that sum to one.

$$
\Pr(Y_i = j) = \frac{\exp(\eta_{ij})}{\sum_{k=1}^{J} \exp(\eta_{ik})},
\qquad j = 1, \ldots, J.
$$

This is a generalization of the inverse-logit to $J$ categories. If $J = 2$, the softmax reduces to the inverse-logit.

## Identification

However, this model is not identified because adding any constant to each of the $\eta_{ij}$ produces the same probabilities.

:::: {.columns}
::: {.column width="50%"}

**Example inputs**

```{r}
#| label: s6-ident-left
#| echo: false
#| cache: false
softmax_ex |>
  transmute(`Input z` = as.character(z),
            `Softmax p` = sprintf("%.4f", p)) |>
  tt() |>
  style_tt(j = 1:2, align = "r")
```

:::
::: {.column width="50%"}

**Adding a constant ($+2$)**

```{r}
#| label: s6-ident-right
#| echo: false
#| cache: false
# add 2 to every input; the softmax values do not change
softmax_ex |>
  transmute(`New input z + 2` = paste0(z, " + 2 = ", z + 2),
            `Softmax p` = sprintf("%.4f", exp(z + 2) / sum(exp(z + 2)))) |>
  tt() |>
  style_tt(j = 1:2, align = "r")
```

:::
::::

Adding a constant to all inputs leaves the softmax unchanged. This is why the multinomial logit model requires an identification constraint.

## Identification

To identify the model, we set $\beta_J = 0$, so that $\eta_{iJ} = 0$. Then for $j = 1, \ldots, J-1$, we have

$$
\Pr(Y_i = j) =
\frac{\exp(X_i \beta_j)}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$

And for the "baseline" category $J$, we have

$$
\Pr(Y_i = J) =
\frac{1}
     {1 + \sum_{k=1}^{J-1} \exp(X_i \beta_k)}.
$$

## Example: the 1992 presidential vote

- 1992 ANES; same respondents as the wk06 Bush approval example
- Outcome: `vote_1992`: Bush, Clinton, or Perot
- Predictors: `party_id` (−3 to 3), `economy` (1 = much better to 5 = much worse)

```{r}
#| label: s6-vote-data
# load data; keep complete cases on the model's variables
vote <- read_csv("https://pos5747.github.io/data/bush-approval-1992.csv") |>
  select(vote_1992, party_id, economy) |>
  drop_na() |>
  mutate(vote_1992 = factor(vote_1992, levels = c("Bush", "Clinton", "Perot"))) |>
  glimpse()
```

::: {.notes}
[CTK: why Perot makes this nominal, not ordered]

[CTK: 592 voters complete on these variables]

[CTK: provenance: Adolph's Problem Set 4; same file as the ordered example; Perot as the reason a nominal model is needed]
:::

## `multinom()` {.smaller}

```{r}
#| label: s6-vote-multinom
# fit multinomial logit; first level (Bush) is the baseline
fit_vote <- multinom(vote_1992 ~ party_id + economy, data = vote, trace = FALSE)
summary(fit_vote)
```

::: {.notes}
[CTK: two rows of coefficients, one linear predictor per non-baseline candidate; each relative to Bush]

[CTK: baseline first here (`multinom()`), last in the math above; the probabilities don't change]

[CTK: signs: party_id negative for both rows (Republicans → Bush); economy positive for both (worse economy → away from Bush)]
:::

## Probabilities with {marginaleffects}

```{r}
#| label: s6-vote-probs
#| cache: false
#| output-location: column
#| fig-width: 7
#| fig-asp: 0.85
# probability of each candidate as party ID varies
p_vote <- predictions(
  fit_vote,
  newdata = datagrid(party_id = -3:3),
  type = "probs")

# colors follow the factor levels:
# Bush, Clinton, Perot
ggplot(p_vote,
       aes(x = party_id, y = estimate,
           ymin = conf.low, ymax = conf.high,
           color = group, fill = group)) +
  geom_ribbon(alpha = 0.2, color = NA) +
  geom_line() +
  scale_color_lecture() +
  scale_fill_lecture() +
  labs(x = "Party ID", y = "Probability",
       color = "Candidate", fill = "Candidate")
```

::: {.notes}
[CTK: Pr(Perot) rises then falls (about 0.08 at −3, peak about 0.28 at 0 to 1, 0.18 at 3), though both linear predictors are linear in party ID]

[CTK: why: Perot's party_id coefficient (−0.44) sits between Bush's (0) and Clinton's (−1.10); the softmax shares probability]

[CTK: so the sign of a coefficient doesn't give the direction of that category's probability; compute probabilities]
:::

## First differences

```{r}
#| label: s6-vote-fd
#| cache: false
# change in each candidate's probability as the economy goes from
# "better" (2) to "worse" (4), for a typical respondent
comparisons(fit_vote,
            variables = list(economy = c(2, 4)),
            newdata = datagrid(grid_type = "mean_or_mode"),
            type = "probs") |>
  as_tibble() |>
  select(term, contrast, group, estimate, conf.low, conf.high)
```

::: {.notes}
[CTK: worse economy: Bush −0.22, Clinton +0.18, Perot +0.04 (CI includes 0)]

[CTK: Perot's economy coefficient is positive (0.34), but Clinton gains most of what Bush loses]
:::

## `brm()`

```{r}
#| label: s6-vote-brm
#| results: hide
# fit the same model with brm(); categorical() is the multinomial logit,
# with the first level (Bush) as the baseline
fit_vote_brm <- brm(vote_1992 ~ party_id + economy,
                    data = vote,
                    family = categorical(link = "logit"),
                    backend = "cmdstanr",
                    chains = 4,
                    cores = 4,
                    seed = 1234)
```

## `brm()` vs. `multinom()` {.compare}

```{r}
#| label: s6-vote-compare
#| cache: false
#| echo: false
# put the multinom() and brm() estimates side by side
compare_mnl <- function(fit_ml, fit_brm) {
  # multinom(): ML estimates and standard errors
  ml <- broom::tidy(fit_ml) |>
    transmute(outcome = y.level,
              term = if_else(term == "(Intercept)", "Intercept", term),
              est_ml = estimate,
              se_ml = std.error)

  # brm(): posterior means and SDs; mu is brms's prefix for each outcome's
  # linear predictor, so the names look like mu<outcome, no spaces>_<term>
  bayes <- fixef(fit_brm) |>
    as_tibble(rownames = "name") |>
    separate_wider_delim(name, delim = "_", names = c("mu", "term"),
                         too_many = "merge") |>
    transmute(mu, term, est_brm = Estimate, se_brm = Est.Error)

  # match brm()'s names to multinom()'s outcome labels
  ml |>
    mutate(mu = paste0("mu", str_remove_all(outcome, "[^[:alnum:]]"))) |>
    left_join(bayes, by = c("mu", "term")) |>
    select(-mu)
}

# make a table with one block of rows per outcome
tt_mnl <- function(comparison) {
  comparison |>
    select(-outcome) |>
    as.data.frame() |>
    setNames(c("Term", "Estimate", "SE", "Estimate", "SE")) |>
    tt(notes = "multinom(): ML estimates and standard errors. brm(): posterior means and standard deviations.") |>
    format_tt(j = 2:5, digits = 3, num_fmt = "decimal") |>
    style_tt(j = 2:5, align = "r") |>
    group_tt(i = comparison$outcome,
             j = list("multinom()" = 2:3, "brm()" = 4:5))
}

compare_mnl(fit_vote, fit_vote_brm) |>
  tt_mnl() |>
  style_tt(i = "groupi", align = "l")  # outcome labels on the left
```

::: {.caption}
Table code: `compare_mnl()` and `tt_mnl()` in the notes (Multinomial Logit, With `brm()`)
:::

::: {.notes}
[CTK: posterior means within about 0.02 of the ML estimates (e.g., Clinton party_id −1.11 vs −1.10; Perot economy 0.34 vs 0.34); default priors barely matter at N = 592]
:::

## `brm()` + {marginaleffects}

```{r}
#| label: s6-vote-brm-probs
#| cache: false
#| output-location: column
#| fig-width: 7
#| fig-asp: 0.85
# for brms categorical fits (and only those), predictions()
# drops the newdata columns and returns rowid + group, so
# build the grid first and join party_id back by rowid
grid_vote <- datagrid(model = fit_vote_brm,
                      party_id = -3:3)
p_vote_brm <- predictions(fit_vote_brm,
                          newdata = grid_vote) |>
  left_join(select(grid_vote, rowid, party_id),
            by = "rowid")
ggplot(p_vote_brm,
       aes(x = party_id, y = estimate,
           ymin = conf.low, ymax = conf.high,
           color = group, fill = group)) +
  geom_ribbon(alpha = 0.2, color = NA) +
  geom_line() +
  scale_color_lecture() +
  scale_fill_lecture() +
  labs(x = "Party ID", y = "Probability",
       color = "Candidate", fill = "Candidate")
```

::: {.notes}
[CTK: same {marginaleffects} code, no `type = "probs"`; `group` column holds the candidate; the grid columns need joining back by `rowid`]

Checked 2026-10-06 (marginaleffects 1.0.0): the same call on the multinom() fit and on a brms bernoulli fit returns the grid columns; only the brms categorical fit drops them. A marginaleffects quirk, not a brms one; an investigation prompt was written for Carlisle.

[CTK: why bother with brm(): priors, hierarchical extensions (weeks 10–11)]
:::

## Another example: Balancing the federal budget

```{r}
#| label: s6-cces-fit
#| echo: false
# 2012 CCES, from the notes' data folder; "Raise taxes" first, so it is the baseline
cces <- read_csv("wk07-lecture-material/data/cces2012-budget.csv") |>
  mutate(budget = factor(budget, levels = c("Raise taxes",
                                            "Cut defense spending",
                                            "Cut domestic spending")))

fit_budget <- multinom(budget ~ party_id + female + income, data = cces, trace = FALSE)
```

```{r}
#| label: s6-cces-numbers
#| echo: false
#| cache: false
# probability of each choice as party ID varies
p_budget <- predictions(fit_budget, newdata = datagrid(party_id = 1:7), type = "probs")

# the numbers quoted in the bullets
pb <- function(choice, pid) {
  sprintf("%.2f", p_budget$estimate[p_budget$group == choice & p_budget$party_id == pid])
}
```

:::: {.columns}
::: {.column width="58%"}

```{r}
#| label: s6-cces-probs
#| echo: false
#| cache: false
#| fig-width: 7.5
#| fig-asp: 0.75
# probability of each choice as party ID varies, with 95% CIs
ggplot(p_budget, aes(x = party_id, y = estimate,
                     ymin = conf.low, ymax = conf.high,
                     color = group, fill = group)) +
  geom_ribbon(alpha = 0.2, color = NA) +
  geom_line() +
  scale_color_lecture() +
  scale_fill_lecture() +
  guides(color = guide_legend(nrow = 2), fill = guide_legend(nrow = 2)) +
  labs(x = "Party ID", y = "Probability",
       color = "Most preferred", fill = "Most preferred")
```

:::
::: {.column width="42%"}

- 2012 CCES; `r format(nrow(cces), big.mark = ",")` respondents
- Three categories: cut defense spending, cut domestic spending, or raise taxes.
- `female` at its mode (`r unique(p_budget$female)`) and `income` at its rounded mean (`r unique(p_budget$income)`, \$50,000--59,999).
- Strong Democrats to strong Republicans: Pr(cut domestic spending) `r pb("Cut domestic spending", 1)` to `r pb("Cut domestic spending", 7)`; Pr(cut defense spending) `r pb("Cut defense spending", 1)` to `r pb("Cut defense spending", 7)`; Pr(raise taxes) `r pb("Raise taxes", 1)` to `r pb("Raise taxes", 7)`.

:::
::::

::: {.notes}
Model: `multinom(budget ~ party_id + female + income, data = cces)`; party ID from 1 (strong Democrat) to 7 (strong Republican). This example is inspired by a SAGE Research Methods Datasets guide (Reid 2020).
:::

## Callback: ordered logit with `brm()` {.smaller}

```{r}
#| label: s6-approval-fits
#| results: hide
# the wk06 Bush approval data and model
approval <- read_csv("https://pos5747.github.io/data/bush-approval-1992.csv") |>
  mutate(bush_approval = factor(bush_approval, ordered = TRUE))

f_approval <- bush_approval ~ military_force + ideology_distance +
  economy + party_id + education

fit_approval <- MASS::polr(f_approval, data = approval, Hess = TRUE)
fit_approval_brm <- brm(f_approval,
                        data = approval,
                        family = cumulative(link = "logit"),
                        backend = "cmdstanr",
                        chains = 4,
                        cores = 4,
                        seed = 1234)
```

::: {.notes}
[CTK: wk06's `polr()` fit, now with `brm(family = cumulative("logit"))`; same parameterization (cutpoints = `Intercept[j]`)]
:::

## Callback: ordered logit with `brm()`

```{r}
#| label: s6-approval-compare
#| cache: false
# posterior means next to the ML estimates (polr() calls the cutpoints zeta)
ml_approval <- c(fit_approval$zeta, coef(fit_approval))
bayes_approval <- fixef(fit_approval_brm)[, "Estimate"]

tibble(polr_term = names(ml_approval),
       polr = ml_approval,
       brm_term = names(bayes_approval),
       brm = bayes_approval)
```

## Ordered vs. nominal

:::: {.columns}
::: {.column width="50%"}

::: {.defbox}
**Ordered**

`cumulative()`

one set of coefficients
:::

:::
::: {.column width="50%"}

::: {.defbox}
**Nominal**

`categorical()`

$J - 1$ sets
:::

:::
::::

::: {.notes}
[CTK: ordered (`cumulative()`) vs nominal (`categorical()`): one set of coefficients vs J − 1 sets; Red State Part 3 in the exercises (did not vote / Harris / Trump, an unordered outcome)]
:::
